Nanophysiology Approach Reveals Diversity in Calcium Microdomains across Zebrafish Retinal Bipolar Ribbon Synapses

Version 1

Abstract

Abstract

Rapid and high local calcium (Ca2+) signals are essential for triggering neurotransmitter release from presynaptic terminals. In specialized bipolar ribbon synapses of the retina, these local Ca2+ signals control multiple processes, including the priming, docking, and translocation of vesicles on the ribbon before exocytosis, endocytosis, and the replenishment of release-ready vesicles to the fusion sites for sustained neurotransmission. However, our knowledge about Ca2+ signals along the axis of the ribbon active zone is limited. Here, we used fast confocal quantitative dual-color ratiometric line-scan imaging of a fluorescently labeled ribbon binding peptide and Ca2+ indicators to monitor the spatial and temporal aspects of Ca2+ transients of individual ribbon active zones in zebrafish retinal rod bipolar cells (RBCs). We observed that a Ca2+ transient elicited a much greater fluorescence amplitude when the Ca2+ indicator was conjugated to a ribeye-binding peptide than when using a soluble Ca2+ indicator, and the estimated Ca2+ levels at the ribbon active zone exceeded 26 μM in response to a 10-millisecond stimulus, as measured by a ribbon-bound low-affinity Ca2+ indicator. Our quantitative modeling of Ca2+ diffusion and buffering is consistent with this estimate and provides a detailed view of the spatiotemporal [Ca2+] dynamics near the ribbon. Importantly, our data demonstrates that the local Ca2+ levels may vary between ribbons of different RBCs and within the same cells. The variation in local Ca2+ signals is correlated to ribbon size, which in turn correlates with active zone extent, as serial electron microscopy provides new information about the heterogeneity in ribbon size, shape, and area of the ribbon in contact with the plasma membrane.

Introduction

Sensory synapses in the retina rely on the proper function of a specialized organelle, the synaptic ribbon 1-5. Retinal bipolar cells serve as the major conduit for transmitting visual information across the vertebrate retina. Rod bipolar cells (RBCs) can release brief bursts of neurotransmitters to signal a change in contrast or sustain the continuous release of neurotransmitters in a graded manner to provide an analog read-out of luminance 6. To maintain this ability, the RBCs must exert dynamic control over neurotransmitter release rate and facilitate efficient recruitment of release-ready vesicles to fusion sites near the synaptic ribbon. It is established that the elevation of presynaptic Ca2+ in RBCs regulates both dynamic changes in the release rate and accelerates the rate of vesicle replacement 6-14. However, the spatiotemporal properties of the Ca2+ signals that control neurotransmitter release and the molecular entities that regulate the interplay between Ca2+ signals and synaptic vesicle dynamics to sustain kinetically distinct neurotransmitter release components remain poorly understood. Here, we begin to address this lack of knowledge by measuring local Ca2+ signals at positions along the synaptic ribbon at different distances from the active zone in retinal RBCs.

Resolving local Ca2+ signals is technically challenging as it requires information about the spatiotemporal properties of Ca2+ signals specific to ribbon sites. Previous studies of Ca2+ dynamics in goldfish and mammalian RBCs focused primarily on the terminal as a whole and used quantitative methods to examine bulk Ca2+ levels, which are significantly lower and slower than those occurring at the active zone 15-18. Our previous studies estimated that the Ca2+ signals at a single ribbon active zone in zebrafish RBCs likely exceed sub-micromolar concentration level within 3 milliseconds after the opening of voltage-gated Ca2+ (Cav) channels 19. However, these studies were based on estimates using soluble Ca2+ indicators to estimate the Ca2+ signals at the plasma membrane, which are free to diffuse and thus cause the spread of the signal. Measurements of Ca2+ signals along the axis of the ribbon have not been attempted, even though these signals likely control replenishment and priming of vesicles for slower phases of neurotransmitter release. To determine Ca2+ signals along the axis of zebrafish RBC ribbons, we establish a nanophysiology ratiometric approach that measures the spatial and temporal properties of Ca2+ transients by targeting the Ca2+ indicator to the ribbon via conjugation to a ribbon binding peptide (RBP) 20. Resolving Ca2+ signals in the immediate vicinity of the active zone is not currently possible due to the finite spatiotemporal resolution of optical imaging. Thus, we employ a quantitative model using Ca2+ diffusion and buffering to resolve the RBC Ca2+ signals associated with the active zone and along the ribbon and provide a detailed description of the spatiotemporal Ca2+ dynamics across zebrafish RBC ribbons. Our nanophysiology approach for measuring local Ca2+ transients at a single ribbon with high spatiotemporal resolution provides the first evidence of heterogeneity of Ca2+ signals at zebrafish RBC ribbon synapses.

Results

Ca2+ concentrations at single ribbon locations measured using high- and low-affinity diffusible indicators

Synaptic vesicles in RBCs are distributed among at least four distinct pools based on their fusion kinetics, which are assumed to reflect the average proximity of vesicles to Cav channels and the state of vesicle preparedness for Ca2+-triggered fusion 1,6,8,11-13,21-25 (see Supplementary Fig. 1). To visualize and measure Ca2+ signals near the RBC synaptic ribbon controlling such kinetically distinct components of neurotransmitter release, we established a quantitative nanophysiology approach shown in Fig.1 (see also Materials and Methods). In this approach, both the ribbon location and the spatiotemporal Ca2+ signal profile were simultaneously measured by dialyzing the zebrafish RBC terminals with both TAMRA (tetramethyl rhodamine)-labeled RIBEYE-binding peptide (RBP) 26 (to label synaptic ribbons; Fig. 1Ai) and a high-affinity Ca2+ indicator Cal-520® (Cal520HA, effective KD 795 nM; see Materials and Methods) using a whole-cell patch pipette placed directly at the cell terminal. A rapid x-t line scan was taken perpendicular to the plasma membrane across a ribbon, extending from the extracellular space to the cytoplasmic region beyond the ribbon (Fig. 1Aii). The spatial resolution was limited by the point spread function (PSF) of the microscope to approximately 270 nm (Supplementary Fig.2A). In our previous studies, line scans have been applied primarily to measure the temporal properties of Ca2+ signals 19. However, x-t raster plots obtained at a ribbon active zone labeled with RBP allow us to characterize the spatial localization of Ca2+ transients relative to the synaptic ribbon and plasma membrane (Fig.1C-D) in addition to characterizing its temporal aspects (Fig.1E-F). We previously used a similar approach for localizing and tracking single synaptic vesicles before and during fusion at a single ribbon 27, and to measure the kinetics of clearance of fused synaptic vesicle membrane in zebrafish RBC 28. We found that depolarization-evoked Ca2+ influx caused a rapid increase in Ca2+ signals at ribbon locations (Fig.1C, cyan, white horizontal arrowhead) and increased more slowly and less dramatically in the cytoplasm (Fig.1C, cyan, white vertical arrow) in zebrafish RBC. The Sigmoid-Gaussian function fitting of the x-t scans horizontal profile scans (see Materials and Methods) show that the local Ca2+ signals increased rapidly during stimuli (Fig.1D, cyan line) and then decreased immediately after the end of depolarization (Fig.1D, black lines) approaching the spatial profile corresponding to the resting Ca2+ levels (Fig.1D, gray lines). As expected, the centroid position of Ca2+ signals during depolarization (Fig.1D, cyan x0) is closer to the plasma membrane (Fig.1D, magenta x1/2) than the centroid position of RBP (Fig.1D magenta, x0), since Cav channels are located at the membrane 19,26,29-42.

Nanophysiology approach unveils spatiotemporal properties of local Ca2+ signaling in retinal RBC terminals.

A. Left panel. Single projection from a series of confocal optical sections through a zebrafish RBC synaptic terminal. A synaptic terminal was voltage-clamped using a whole-cell pipette with an internal solution containing TAMRA-RBP (magenta) to label synaptic ribbons (Ai, magenta). RBP fluorescence was concentrated at ribbons and also filled the entire terminal, allowing visualization of the terminal border. Experiments were carried out on ribbons that could be distinguished from adjacent ribbons (white asterisks). Scale, 5 μm. Right panel, Close-up view of a single synaptic ribbon. The outside of the cell is to the right, and x-t scan lines (dotted lines) were positioned perpendicular to the plasma membrane, extending from the intracellular side of the ribbon to the extracellular space. A rapid x-t line scan was taken at a ribbon location perpendicular to the plasma membrane across a ribbon (Aii) with sequential dual laser scanning performed at rates of 1.51 milliseconds per line per channel (3.02 milliseconds per line for both channels). The resulting x-t raster plots were used to measure the fluorescence intensity profiles of RBP (magenta) and the Ca2+ transient (cyan; Cal520HA) shown in panels C and E. Scale, 1.6 μm.

B. Voltage-clamp recording of a RBC terminal. Terminals were held at -65 mV and stepped to 0 mV (t0) for 10 ms (red) to evoke a brief Ca2+ current (black). A sinusoidal voltage stimulus was used to monitor membrane capacitance (blue). A typical experiment began with a voltage command (VH= -65 mV), and a TTL pulse generated by the Patch Master software (t1) triggers image acquisition (t2). t0 is the time of depolarization.

C. Illustration of the approach used to obtain the spatial location of Ca2+ signals with respect to the ribbon. Example of an x-t raster plot that is oriented to illustrate the x-axis intensity profiles of RBP (magenta) and Ca2+ signal (cyan) fluorescence during a brief depolarization. Sequential dual laser scanning was performed at 1.51 milliseconds per line for one channel (3.02 milliseconds per line for both channels).

D. Fluorescence intensity profiles along the x-axis for RBP (magenta) and Cal520HA before stimuli (gray line), during 10 ms depolarization (cyan line), and after stimuli (black line) depolarization were obtained by averaging three pixels along the time-axis. RBP (magenta) and Ca2+ signals during (light cyan) were fit with a Sigmoid-Gaussian function 1. The centroid (x-axis position) of the RBP (magenta) and Ca2+ signals during (cyan) were taken as the peak of the Gaussian fit (x0). The parameter x1/2 (dotted magenta line) from the Sigmoid fit to the RBP fluorescence (magenta trace) was used to estimate the location of the plasma membrane.

E. The x-t raster plot shown was from the same recording as in panel C but re-oriented to demonstrate the t-axis intensity profiles of RBP (magenta) and Ca2+ transient (cyan).

F. Spatially averaged Cal520HA fluorescence as a function of time at ribbon proximal (light cyan), distal (dark cyan), and cytoplasmic (gray) locations from the single ribbon shown in D, upper panel.

To characterize the temporal Ca2+ profile at different distances from the plasma membrane, we analyzed the x-t line scans along the time-axis at three distinct distances (Fig. 1E-F) based on the spatial profile of the ribbon described in Fig.1C-D (see also Materials and Methods). We will refer to the corresponding three measurements as ribbon-proximal, ribbon-distal, and cytoplasmic (see Materials and Methods). We found that at the onset of the stimulus (Fig. 1F, black arrow), the fluorescence at the location near the ribbon proximal to the membrane (ribbon-proximal) rose more rapidly and to a higher level (Fig. 1F, light cyan trace) than that at the ribbon-distal and cytoplasmic locations (Fig. 1F, dark cyan, and gray trace). To quantify the kinetics of ribbon-proximal and distal Ca²⁺ signals, we averaged multiple x-t scans acquired with Cal520HA under the same imaging conditions (Fig. 2A). The rise time of the Ca2+ signals produced in response to brief 10 milliseconds stimuli, defined here as the time for the averaged Ca2+ transient measured with Cal520HA to reach half-maximal peak fluorescence amplitude from the resting level (Fig. 2A), is faster for the proximal (∼2.81 milliseconds) than distal (∼3.44 millisecond) signals (Fig. 2A). The corresponding maximum values of trial-averaged ΔF/Frest (changes in the Cal520HA fluorescence during depolarization normalized to background level before depolarization) was significantly higher at ribbon-proximal locations (nearest the Cav channels) than those at ribbon-distal locations, with mean ± SEM of 1.9 ± 0.3 and 1.5 ± 0.25, respectively (Fig. 2A; p< 0.001, N=24; 95% confidence limit). These findings suggest that the spatial resolution of our Ca2+ imaging using Cal520HA is sufficient to resolve differences in the smaller Ca2+ signals at ribbon proximal and distal locations.

Kinetics of Ca2+ transients in response to brief stimuli recorded with freely diffusible indicators.

A. Spatially averaged Cal520HA fluorescence as a function of time at ribbon proximal (light cyan) and distal (dark cyan) locations from a single ribbon, as shown in Fig.1 (n=24 ribbons from 7 different RBCs).

A inset. The temporal profile of events between 20-60 ms is shown in an expanded view for better visualization. Scale bars: vertical, 0.5 (ΔF/Frest); horizontal,10 ms.

B. Spatially averaged Cal520LA fluorescence as a function of time at ribbon proximal (light cyan) and distal (dark cyan) locations. Data points show the average intensity (± SEM) in each horizontal row of 5 pixels for three 10 ms depolarizations at distinct ribbon locations (see Materials and Methods and Fig.1). Fluorescence intensity is normalized with respect to the baseline fluorescence before stimulation, and averaged over all pixels (i.e., over space and time). The arrow indicates the onset of the 10-ms depolarizing stimulus. (n=21 ribbons from 4 different RBCs)

B inset. Temporal profile of events between 20-60 ms was expanded for better visualization. Scale bars: vertical, 0.5 (ΔF/Frest); horizontal, 10 ms.

Since Cal520HA will be saturated at the Ca2+ levels expected in Ca2+ microdomains relevant for vesicle exocytosis 43, we repeated the x-t line scan analysis with a lower-affinity soluble Ca2+ indicator Cal520LA (KD 90 μM; Fig. 2B) allowing us to better define the typical Ca2+ signals controlling distinct neurotransmitter release components corresponding to locations proximal and distal relative to the synaptic ribbon (Fig. 2B). We found that the ribbon-proximal signals detected with Cal520LA (Fig. 2B, light cyan) showed a sharper rise and decay at the onset and termination of the stimulus (Fig. 2B, black arrow), when compared to ribbon-distal signals (Fig. 2B, dark cyan), as one would expect for nanodomain Ca2+ elevations 44-46. The relationship between rise times for the proximal (4.58 millisecond) and distal (7.10 millisecond) signals were similar to those we found with Cal520HA, with both proximal Ca2+ signals being faster than the distal signals. In twenty-one similar experiments, the peak ΔF/Frest at the membrane after 10-ms depolarization was significantly larger for proximal than distal signals (ΔF/Frest: 3.1 ± 0.4 and 1.9 ± 0.2 respectively P=0.001, N= 21, 95% confidence limit). As expected, the ratio of proximal to distal signals measured with Cal520LA (1.6) was significantly higher than that measured with Cal520HA (1.3).

Ca2+ concentrations at single ribbon locations measured using ribbon-bound indicators

Although Ca²⁺-sensitive fluorescent chemical dyes have been used previously 7,20,29,36,42,43,45,46,49-59, the visualization of signals within smaller domains using freely diffusible Ca²⁺ reporters is limited by the resolution of light microscopy and the spread of the indicator by diffusion 60. Diffusible Ca²⁺ indicators report space-averaged Ca²⁺ concentrations, and their intracellular diffusion inherently broadens the spatial resolution of Ca²⁺ nanodomains. To partially overcome this problem, we targeted the Ca2+ indicators to the ribbon by fusing them to the RBP, as described previously 20. RBP-conjugated Ca2+ indicators Cal520HA-RBP or Cal520LA-RBP were introduced to the RBC terminal together with fluorescently labeled RBP via whole-cell voltage clamp by placing the patch pipette directly at the terminal while imaging the terminal using laser scanning confocal microscopy. For two-color imaging, ribbons were labeled with TAMRA-RBP that did not interfere with Cal520-RBP fluorescence, and both channels were scanned sequentially to prevent possible bleed-through. Under these conditions, we used the spots detected by TAMRA-RBP to define the locus of the synaptic ribbon (Fig 3A&B, magenta). We also found punctate regions with both Cal520HA-RBP (Fig 3A, cyan) and Cal520LA-RBP (Fig 3B, cyan) at the same location, on a dimmer fluorescent background of the synaptic terminal, which correspond to the Ca2+ indicator-peptide complexes that are bound and not bound to the ribbon, respectively 20. The overall changes in ΔF/Frest were averaged over several trials for proximal and distal Ca2+ signals measured with Cal520HA-RBP (Fig. 3C, light cyan vs. dark cyan) and Cal520LA-RBP (Fig. 3D, light cyan vs. dark cyan). When compared to their distal location counterparts, the Ca2+ signals proximal to the membrane showed a sharper rise and decay at both the onset and termination of stimuli (Figs. 3C and 3D, light vs. dark cyan trace), as expected when comparing nano- and microdomain Ca2+ profiles. Our data shows that the amplitude differences between ribbon-proximal and ribbon-distal Ca2+ signals were well resolvable using the ribbon-bound Cal520HA-RBP indicator (Fig. 3C, light vs. dark cyan traces: ΔF/Frest = 3 ± 0.4 vs. 1.9 ± 0.3, respectively, p=0.001) and Cal520LA-RBP indicator (Fig. 3D, light vs. dark cyan traces: ΔF/Frest = 5.5 ± 0.9 vs. 3.3 ± 0.8, respectively, p=0.003). The amplitudes of ribbon-proximal Ca2+ signals were higher when measured with Cal520LA-RBP than with Cal520LA-free (Fig. 3E, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray): ΔF/Frest = 5.5 ± 0.9, n=30 vs. 3 ± 0.4, n=19, p=0.04) but this was not the case for distal Ca2+ signals (Fig. 3F, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray): ΔF/Frest = 3.3 ± 0.8, n=30 vs. 1.9 ± 0.2, n=21, p=0.15). Notably, the Ca2+ signals at distal sites measured with Cal520LA-RBP reached their peak amplitude earlier (Fig. 3F inset, dark cyan) than those measured with Cal520LA-free (Fig. 3F inset, light gray). Together, these findings strongly suggest that conjugating the Cal520LA indicator to RBP provides a more accurate, promising approach for measuring the distinct local Ca2+ signals at ribbon-proximal vs. ribbon-distal locations.

Temporal properties of Ca2+ transients recorded with free and ribeye bound Ca2+ indicators.

A-B. Confocal images of the isolated RBCs that were whole-cell voltage-clamped using an internal solution containing the TAMRA-RBP (magenta) and either (A) Cal520HA-RBP (cyan) or (B) Cal520LA-RBP (cyan). Note prominent spots in both TAMRA-RBP and ribeye-bound Ca2+ indicators (A) Cal520HA-RBP (cyan) or (B) Cal520LA-RBP, showing the location of the ribbon. Scale bars, 2 μm.

C-D. Spatially averaged fluorescence of (C) Cal520HA-RBP (n=19) or (D) Cal520LA-RBP (n=30) as a function of time at ribbon proximal (light cyan), and distal (dark cyan) locations. Data points show the average intensity (± SEM) in each horizontal row of five pixels for 10 ms depolarizations at distinct ribbon locations. Fluorescence intensity at the onset of the 10 ms depolarizing stimulus (arrow) was normalized to the baseline fluorescence before stimulation and averaged over all pixels (i.e., over space and time).

C-D inset. Temporal profile of events between 10-50 ms were expanded for better visualization. Scale bars: vertical, 1 (ΔF/Frest, C inset) or 2 (ΔF/Frest, D inset); horizontal, 10 ms.

E-F. Average fluorescence intensity of (E) proximal and (F) distal Ca2+ signals obtained with Cal520LA-RBP (light cyan and dark cyan, respectively) and Cal520L-free (grey).

E-F inset. Events between 10-50 ms was expanded for better visualization. Scale bars: vertical, 2 (ΔF/Frest, E inset) and 1 (ΔF/Frest, F inset); horizontal, 10 ms.

This method also enables ratiometric measurements with RBP-conjugated Ca2+ indicator by normalizing its fluorescence to that of RBP-peptides conjugated to Ca2+-insensitive fluorophores (TAMRA-RBP) to provide an estimate of Ca2+ concentration in RBC synaptic ribbons. We obtained the effective K1/2 (Keff) by measuring the Cal520HA-RBP/ TAMRA-RBP fluorescence ratio in buffered Ca2+ solutions and using the Grynkiewicz equation55. For Cal520HA-RBP, we found the Keff to be ∼795 nM, which is higher than the value of 320 nM reported by the manufacturer for Cal520HA. The differences between the in-cell measurements and the manufacturer’s values are likely to arise from differences in the cellular buffering capabilities, changes in dye properties due to interactions with the molecules inside the cell 36,61-63, and possible differences in the binding properties of the peptide-conjugated Ca2+ indicators when bound to synaptic ribbons. Since the in-cell approach most closely matched the experimental conditions, we used this value (Keff) for all further calculations.

We first measured the ribbon-proximal and ribbon-distal Ca2+ concentrations using Cal520HA-RBP with 0.2 mM EGTA in the patch pipette as it allowed to resolve the gradient between the two signals (Fig. 3C). Under these conditions, we found a maximum ribbon-proximal Ca2+ concentration produced in response to a brief 10-ms pulse of 3.7 µM, and a maximum ribbon-distal Ca2+ concentration of 0.7 µM. These apparent Ca²⁺ concentration amplitudes are well below levels required to trigger exocytosis of the ultrafast releasable pool (UFRP) and readily releasable pool (RRP) 43 and, as discussed earlier, likely represent the lower bounds of the Ca²⁺ concentration at a single ribbon location due to local saturation of the high-affinity indicator and/or due to nanodomains that are smaller than the resolution attainable using light microscopy. It should also be noted that unlike freely diffusing Ca2+ indicators, RBP-conjugated indicators that slowly unbind from the ribbon are not readily replaced by free Ca2+ indicators, rendering them even more prone to saturation. Nevertheless, our data demonstrate that ribbon-bound indicators may better report local Ca2+ concentrations due to their localization, albeit still subject to the limitations of light microscopy.

To test the contribution of local saturation and to better estimate ribbon Ca2+ concentrations, we repeated our measurements using the low-affinity Ca2+ indicator Cal520LA conjugated to the RBP peptide. Because it was difficult to perform in-cell measurements to determine the Keff for Cal520LA-RBP given the large amounts of Ca2+ required to calibrate the Cal 520LA indicator (see Methods and Materials), we used the K1/2 value provided by the manufacturer for free Cal520LA (90 µM). However, we expect that in-cell measurements of Keff, are likely to be different due to the cellular buffering properties reported previously for Keff, measurements of Oregon Green BAPTA-5N in inner hair cells 36.

Bipolar cells release neurotransmitters primarily from ribbon active zones, although some release also occurs at non-ribbon sites (referred to as NR, Fig. 4A) 64-66. To reveal the Ca2+ signaling at and away from the ribbon, we performed whole-cell patch clamping and x-t line scans at ribbon (Fig. 4Ba, R) and non-ribbon (Fig. 4Bb, NR) sites perpendicular to the plasma membrane using TAMRA-RBP and Cal520LA-RBP. A brief (10-ms) depolarizing voltage-clamp pulse evoked rapid high Ca2+ signals (Fig. 4Ba, cyan raster plot) at ribbon locations (Fig. 4Ba, magenta raster plot) but not at non-ribbon locations (Fig. 4Bb, cyan raster plot). The amplitude of Ca2+ signals elicited by brief depolarization and detected with Cal520LA-RBP at the ribbon-proximal site (Fig. 4C, light cyan traces) were significantly higher than those at ribbon-distal (Fig. 4C, dark cyan traces) and non-ribbon (Fig. 4C, blue traces) sites. The Ca2+ concentration gradient along the ribbon is summarized in Fig. 4D. We found that the average Ca2+ concentration at the proximal side of the ribbon (26.4 ± 3.1 µM, n=26) was significantly different from that at the ribbon-distal (15.6 ± 1.5 µM, n=26) and non-ribbon (10.4 ± 0.4 µM, n=15) sites, and that Ca2+ concentrations at ribbon-distal sites is higher than those at non-ribbon sites (Fig. 4D). These measurements display large heterogeneity across distinct ribbons and distinct cells, with the coefficient of variation of about 60% for [Ca2+] measurements at locations proximal to the ribbon, compared to a 10% CV for multiple depolarizations for the same ribbon. The Ca2+ signals at distal sites also reached their peak amplitude earlier (Fig. 4C inset, dark cyan arrowhead) than those at the non-ribbon sites (Fig. 4C inset, blue arrowhead). These findings are consistent with non-ribbon vesicle release being governed by cytoplasmic residual Ca2+ or Ca2+ influx via clustered Cav channels at ribbon sites rather than non-ribbon active zones with clusters of Cav channels and are also in agreement with previous reports regarding the temporal delay and sensitivity of non-ribbon exocytosis to EGTA 65,67.

Ca2+ signals at synaptic ribbon at different distances from the plasma membrane.

A. Ultrastructure of a zebrafish RBC with kinetically distinct vesicle pools, as described in Supplementary Fig. 1 UFRP (green vesicles), and RRP (yellow vesicles) are primarily released via ribbon sites (R) at the cytomatrix of the active zone (arrowhead). Recycling pool (RP, orange vesicles) in the cytoplasm (C), likely to be released via non-ribbon (NR) sites. Scale bar: 500 nm.

B. Representative x-t plots show the fluorescence intensity of Cal520LA-RBP (cyan) and TAMRA-RBP (magenta) as a function of distance (vertical axis) and time (horizontal axis) at (Ba) ribbon sites and (Bb) non-ribbon (NR) locations. The darker region at the upper edge of each plot is the extracellular space and the arrowheads show the timing of depolarization. Scale bars: vertical, 1.6 μm and horizontal, 75 ms.

C. Spatially averaged Cal520LA-RBP as a function of time at ribbon proximal (light cyan), distal (dark cyan), and non-ribbon locations (blue) n=26 (proximal ribbons), 26 (distal ribbons), and 15 (non-ribbon) from 5∼8 different RBCs.

C inset. Temporal profile of events between 0-100 ms were expanded for better visualization. Scale bars: vertical, 0.2 (ΔF/Frest); horizontal, 20 ms.

D. Ca2+ measurements along the ribbon axis using the nanophysiology approach demonstrate the proximal Ca2+ signals can go as high as 26.4 ± 3.1 μM (light cyan, N=26) and distal as 15.6 ± 1.5 μM (dark cyan, N=26), and non-ribbon 10.4 ± 0.4 (NR, blue, N=15) respectively, in response to 10 ms stimuli. Error bars show standard errors.

Sensitivity of microdomain Ca2+ to exogenous buffers

Previous work has shown that the rate of vesicle replacement in RBCs is accelerated by elevated Ca2+ levels at sites along the ribbon, and that while millimolar levels of EGTA have little effect on the fast exocytosis component near Cav channels, they selectively block sustained exocytosis, likely by preventing Ca2+ from reaching the distal locations on the ribbon 11,68-71. These findings raise the possibility that Ca2+ has two sites of action, one near the Cav channels that trigger vesicle release and one further away that replenishes the supply of releasable vesicles. Burrone et al. (2002) proposed that endogenous Ca2+ buffers regulate the size of the RRP by limiting the spatial spread of Ca2+ ions and could suppress vesicle release at the periphery of the active zone in bipolar cells 7. However, the Ca2+ gradient at different locations along the ribbon controlling RRP release has never been examined. Thus, we used our nanophysiology approach (Fig.1) to measure the Ca2+ gradient along the synaptic ribbon under buffering conditions that differentially modulate vesicle release and resupply. To determine the spatiotemporal properties of Ca2+ signals under these conditions, we performed rapid x-t line scans at a single ribbon location in the presence of ribbon-bound Cal520LA-RBP and under varying concentrations of exogenous buffer in the patch pipette solution, including (EGTA at 0.2 (Fig. 5A), 2 (Fig. 5B), and 10 mM (Fig. 5C) or BAPTA at 2 mM (Fig. 5D)). We found that the ratio between proximal and distal Ca2+ signal amplitudes was similar in 0.2 and 2 mM EGTA (proximal-to-distal ratio ∼1.7, Supplementary Table 1) but was enhanced with 10 mM EGTA (proximal-to-distal ratio ∼1.9, Supplementary Table 1) and further enhanced with 2 mM BAPTA (proximal-to-distal ratio ∼4.5, Supplementary Table 1). These findings once again emphasize the reliable measurement of resolving ribbon-proximal vs. ribbon distal Ca2+ signals using our nano-physiology approach. Our experimental findings of the increase in the proximal-to-distal Ca2+ concentration ratio with increasing EGTA concentration are consistent with our simulation results (Fig. 7), although the corresponding ratios are greater in the simulation than in the experiment due to the large size of the microscope’s point-spread function.

Effect of exogenous Ca2+ buffers on spatiotemporal properties of Ca2+ microdomains in RBC terminal recorded with low-affinity ribbon-bound dye.

A-D. Average temporal fluorescence intensity (normalized to ΔF/Frest) of proximal (red), distal (blue), and cytoplasmic (green) Ca2+ signals with Cal520LA-RBP as a function of time at distinct ribbon locations with pipette solution containing (A) 0.2 mM EGTA, (B) 2 mM EGTA, (C) 10 mM EGTA, or (D) 2 mM BAPTA.

Inset. The temporal profiles of events between 0-50 ms were expanded for better visualization. Scale bars: vertical, 2 (ΔF/Frest; panels A and B) and 1 (ΔF/Frest; panels C and D); horizontal, 10 ms.

E. Average temporal fluorescence intensity (normalized to ΔF/Frest) of proximal Ca2+ signals measured with Cal520LA-RBP (cyan) and Cal520LA-free (gray) as a function of time with pipette solution containing 2 mM BAPTA.

Though BAPTA significantly abolished the spread of Ca2+ signals, it is impressive that a substantial amount of ribbon-proximal Ca2+ concentration measured with ribbon-bound indicators was still present with 2 mM BAPTA. One reason for this observation could be that ribbon-bound Ca2+ indicators are likely to measure Ca2+ signals very close to Cav channels without diffusing away, which makes it impossible for BAPTA to intercept Ca2+ ions. If so, similar experiments conducted with free Ca2+ indicators should report a significantly lower proximal Ca2+ signal since the measurement by a diffusible dye would effectively spread the signal over a larger volume. Indeed, this is what we found. As shown in Fig. 5E, the apparent proximal Ca2+ signal in response to a 10-ms brief pulse measured with Cal520LA-free in the presence of 2 mM BAPTA was 0.9 ± 0.2 (n=9), 4-fold lower than proximal Ca2+ signals measured with Cal520LA-RBP (3.6 ± 1, n=20). The latter value is closer to the true measure of the Ca2+ concentration in the vicinity of the ribbon base. These findings emphasize that the reliable measurement of ribbon-proximal Ca2+ signals in the vicinity of Cav channels greatly benefits from the increased spatial resolution of our nano-physiology approach. We also measured the spatiotemporal properties of local Ca2+ signals using Cal520HA-RBP as we have done above for Cal520LA-RBP (Supplementary Fig. 3). We found that the ratio between proximal and distal Ca2+ signals was similar in 0.2 and 2 mM EGTA (proximal-to-distal ratio ∼1.5, Supplementary Table 2) but was enhanced with 10 mM EGTA (proximal-to-distal ratio ∼2.6, Supplementary Table 2) and further enhanced with 2 mM BAPTA (proximal-to-distal ratio ∼2.8, Supplementary Table 2). Cal520 acts as a buffer that binds Ca2+ and carries it away by diffusion. Thus, the higher affinity indicator Cal520HA will bind and shuttle away more Ca2+ than Cal520LA. If the Cal520HA indicator is saturated near the Cav channels, it may be underreporting the fast Ca2+ transients that occur in those locations. However, Cal520 LA and Cal520 HA show similar increase in the ribbon proximal-to-distal ratios with increasing concentrations of exogenous buffer.

Computational models of Ca2+ signals along the axis of the RBC synaptic ribbon

Since we could identify the position of the ribbon and we used RBP-fused Ca2+ indicator-RBP, we were able to measure and distinguish ribbon-proximal vs. ribbon-distal Ca2+ signals that drive the release of UFRP and RRP. However, several factors, including finite spatiotemporal resolution of optical imaging, spatial diffusion, dye saturation, and dye binding kinetics, may limit our ability to achieve optimal resolution. These limitations, however, can be overcome by the use of quantitative models. Thus, we developed a model based on our data with published information on Ca2+ diffusion and buffering to estimate more accurately the [Ca2+]I gradient along the ribbon at distinct distances from the plasma membrane.

Combining whole-terminal Ca2+ current measurements and our estimate for the number of synaptic ribbons per terminal allowed us to infer the magnitude of Ca2+ current per single ribbon, which we used in our model to determine the spatiotemporal [Ca2+] dynamics near the ribbon. Fig. 7 shows the results of the simulation of [Ca2+] at various distances from the ribbon during and after a depolarizing pulse of 10ms duration in the presence of different concentrations of exogenous and endogenous buffers, replicating the conditions used in our experiments. The geometry of the simulation domain box is shown in Fig. 6; it represents the fraction of total synaptic terminal volume per single ribbon (see Methods, Supplementary Table 3 and Supplementary Fig.4A.). Fig. 7 shows the time-dependence of [Ca2+] during and after the depolarization pulse at 5 specific locations both near and distant from the ribbon (left-hand panels in each subplot). These five spatial locations are marked by circles of different colors in the right-hand panel of each subplot, which shows [Ca2+] at the end of the Ca2+ current pulse in pseudo-color in the entire planar cross-section of the simulation domain cutting through the middle of the ribbon, as shown in Fig. 6. We assumed a highly simplified Cav channel arrangement into four clusters forming a square with a side length of 80nm. The closest of the five spatial locations (red curves and circles in Fig. 7) is X=20 nm away from the base of the ribbon center-line, Z=10 nm above the ribbon, about 45nm away from the closest Cav channel cluster (see Fig. 6). We assumed that this location was within the Ca2+ microdomain that triggered the release of the UFRP. Vesicles were not included in the simulation since the total exclusion volume attributed to them represented only a small fraction of the total inter-terminal volume and, therefore, did not significantly impact [Ca2+] at the qualitative resolution level we are interested in. A couple of simulations with vesicles included were performed to confirm this statement, requiring much finer spatial resolution and longer computational time. We note also that [Ca2+] for short pulse durations considered here is relatively insensitive to our assumptions on the membrane Ca2+ extrusion mechanisms, listed in Supplementary Table 3.

Geometry of the computational model of intra-terminal Ca2+ dynamics.

Simulation domain is a box with dimensions (1.28 x 1.28 x 1.1) µm3, approximating the fraction of synaptic terminal volume per single ribbon. Ca2+ ions enters near the base of the ellipsoidal ribbon at four locations marked by black disks representing Ca2+ channels or their clusters. The semi-transparent gray coordinate plane Y=0 corresponds to the section used for the pseudo-color 2D [Ca2+] plots in Figure 7. Ca2+ is extruded on all surfaces of this domain, simulating combined clearance by pumps and exchangers on the plasmalemmal as well as internal endoplasmic reticulum and mitochondrial membranes.

Simulation of the effect of an endogenous immobile buffer of different concentrations on [Ca2+] dynamics in response to a 10ms pulse.

(A1-C2) In each sub-plot, the left-hand panels show the [Ca2+] time course in response to a 10 ms constant current pulse (total current of 1 pA) at five select locations marked by colored circles in the right panel. The right-hand panels show a pseudo-color plot of [Ca2+] in a 2D section of the 3D simulation volume in Fig. 6, at a fixed point in time corresponding to the end of the current pulse. Concentration values for each level curve are indicated in the color bar. Endogenous buffer is immobile, with a concentration of 200 µM in panels A1-C1 (resting buffering capacity 100 µM), vs. 1.44 mM in panels A2-C2 (resting buffering capacity 720). Exogenous buffer concentrations were 0.2 mM EGTA (panels A1, A2), 10 mM EGTA (panels B1, B2), 2 mM BAPTA (panels C1, C2).

Fig. 7A1-B2 reveals that [Ca2+] could rise above 50 µM within the microdomain near the base of the ribbon in the presence of 0.2 mM up to 10 mM EGTA. Given the size of the microscope point-spread function, this is in qualitative agreement with the 26 μM estimate of ribbon-proximal [Ca2+] that we recorded using the ribeye-bound low-affinity indicator. In addition, this concentration level is reached very soon after the channel opening event due to the rapid formation of the Ca2+ microdomain 72,73, and therefore, this estimate is expected to hold for shorter pulses as well. However, [Ca2+] decayed rapidly with distance, with the rate of decay significantly increasing when EGTA concentration was increased to 10mM (Fig. 7B1, B2). Note that a 7-fold change in the capacity of the immobile endogenous Ca2+ buffer (cf. Fig. 7A1 vs. 7A2) had a relatively modest effect on the [Ca2+] level up to about 150 nm distance from the ribbon base. This effect of increasing endogenous buffer concentration was further reduced in 10 mM EGTA (cf. Fig. 7B1 vs. 7B2), due to strong competition of endogenous buffer with large concentrations of EGTA. This agrees with the expectation that immobile buffers primarily slow down Ca2+ signals but do not affect the rapidly forming quasi-steady-state Ca2+ microdomains in the immediate vicinity of the channel 72-74. Finally, Fig. 7C1-C2 shows that 2 mM BAPTA had a much greater effect on localizing the Ca2+ signal to the immediate vicinity of the ribbon base.

Since a native unpatched cell may well contain mobile rather than immobile buffers, it was interesting to examine the effect of increasing the concentration of mobile endogenous buffer on Ca2+ dynamics in the ribbon vicinity, as compared to the corresponding effect of immobile buffer. Supplemental Figure 4B-C shows that a 7-fold increase in mobile buffer concentration reduces the microdomain Ca2+ by a factor of 2 and greatly reduces the size of the microdomain. This contrasts with the effect of increasing the concentration of the immobile buffer, which has a much more subtle effect (cf. Fig. 7A1, A2). We note that 0.2 mM EGTA was absent in the simulation with mobile buffer shown in Supplemental Figure 4B-C; in general, the effect of modest concentrations of EGTA is expected to be negligible due to its slow Ca2+ binding speed compared to the endogenous buffer.

Variability of local Ca2+ signals across the RBC synaptic ribbons

Bipolar cells release neurotransmitters primarily from ribbon active zones 1,6,8,11-13,21-25. However, the factors that shape the synaptic ribbon microdomains at retinal ribbon synapses have not been examined. We wondered whether all of the 30∼50 synaptic ribbons at an RBC terminal release glutamate in a similar fashion or whether there is some heterogeneity in glutamate release from different ribbon active zones. In particular, we asked what underlying mechanisms could differentiate the Ca2+ signals between ribbons of the same RBC terminal and across different RBCs. We first found evidence for such variability in local Ca2+ signals at single-ribbon locations of different cells using the freely diffusible high-affinity indicator Cal520HA, observing in particular high variability of the Ca2+ transient amplitudes, even for RBCs with similar depolarization-evoked Ca2+ current amplitudes (cf. Supplementary Figs. 5A vs. 5B. However, we did not observe such variability when we obtained multiple line scans across the same ribbon (Supplementary Figs. 5A vs. 5B, gray traces). We wondered whether the observed variability between cells could be due to distinct subtypes of zebrafish RBCs 47, which might have different subtypes or numbers of Cav channels near the ribbon. Recent work in zebrafish retina identified two distinct RBC subtypes: RBC1 and RBC2 47. The wiring pattern of amacrine cells postsynaptic to RBC1 closely resembles the circuitry of mammalian RBCs, whereas RBC2 forms distinct pathways. Furthermore, RBC1 specifically labels for the known marker of mammalian RBCs, PKC-α 47. Due to these similarities, RBC1 in zebrafish is classified as analogous to mammalian RBCs. Moreover, RBC1s have expected morphological characteristics, in particular, the shape and size of soma and the presence of a single synaptic terminal 48. Immunolabeling of isolated RBC preparation used in these experiments showed primarily intact RBCs, which are PKC-α-specific RBC1 (Supplementary Fig. 2B). For simplicity, and because this study focuses exclusively on RBC1, we will refer to RBC1 as ‘RBC’ throughout this work. Thus, we attribute the variability in the RBC Ca2+ transients to the variability between the Ca2+ microdomains within the same cell type.

To identify mechanisms that contribute to the heterogeneity in the local Ca2+ signals we have reported in this study, we began by asking whether differences in local Ca2+ buffering could account for this heterogeneity. To examine the role of Ca2+ buffering in the variability of local Ca2+ signals across different cells and different ribbons (Figs. 8A and 8D, respectively), we compared proximal (Figs. 8B and 8E) and distal (Figs. 8C and 8F) Ca2+ signals in 10 mM EGTA by averaging the Ca2+ recordings from all the ribbons of a given cell (Supplementary Figs. 6 and 7). When the Ca2+ signals were restricted to the ribbon sites with 10 mM EGTA in the pipette solution (Supplementary Tables 1 and 2), the amplitude of the ribbon-proximal and ribbon-distal Ca2+ signals averaged over all ribbons of a given cell exhibited significant variation across different cells (Figs. 8B, 8C, and Supplementary Fig. 6). We next compared variability in Ca2+ signals in the presence of 10 mM EGTA at individual ribbons within the same cell (Fig. 8D and Supplementary Fig. 7) at proximal (Fig. 8E) and distal (Fig. 8F) locations. For the cell described in Figs. 8D-8F, the proximal Ca2+ signals were significantly different across all ribbons examined, and there were considerable differences in distal Ca2+ signals between ribbons, with the exception of ribbons numbered 2 and 5. These findings suggest that the heterogeneity in proximal and distal Ca²⁺ signals at distinct ribbons within the same cell may result from different underlying mechanisms, for example, heterogeneity in Cav expression or subtype and Ca2+ handling mechanisms. These findings suggest that exogenous Ca2+ buffering has a negligible effect on experimentally observed heterogeneity and variability of the proximal Ca2+ signals and that local Ca2+ signals at RBC ribbons are dominated by Ca2+ in regions close to the ribbon base where Cav channels are located.

Heterogeneity in the spatiotemporal properties of Ca2+ microdomains in RBC terminal.

A. Top panel. Cartoon of two representative RBCs (cell A and cell B), each containing differently colored ribbons. Bottom panel. Ribbon-to-ribbon variability was measured by recording local Ca2+ signals near different ribbons (yellow and purple traces) for each RBC. If a cell had multiple readings for a single ribbon, averages were obtained for comparisons as described in Supplementary Fig. 6.

B-C. Variability of Ca2+ signals in different ribbons across different cells in the presence of 10 mM EGTA, recorded using Cal520LA-RBP. Proximal Ca2+ amplitude values were significantly different between cells A and B (p = 0.022), cells A and D (p = 0.023), and cells B and D (p = 0.005) but similar between A and C (n = 4 cells, 4 fish).

(C) In distal locations, ribbon amplitude values were significantly different between cells A and B (p = 0.031), but similar across all other cell comparisons (n = 4 cells, 4 fish).

D. Top panel. Illustration of a RBC containing three ribbons (numbered 1-3). Bottom panel. Ca2+ signal measurements from three distinct ribbons (black, gray, and blue traces) were compared to determine the ribbon-to-ribbon variability within each RBC, as described in Supplementary Fig. 7.

E-F. Box plot illustrating [Ca2+] across various ribbons of an individual RBC. Ribbon variability within individual cells was measured with 10 mM EGTA using Cal520LA-RBP at (E) proximal and (F) distal locations. (E) Proximal Ca2+ amplitude values were significantly different among all ribbons (p < 0.001) (n = 5 ribbons, 1 RBCs, 1 fish). (F) Distal Ca2+ amplitudes were significantly different among all ribbon comparisons (p < 0.001) except for ribbons 2 and 5 (n = 5 ribbons, 1 RBCs, 1 fish). Similar analyses were conducted in two more cells and found similar observations (data not shown).

The ultrastructure of the zebrafish RBC terminal reveals diversity in the size of the synaptic ribbons across the terminal

In hair cells, Ca2+ microdomain signaling varies with ribbon size, reflecting larger patches of Cav channels aligned with larger ribbons 53. However, the number, size, and shape of ribbon active zones per terminal have not been established in our experimental system, the synaptic terminals of zebrafish RBCs. To provide quantitative measurement of ribbon microdomains, we examined the ultrastructure of zebrafish RBC synaptic ribbons by serial block face scanning electron microscopy (SBF-SEM) and reconstructed three RBCs from serial section electron micrograph (Fig. 9A). We analyzed the bipolar cell terminals that are closest to the ganglion cell layer with a morphology similar to mammalian RBCs and zebrafish RBC1s 47. Our reconstruction of three RBCs revealed 30-41 ribbons within RBC terminals (Fig. 9A), similar to what was previously reported in goldfish giant ON-type mixed RBCs (range 45-60 75).

Distribution of measured distance between synaptic ribbons in three zebrafish RBCs.

A. Reconstruction of three RBC terminals closest to the ganglion cell layer resembling the shape and size of the mammalian RBC1. The ribbons are shown in magenta. Note that the total number of ribbons includes the “floating” ribbons detached from the plasma membrane. RBC II contained 7 floating ribbons. Scale bar, 5 µm. RBCs, rod bipolar cells.

B. Overview of the distance measurements. Top. 3D rendering of a single RBC terminal is shown in light green with ribbons in magenta. Black lines show an example of the different SBF-SEM layers that are cut to obtain individual images. Bottom. Four sample SBF-SEM layers are shown with two example ribbons (magenta) located near each other but in different layers. If two ribbons were on the same layer, their linear distance was taken, whereas if they were located in different layers, their linear distance and height difference were used to calculate their actual distance using the Pythagorean Theorem. Each SBF-SEM layer has a thickness of 50 nm. A 3D volume movie of the synaptic terminal of a zebrafish bipolar neuron synaptic ribbon distribution and measurement is provided in Video 1.

C. Histograms showing the ribbon distance for the three RBCs. x0 shows the mean of the distribution.

D. Histogram of the distances between each ribbon and its five nearest ribbons for all ribbons contained in the three RBCs. x0 shows the mean of the distribution. Please note that the y-axis for B and C have different sizes, given that they represent the number of occurrences of each event.

We examined the distribution of these 30–41 synaptic ribbons within the zebrafish RBC terminals. The distribution of RBC ribbons as estimated by the distance between ribbon sites (Fig. 9B and Video 1; see Materials and Methods) revealed a wide distribution in synaptic ribbon distance to its nearest neighbor across the three reconstructed RBCs (means of 2.48 ± 0.12 µm, 1.84 ± 0.06 µm and 1.98 ± 0.08 µm; Fig. 9C). The distance between RBC synaptic ribbons ranged from 141 nm to 6 µm, with a mean of 2 ± 0.05 µm (Fig. 9D; across all three reconstructed RBCs). Of the 510 zebrafish RBC synaptic ribbons, only 4 (0.8%) were separated by distances smaller than our confocal microscope resolution limit of 270 nm.

To reveal whether heterogeneity in synaptic Ca2+ signals correlates with active zone size, we compared the shape and size of the ribbon active zones across the three reconstructed RBCs. Our results show significant variations in the shape and size of zebrafish RBC synaptic ribbons in RBCs and associated active zones (Fig. 10). On average, individual ribbons spanned 2–5 consecutive sections, with some located within the axons or closer to the terminal. The SBF-SEM images and 3D projections of ribbon structures (colored in magenta) revealed considerable variability in shape and size across all dimensions, as illustrated in Figs. 10A–C. The number of Cav channels per active zone in hair cells from chicken, frog, and turtle varies with the size of the synaptic ribbon 76. We thus measured the area of the ribbon facing the plasma membrane where the Cav channels are known to be located to estimate the number of Cav channels per active zone and to determine whether variation in active zone size might plausibly contribute to the heterogeneity in Ca2+ signaling. The EM images and 3D projection of the plasma membrane (yellow) and the area of the ribbon facing the plasma membrane (cyan), representing the active zone, are shown in Figs. 10A-C. We observed large variability in the area of the ribbon facing the plasma membrane or active zone within the RBCs, with substantial variability in their average distribution across the three RBCs (Figs. 10C, cyan, 10E, Video 2 and Supplementary Fig. 9), suggesting that the number of Cav channels per RBC active zone could plausibly be heterogeneous.

Serial block-face scanning electron microscopy analysis reveals heterogenous RBC ribbon shape, size and area of the ribbon facing the plasma membrane.

A-C. EM images of zebrafish RBC ribbon structures (A) and their respective 3D reconstructions to illustrate different shapes and sizes of synaptic ribbons (A, B, and C, magenta), plasma membrane (A and B, yellow), and the area of the ribbon facing the plasma membrane (A, B, and C, cyan). Each of the 5 rows of images illustrates one ribbon synapse from a zebrafish retinal RBC. A 3D reconstruction of the RBC synaptic terminal and ribbon from SBF-SEM stacks is provided in Video 2.

D. 3D reconstruction of the RBC terminal closest to the ganglion cell layer resembles the shape and the size of the mammalian RBC1 2. The ribbons are colored magenta.

E. Summary of the active zone size, the area of the ribbon associated with the plasma membrane measured in serial sections across the three RBCs from Fig 9A. The individual distribution of the three RBCs active zone sizes is provided in Supplementary Fig. 8. The z-size of the SBF-SEM sections is 50 nm, and each ribbon spans 2-5 consecutive sections. The solid cyan circles in the violin plots show individual synaptic ribbon measurements from three RBCs, with the average measurements shown in the solid black circle.

The size of the active zone and maximal Ca2+ influx correlate with the size of the synaptic ribbon

To reveal whether larger ribbons display stronger maximal Ca2+ influx, we measured Ca2+ signals in response to a series of 200-ms depolarizations. We first imaged TAMRA-RBP and Cal520LA-RBP in sequential scans to obtain ribbon location and resting Ca2+ levels. To maximize the capture of Ca2+ signals during brief stimuli, we image the Cal520LA-RBP channel, followed by TAMRA-RBP and Cal520LA-RBP sequential scans to confirm the ribbon locations. We found that the maximum amplitude of depolarization-evoked Ca2+ signals increased with an increase in ribeye fluorescence (Fig. 11A & B and Video 3; r = 0.51, N = 121 ribbons, p< 0.001), consistent with findings reported in cochlear inner hair cells 53,58. Since ribeye fluorescence correlates with the number of ribeye molecules per ribbon, and, therefore, ribbon size 26,53,58,77, our findings suggest that larger ribbons display stronger Ca²⁺ signals as active zone size scales with ribbon size 58. Given the diverse shapes of RBC synaptic ribbons (Fig. 10A & B, magenta), we measured the longest length and width of ribbons from EM images and compared this with the active zone size. We found that RBC ribbon length and width have a positive correlation (Fig. 11C, r = 0.47, N = 102 ribbons, p< 0.001), and both dimensions have a positive correlation with active zone size, albeit the ribbon width (Fig. 11D, r = 0.52, N = 102 ribbons, p< 0.001) has a stronger correlation with active zone size than the ribbon length (Fig. 11D, r = 0.32, N = 102 ribbons, p< 0.001). These results suggest that heterogeneity in synaptic Ca2+ signals correlates with ribbon dimensions and that active zone and the number of Cav channels 53,58 scale with ribbon size.

Heterogeneity of Ca2+ microdomains in RBC terminals.

Larger ribbons have stronger maximal Ca2+ influx and larger CAZ. A. Images show voltage-clamped RBC filled with a solution containing TAMRA-RBP to label ribbons before depolarization (left, magenta) and Cal520LA-RBP to measure the amplitude of Ca2+ influx (middle, cyan) during depolarization and superimposed of the two (right) to compare the size or ribbon vs. maximal Ca2+ influx. A time-lapse movie of the synaptic terminal of a zebrafish bipolar neuron during Ca2+ influx is provided in Video 3. B. Scatter plot of maximal Ca2+ influx (F/Frest) vs. TAMRA-RBP. Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. C. Scatter plot of maximal synaptic ribbon length vs. maximal synaptic ribbon width as estimated from SBF-SEM images. Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. D. Scatter plot of area of AZ vs. ribbon width (filled black circles) and of area of AZ vs. ribbon length (filled grey circles). Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. AZ, Active Zone.

Discussion

A deep look at the active zone Ca2+ microdomain in RBC terminals

In this study, we developed a quantitative nanophysiology approach using ribeye-bound Ca²⁺ indicators and fluorescently labeled RBP, combined with ratiometric dual-color imaging, to measure changes in Ca²⁺ concentration at and along the ribbon axis. By targeting Ca²⁺ indicators to the ribbon combined with fluorescently labeled RBP, we determined the Ca2+ concentration along the ribbon at different distances from active zone, and found that near the ribbon base the Ca2+ concentration could increase to an average of 26 μM upon depolarization. Given the localization of our indicator on the ribbon and minimal sensitivity to exogenous buffers, we believe that our ribbon-proximal signal represents Ca2+ concentrations on the ribbon in locations very near release sites, likely some of those that contribute to vesicle resupply. Notably, the [Ca2+] i levels obtained from our nanophysiology approach using confocal microscopy are similar to those that were reported for measurements in hair cells using Stimulated Emission Depletion (STED) lifetime measurements 36. Previous work on RBCs found slowing of recovery from paired-pulse depression by EGTA, suggesting that Ca2+ levels at distal sites on the ribbon might be important for the recruitment of new vesicles following depletion of the RRP and UFRP 11,24. Here, we measured directly the effect of EGTA on the spread of Ca2+ signals and found that the Ca2+ signal at the distal half of the ribbon was two-fold lower than that at the location proximal to the membrane under conditions of low Ca2+ buffering. It should be noted, however, that diffraction of light during the optical imaging procedure is likely to underestimate spatiotemporal [Ca2+] along the ribbon. Therefore, we used our measurement of the Ca2+ current per ribbon to develop a detailed computational model that allowed us to resolve the expected profile of Ca2+ microdomains governing UFRP fusion. The model results shown in Fig. 7 reveal, on a finer spatial scale, the distribution of Ca2+ around a ribbon in the presence of various quantities of immobile endogenous buffer and added exogenous buffers. Our finding of high proximal [Ca2+] levels on the order of 26 µM recorded using a ribeye-bound low-affinity indicator dye is in qualitative agreement with the simulated high concentration of Ca2+ in the microdomain at the base of the ribbon. Simulations also show that, as expected, the peak Ca2+ microdomain amplitude is relatively insensitive to the concentration of endogenous immobile buffer and added EGTA, but its decay with distance is accelerated by increasing the concentrations of exogenous and endogenous buffers. In contrast, the addition of 2 mM BAPTA causes a profound reduction of the Ca2+ microdomain amplitude and localizes the signal to a small volume around the ribbon. We also compared the impact of mobile vs immobile endogenous buffers on the Ca2+ distribution, showing the much greater effect of mobile buffers on the microdomain amplitude and distance dependence. Although these results are generally expected72-74, the simulation results estimate the [Ca2+] levels in the vicinity of the ribbon with greater spatial and temporal resolution. As expected, the ratios of Ca2+ concentration values at the base of the ribbon vs. further away from the ribbon are much greater in the model simulation than our experimentally determined proximal-to-distal Ca2+ ratio values due to the large size of the microscope PSF, which precludes precise localization of the corresponding Ca2+ signals. Further, simulations show the effect of exogenous buffers with greater spatial resolution. However, the simulations fully agree with the overall increase of the proximal-to-distal Ca2+ concentration ratios by the application of exogenous Ca2+ buffers, especially BAPTA. Despite the technical challenges posed by the limited spatiotemporal resolution of optical imaging, the spatial resolution attained with the nanophysiology approach developed in our study is sufficient to differentiate Ca²⁺ signals associated with distal sites on the ribbon from those localized near the plasma membrane. Conceivably, the nanophysiology approaches established here could generally be applied to other classes of ribbon-containing cells, such as rods, cones and hair cells.

Heterogeneity of Ca2+ microdomains across RBCs terminals

Our quantitative nanophysiology provides evidence for the heterogeneity in synaptic Ca2+ signals across zebrafish RBCs, suggesting variability in the Ca2+ microdomains as has been previously reported for hair cells 53,58 and cone photoreceptors. Retinal bipolar neurons contain many ribbon active zones. This ranges from 45-60 in adult goldfish, as quantified by EM 75, or 25-45 in RBCs of adult living zebrafish, as estimated from laser scanning confocal microscopy78. Our analyses of ribbon number across zebrafish RBCs through SBF-SEM arrived at an estimate of 31-40. The differences in the number of ribbons as determined through confocal microscopy compared to SBF-SEM could be explained by the resolution limit encountered by the confocal image analyses. Our analyses of zebrafish RBC terminal SBF-SEM images and 3D projections provide the first evidence of diversity in synaptic ribbon shape and sizes across zebrafish RBC terminals. Further, our quantitative measurements of SBF-SEM images reveal that synaptic ribbon width and length correlate with the active zone size, consistent with previous findings in hair cell ribbon synapses 76,77 and cone photoreceptor ribbon synapses54. Our observation that maximal active zone Ca²⁺ influx correlates with ribeye fluorescence in live imaging is consistent with findings previously reported in hair cells 58. Thus, we expect larger active zones, with more Ca2+ influx, to have more Cav channel expression, as demonstrated for hair cell ribbon synapses 58. Heterogeneity among ribbon active zones has been previously reported in hair cells where synaptic Ca2+ microdomains varied substantially in amplitude and voltage-dependence within individual inner hair cells 53,58. Studies in aquatic tiger salamander cone photoreceptors revealed that the amplitude and midpoint activation voltage of Ca²⁺ signals varied across individual ribbons within the same cone. Additionally, local Ca²⁺ signal dynamics at cone photoreceptor ribbons were found to be independently regulated 54.

The faster, smaller, and more spatially confined Ca²⁺ signals that are insensitive to the application of high concentrations of exogenous Ca²⁺ buffers, referred to here as ribbon proximal Ca²⁺ signals, could be due to Ca²⁺ influx through Cav channel clusters beneath the synaptic ribbon 19,26,29-42. However, the variability in Ca²⁺ signals at the distal ribbon, away from the plasma membrane, may result from the spread between two closely spaced microdomains; thus, if proximal Ca²⁺ signals are variable, the distal Ca²⁺ signals are also variable between ribbons. Alternatively, the variability in distal Ca²⁺ signals could be due to additional mechanisms, for example, Ca²⁺ influx from internal stores. Future studies should aim to elucidate the specific mechanisms underlying the observed heterogeneity in distal Ca²⁺ signals. For instance, exploring differences in Ca²⁺ release from intracellular stores or variations in Ca²⁺ sequestration, as reported in goldfish, could reveal key contributing factors 79. Although the potential impact of these intracellular Ca²⁺ stores on Ca²⁺ microdomain heterogeneity in hair cells has not been reported,53 previous studies in goldfish bipolar cell terminals have documented spatially restricted Ca²⁺ oscillations in voltage-clamped retinal bipolar cells, occurring independently of membrane potential 20. To our knowledge, the sources of Ca²⁺ for these oscillations remain unknown. However, previous studies in goldfish bipolar cells suggest that the endoplasmic reticulum79 and mitochondria18 may act as internal Ca2+ stores. Notably, IP₃ receptors have been localized to retinal bipolar cell terminals in some species.80,81

In the current study, we investigated possible mechanisms for heterogeneity in proximal Ca²⁺ signals by measuring the ribbon size, particularly the area of the ribbon adjacent to the plasma membrane, where Cav channel clusters are located. Our SBF-SEM shows substantial variability in synaptic ribbon size, shape, and the area of the ribbon facing the plasma membrane where Cav channel clusters are tethered. Since both Ca²⁺ signals and the area of the synaptic ribbon facing the plasma membrane are heterogeneous, we propose that the larger ribbons may anchor more channels, leading to larger Ca²⁺ microdomains. Since local Ca2+ signals control kinetically distinct neurotransmitter release components, heterogeneity in local Ca2+ signals may alter the rate of vesicle release and allow them to function independently. Indeed, heterogeneity in neurotransmitter release kinetics has been proposed for the observed diversity in excitatory postsynaptic current amplitudes and kinetics reported from paired recordings of goldfish bipolar cells82 and ON-type mixed RBCs 83. The heterogeneity in the RBC Ca²⁺ microdomains, synaptic ribbon shape, size, and active zone area reported in this study may contribute to regulating the dynamic range of RBC ribbon synapses22,84,85 - a hypothesis that needs to be tested in future studies.

Methods

Rearing of zebrafish

Male and female zebrafish (Danio rerio; 16 ∼20 months) were raised under a 14 h light/10 h dark cycle and housed according to NIH guidelines and the University of Tennessee Health Science Center (UTHSC) Guidelines for Animals in Research. All procedures were approved by the UTHSC Institutional Animal Care and Use Committee (IACUC; protocol # 23-0459).

Isolation of zebrafish retinal RBCs

Dissociation of RBCs was performed using established procedures 86. Briefly, retinas were dissected from zebrafish eyes and incubated in hyaluronidase (1100 units/ml) for 20 minutes. The tissue was washed with a saline solution containing 120 mM NaCl, 2.5 mM KCl, 0.5 mM CaCl2, 1 mM MgCl2, 10 mM glucose, and 10 mM HEPES, pH = 7.4] before being cut into quadrants. Each quadrant was incubated at room temperature for 25 to 40 min in the same saline solution, to which was added DL-cysteine and papain (20-30 units/ml; Sigma Millipore, St. Louis, MO) and triturated using a fire-polished glass Pasteur pipette. Individual cells were transferred to glass-bottomed dishes, allowed to attach for 30 minutes, and washed with saline solution before being used for experiments.

Ribeye binding peptides

As a means of localizing the ribbons, custom peptides containing the ribbon binding sequence fused to TAMRA (tetramethylrhodamine; TAMRA: GIDEEKPVDLTAGRRAG) dye were synthesized, purified, and purchased from LifeTein (>95% purity, LifeTein LLC, NJ).

Ca2+ indicators

Free Ca2+ indicators. The potassium salts of high and low-affinity Ca2+ indicators Cal-520® (high affinity, KD 320 nM) and Cal-520N™ AM (low affinity, KD 90 µM), referred to as Cal520HA and Cal520LA, respectively were purchased from AAT Bioquest.

Direct conjugation of Ca2+ indicator to cysteine-containing ribbon-binding peptides. To target Ca2+ indicators to the ribbon, custom-made cysteine-containing ribeye binding peptides NH2-CIEDEEKPVDLTAGRRAC-COOH were synthesized and purchased from LifeTein to directly fuse with the fluorogenic 520® maleimide (purchased from AAT Bioquest) for high affinity (HA) and low affinity (LA Ca2+ indicator dyes). Each peptide at one mM concentration was mixed and incubated with two mM Cal-520® maleimide (20 mM stock solution in DMSO purchased from AAT Bioquest) for one hour at room temperature, then overnight at 4°C. Calibration of Ca2+ indicator dyes was performed as described 19 and as detailed below. The conjugated Ca²⁺ indicators were stored at -20°C in smaller aliquots, each sufficient for a single day’s experiments.

Measurement of dissociation constants (KD) for Ca2+ indicator peptides

The effective KD (Keff) was obtained by measuring the fluorescence of Cal520HA-RBP and TAMRA-RBP in buffered Ca2+ solutions, determining the ratio between them 19, and using the Grynkiewicz equation to determine the Ca2+ concentration [Ca2+] from this ratio 55. However, this was not possible for the low-affinity indicator Cal520LA due to the large Ca2+ levels required to calibrate it. Thus, Keff for Cal520LA-RBP could be larger than the K1/2 provided by the manufacturer for Cal520LA (KD, 90 μM), as reported previously with Keff, measurements of OB-5N in inner hair cells 36. Thus, our estimates of local Ca2+ concentrations obtained using Cal520LA represent the lower bounds of the underlying true values. As noted in the Results section, the same may be true for the high-affinity Cal520HA despite its accurate Keff estimate, due to potential dye saturation effects.

RBC voltage clamp recording

Whole-cell patch-clamp recordings were made from isolated RBCs, as described previously 19,27,87. Briefly, a patch pipette containing pipette solution (120 mM Cs-gluconate, 10 mM tetraethyl-ammonium-Cl, 3 mM MgCl2, 0.2 mM N-methyl-d-glucamine-EGTA, 2 mM Na2ATP, 0.5 mM Na2GTP, 20 mM HEPES, pH = 7.4) was placed on the synaptic terminal, as described previously. The patch pipette solution also contained a fluorescently-labeled RBP peptide (TAMRA-RBP) to mark the positions of the ribbons and either 1) free Ca2+ indicators Cal520HA-free (Figs. 1 and 2) and Cal520LA-free (Fig. 2, 3 and 5) to demonstrate our nanophysiological approach or 2) ribeye-bound Ca2+ indicators Cal520HA-RBP (Figs. 3 and Supplementary Figs. 3 and 5) and Cal520LA-RBP (Figs. 3, 4, 5, and 8) to measure local ribbon-associated Ca2+ signals. Current responses from the cell membrane were recorded under a voltage clamp with a holding potential (VH) of -65 mV that was stepped to 0 mV (t0) for 10 milliseconds. These responses were recorded with a patch clamp amplifier running PatchMaster software (version v2x90.4; HEKA Instruments, Inc., Holliston, MA). Membrane capacitance, series conductance, and membrane conductance were measured via the sine DC lock-in extension in PatchMaster and a 1600 Hz sinusoidal stimulus with a peak-to-peak amplitude of 10 mV centered on the holding potential 88.

Acquisition of confocal images

Confocal images were acquired using an Olympus model IX 83 motorized inverted FV3000RS laser-scanning confocal microscopy system (Olympus, Shinjuku, Tokyo, Japan) running FluoView FV31S-SW software (Version 2.3.1.163; Olympus, Center Valley, PA) equipped with a 60 X silicon objective (NA 1.3), all diode laser combiner with five laser lines (405, 488, 515, 561 & 640 nm), a true spectral detection system, a hybrid galvanometer, and a resonant scanning unit. Fluorescently labeled ribeye binding peptide (RBP) 26 and Ca2+ indicator were delivered to RBC via a whole-cell patch pipette placed directly at the cell terminal. We waited for 30 s after break-in to allow Cal520HA to reach equilibrium with the patch-pipette before obtaining the first fluorescence image. Rapid x-t line scans at the ribbon location were performed to localize synaptic ribbons (Fig.1Aii) and to monitor local changes in Ca2+ concentration at a single ribbon, as we demonstrated previously, to estimate the Ca2+ levels at the plasma membrane and to track a single synaptic vesicle at ribbon locations 19,20. The z-projection from a series of confocal optical sections through the synaptic terminal (Fig. 1Ai) illustrates ribbon labeling (magenta spots). RBP fluorescence was used to localize a synaptic ribbon and to define a region for placing a scan line perpendicular to the plasma membrane, extending from the extracellular space to the cytoplasmic region beyond the ribbon to monitor changes in the Ca2+ concentration along the ribbon axis (Fig. 1Aii). The focal plane of the TAMRA-RBP signal was carefully adjusted for sharp focus to avoid potential errors arising from the high curvature near the top of the terminal and the plane of membrane adherence to the glass coverslip at the bottom of the terminal. Sequential dual laser scanning was performed at rates of 1.51 milliseconds per line. Two-color laser scanning methods allowed observation of Ca2+ signals (Fig.1C, cyan) throughout the full extent of the ribbon in voltage-clamped synaptic terminals, while the ribbon and cell border were imaged with a second fluorescent label. The Exchange of TTL (transistor-transistor logic) pulses between the patch-clamp and imaging computers synchronized the acquisition of electrophysiological and imaging data. The precise timing of imaging relative to voltage-clamp stimuli was established using PatchMaster software to digitize horizontal-scan synced pulses from the imaging computer in parallel with the electrophysiological data (Fig.1B). Acquisition parameters, such as pinhole diameter, laser power, PMT gain, scan speed, optical zoom, offset, and step size were kept constant between experiments. Sequential line scans were acquired at 1-2 millisecond/line and 10 μs/pixel with a scan size of 256 × 256 pixels. Bleed-through between the channels was confirmed with both lasers using the imaging parameters we typically use for experiments. To test bleed-through from the RBP channel (TAMRA-RBP) to the Ca2+ indicator (Cal-520HA and Cal-520LA), whole-cell recordings from RBC terminals were performed with patch pipette solution that contained TAMRA-RBP or the aforementioned Ca2+ indicators, and line-scan images were collected and analyzed using the same procedures used for experimental samples.

Point spread function

The lateral and axial point spread function (PSF) was obtained as described previously 19,28,86. Briefly, an XYZ scan was performed through a single 27 nm bead and the maximum projection in the xy-plane was fit to the Gaussian function using Igor Pro software. We obtained the full width at half maximum (FWHM) values for x and y-width of 268 and 273 nm, respectively, in the lateral (x-y plane) and for y-z width of 561 nm in the axial (y-z axis) resolution.

Photobleaching

We minimized photobleaching and phototoxicity during live-cell scanning by using fast scan speed (10 microsec/pixel), low laser intensity (0.01–0.06% of maximum), and low pixel density (frame size, 256 × 256 pixels). We estimated photobleaching using x-t line scans of Cal520HA or Cal520LA and TAMRA-RBP in the absence of stimulation, with the same imaging parameters used for experimental samples.

Data analysis

Quantitative FluoView x-t and x-y scans were analyzed initially with ImageJ software (imagej.nih.gov) and subsequently with Igor Pro software (Wavemetrics, Portland OR) for curve fitting and production of the figures. Data from PatchMaster software were initially exported to Microsoft Excel (Version 16.81) for normalizing and averaging and exported from MS Excel to Igor Pro (Version 9.05) for curve fitting and production of the l figures.

Analysis of x-t scan data

X-axis profile. To determine the Ca2+ signals along the ribbon axis, we spatially averaged the x-axis profile intensity of RBP (i.e., a horizontal row of pixels, see Fig.1C, top) to determine the position of the center of the ribbon and estimate the location of the plasma membrane. The parameter x0 is the peak of the Gaussian fit, giving the x-position of the center of the ribbon. The x-axis profile is also used to obtain the spatial profile of Ca2+ signals before, during, and after with respect to the ribbon profile. To identify the Ca2+ signals specific to the ribbon location, we fit x-axis intensity profiles with the equation f(x) = s(x) + g(x), where s(x) is a sigmoid function that describes the transition from intracellular to extracellular background fluorescence at the edge of the cell, given by s(x) = b – c / (1–exp((x1/2 – x)/d)), and g(x) is a Gaussian function that represents the fluorescence of RBP, given by g(x) = a exp(-(x-x0)2/w2), as described 27. The parameters x1/2 and x0 were taken as the x-axis positions of the plasma membrane and the fluorescence emitter, respectively. While parameter b is intracellular background fluorescence, c is extracellular background fluorescence, d is the slope of the sigmoid, a is the peak amplitude of emitter fluorescence, and w is √2* the standard deviation of the Gaussian function, in practice, the latter parameters were highly constrained by the data or by the measured PSF, leaving only x1/2 and x0 as free parameters in the fitting. Fig.1D, demonstrates that the peak of the Ca2+ signals (x0, cyan; Fig.1D) during the stimuli is proximal to the ribbon center (x0, magenta; Fig.1D) towards the plasma membrane (x1/2, magenta; Fig.1D), as expected for Ca2+ influx originating from Ca2+ channels localized in the plasma membrane 44-46.

T-axis profile. The temporal profiles of the Cal520, Cal520-RBP, and TAMRA-RBP signals were determined by analyzing the time-axis profile of the x-t line scan to obtain the kinetics of the Ca2+ transient with respect to the ribbon. We determined the baseline kinetics by averaging the fluorescence obtained immediately before depolarization. The timing of depolarization and the amplitude of the Ca2+ current was obtained from the PatchMaster software. The rising phase of the Ca2+ transient was fit with the sigmoid function, the peak of which is referred to as the peak amplitude Ca2+ transient.

We used this baseline profile of ribbon-proximal Ca2+ signals to distinguish signals proximal or distal to the ribbon despite the distance between these two signals being within the PSF. For example, the ribbon-proximal signals were obtained by averaging 5 pixels of the temporal profile of the Ca2+ signals (obtained with Cal520 or Cal520-RBP) between the x1/2 and x0 values obtained in the x-axis profile of TAMRA-RBP. The distal profile was obtained similarly but was 5-pixels after x0 towards the cytoplasm.

Quantifying the rise-time of the Ca2+ transients. The rise-time of the fluoresence transients (Figs. 2A and 2B) was defined as the time to reach half-maximal value from the start of the stimulation, after subtracting its baseline. It was determined by fitting a hyperbolic function with an argument shifted by a constant value β, which serves as one of two free parameter of the fit, along with the argument scale factor α characterizing the steepness of the rising phase of the signal: The constant C = tanh(α*β) ensures that f(0) = 0 at the start of the stimulation, t = 0. This fitting function reaches its peak value Fpeak at t = tpeak, as determined from the averaged fluorescence trace. The sign of the argument shift β controls the shape of the signal near t = 0: β < 0 corresponds to a signal which is concave downward (i.e. very rapid onset), while β > 0 corresponds to a signal which is concave upward near t = 0 (i.e. gradual rise). Therefore, this function provides a more flexible functional relationship than the Hill function or the standard sigmoidal function.

Analysis of x-y scan data

We analyzed the rate of loading the TAMRA-RBP and Ca2+ indicator into the RBC terminal with ImageJ software by placing a square region of interest (ROI; 5 x 5 μm) and using a Plot z-axis profile function to obtain spatially averaged TAMRA-RBP and Ca2+ indicator fluorescence as a function of time. The rising phase of TAMRA-RBP and Ca2+ indicator fluorescence was obtained by fitting the rising to the peak to the single exponential function using the curve fitting function in Igor Pro 9 software. The rate of the exponential function is defined as the rate of fluorescence loading to the terminal.

For analysis comparing the fluorescence intensity of ribbons and Ca2+ influx elicited by 100-ms stimuli (Fig. 11A & B), 10 images of both TAMRA-RBP and Cal520LA-RBP were collected in sequential, followed by imaged Cal520LA-RBP only during 200-ms depolarization, and the sequence ended with 10 images of both TAMRA-RBP and Cal520LA-RBP in sequential to confirm the ribbon locations. Two color imaging were obtained sequentially with a frame interval of 407 ms, and Cal520LA-RBP only during 100ms-stimuli was 205 ms. Images before and after were then averaged to obtain the ribbon location. Synaptic ribbon fluorescence was quantified as the ratio of TAMRA fluorescence to the fluorescence of the nearby RBC cytoplasm, measured approximately eight to nine pixels away (Fribbon/Fnear by), measuring the pixel with the highest intensity. The change in Cal520LA-RBP fluorescence, used as a proxy for active zone Ca2+ influx, was estimated as ΔF/Frest, where Frest is the fluorescence intensity before depolarization at −65 mV, and ΔF is the difference when depolarized to 0 mV. Ca²⁺ indicator intensity was calculated as the average fluorescence of nine pixels centered on the pixel with the greatest fluorescence increase.

Experimental design and statistical methods

We did not use any statistical methods to determine the sample size prior to experiments. Mean value variances were reported as ± the standard error of the mean (SEM). The statistical significance of differences in average amplitudes of Ca2+ current, capacitance, synaptic ribbon size and number, and Ca2+ transients were assessed using unpaired, two-tailed t-tests with unequal variance using R Studio (Version 2023.09.0+463) and Igor Pro software.

Amplitude analysis for Figure 8. Amplitude data were processed and analyzed using R Studio software (Version 2023.09.0+463). The data were divided into two groups, that of individual cells and of multiple cells, and analyzed as follows. Multiple individual cells from each condition (proximal 10 mM EGTA Cal520LA-RBP and distal 10 mM EGTA Cal520LA-RBP) were analyzed to compare the variability between the Ca2+ amplitude measurements of the ribbons from each cell. Having verified the normality of the data to ensure the validity of the statistical test, we used one-way ANOVA and Tukey’s honest significant difference test to assess the specific statistical differences between ribbons. To compare the variability between the amplitude measurements of different cells, we compared readings for various cells. If there were multiple measurements for a single ribbon in a specific cell, those readings were averaged. In this case, the normality of the data was again verified and Welch’s ANOVA was employed to assess the between-group differences as it does not assume homogeneity of variance, with the Games-Howell post-hoc test being used to understand the specific differences between cells.

Computational modeling of Ca2+ dynamics

Spatio-temporal Ca2+ dynamics is simulated in a volume shown in Fig. 9, which is a 3D box of dimensions (1.28 x 1.28 x 1.1) µm3. Assuming an approximately spherical bouton of diameter 5µm and an average of 36 ribbons per terminal, one obtains synaptic volume per ribbon of 1.81 µm3, which matches the volume of our box domain. Ribbon serves as an obstacle to Ca2+ and buffer diffusion, and has an ellipsoidal shape with semi-axis of 190nm along the Y- and Z-axes (for a total height of 380nm), and a semi-axis of 70 nm along the X-axis. It is attached to the plasma membrane by a thin ridge (arciform density) of dimensions 60x30x30 nm. Ca2+ ions enter through 4 channels (or clusters of channels) at the base of the ribbon indicated by black disks in Fig. 6, forming a square with a side length of 80 nm; this highly simplified channel arrangement is sufficient to capture the level of detail in spatial [Ca2+] distribution that we seek to resolve. Each of the four clusters admits a quarter of the total Ca2+ current. Ca2+ current values and current pulse durations are listed in Fig. 7. Although the actual distribution of Cav channels is more heterogeneous, channels close to the ribbon play a greater role in neurotransmitter release 2.

The simulation includes a single dominant Ca2+ buffer with molecules possessing a single Ca2+ ion binding site. We assume that only immobile buffer is present in a patch-clamped cell, since any mobile buffer would diffuse into recording pipette. Simulations with mobile buffer in an un-patched cell are also provided (Supplementary Fig. 4B&C). Buffer parameter values were set to values reported by 7, namely dissociation constant of 2 µM, and total concentration of 1.44 mM (total buffering capacity at rest of 720). We also examine the impact of the assumption on buffer concentration by repeating the simulation with a lower buffering capacity of 100 (total buffer concentration of 200 µM). The Ca2+ ion diffusivity is set to 0.22 µm2/ms 89.

To minimize the number of undetermined parameters, we assumed the same parameters for the Ca2+ clearance on all surfaces, with one high-capacity but lower affinity process simulating Na+/Ca2+ exchangers, and one lower capacity but higher affinity process simulating Ca2+ extrusion by ATPase PMCA and SERCA pumps. We used the Hill coefficient of m=2 for the Ca2+ sensitivity of the SERCA pumps on all surfaces except the bottom surface 90, whereas Hill coefficient of m=1 corresponds to PMCA pumps on the bottom surface 91. Thus, the flux of [Ca2+] across the surface (ions extruded per unit area per unit time) is described by where C is the Ca2+ concentration, DC is the Ca2+ diffusivity, and n ⋅ ▽C is the gradient of [Ca2+] normal to the boundary. The constant leak term ΦLeak ensures that the flux is zero at the resting value [Ca2+]rest = C0 = 0.1 µM. The Ca2+ clearance rate and affinity parameters for the pumps and exchangers are listed in Supplementary Table 3. We note however that the density of Na+-Ca2+ exchangers and PMCA/SERCA pumps are rough estimates and have a significant effect on results only for longer stimulation durations of over 100-200 ms. Reaction-diffusion equations for [Ca2+] and buffer are solved numerically using Calcium Calculator software version 6.10.7 92,93.

Serial block-face scanning electron microscopy (SBF-SEM)

The eyecups of adult (16 ∼20-month-old) zebrafish were dissected in bicarbonate-buffered Ames’ medium and, after removing the lens, the eyecups were fixed in 4% glutaraldehyde in 0.1 M sodium-cacodylate buffer at RT for one hour, followed by overnight fixation at 4°C. Thereafter, samples were treated to generate blocks for serial block face scanning electron microscopy (SBF-SEM) according to the published protocol 94, in this case using Zeiss 3-View SBF-SEM. Multiple 35 µm montages were acquired at a 5 nm X-Y resolution and a 50nm Z resolution across the entire retina from the outer (photoreceptor) layer to the inner (ganglion) cell layer. Images were inspected visually and structures of interest were traced using the TrakEM plug-in in ImageJ software (NIH/Fiji), as described below.

Analysis of serial block-face scanning electron microscopy images. SBF-SEM image stacks of the zebrafish retina were imported, aligned, visualized, and analyzed using the TrakEM plugin (version 1.5h) in ImageJ software (NIH). RBC1 were identified based on their characteristic morphology and 4-7 μm terminal size, as previously described 48. RBCs, synaptic ribbons, pre-synaptic membranes, and the area of the ribbon touching the plasma membrane were traced, painted using the TrakEM brush features, and rendered in 3D for structural visualization. The area of the synaptic ribbon touching the RBC membrane was quantified using the measuring tools in ImageJ and images were prepared using Adobe Photoshop software (version 25.9). The distance between each ribbon from three different RBCs and its nearest five ribbons (Fig. 9B and Video 1) was measured as follows: Ribbons on the same plane had their linear distance measured, whereas the distance between ribbons that were close to each other but in different layers was calculated using the Pythagorean Theorem.

eLife Assessment

The study introduces new tools for measuring the intracellular calcium concentration close to transmitter release sites, which may be relevant for synaptic vesicle fusion and replenishment. This approach yields important new information about the spatial and temporal profile of calcium concentrations near the site of entry at the plasma membrane. This experimental work is complemented by a coherent, open-source, computational model that successfully describes changes in calcium domains. Key gaps in the data presented mean that the evidence for the main conclusions is currently incomplete.

Reviewer #1 (Public review):

This paper describes technically-impressive measurements of calcium signals near synaptic ribbons in goldfish bipolar cells. The data presented provides high spatial and temporal resolution information about calcium concentrations along the ribbon at various distances from the site of entry at the plasma membrane. This is important information. Important gaps in the data presented mean that the evidence for the main conclusions is currently inadequate.

Strengths

(1) The technical aspects of the measurements are impressive. The authors use calcium indicators bound to the ribbon and high-speed line scans to resolve changes with a spatial resolution of ~250 nm and a temporal resolution of less than 10 ms. These spatial and temporal scales are much closer to those relevant for vesicle release than previous measurements.

(2) The use of calcium indicators with very different affinities and different intracellular calcium buffers helps provide confirmation of key results.

Weaknesses

(1) Multiple key points of the paper lack statistical tests or summary data from populations of cells. For example, the text states that the proximal and distal calcium kinetics in Figure 2A differ. This is not clear from the inset to Figure 2A - where the traces look like scaled versions of each other. Values for time to half-maximal peak fluorescence are given for one example cell but no statistics or summary are provided. Figure 8 shows examples from one cell with no summary data. This issue comes up in other places as well.

(2) Figure 5 is confusing. The figure caption describes red, green, and blue traces, but the figure itself has only two traces in each panel and none are red, green, or blue. It's not possible currently to evaluate this figure.

(3) The rise time measurements in Figure 2 are very different for low and high-affinity indicators, but no explanation is given for this difference. Similarly, the measurements of peak calcium concentration in Figure 4 are very different from the two indicators. That might suggest that the high-affinity indicator is strongly saturated, which raises concerns about whether that is impacting the kinetic measurements.

Reviewer #2 (Public review):

Summary:

The study introduces new tools for measuring intracellular Ca2+ concentration gradients around retinal rod bipolar cell (rbc) synaptic ribbons. This is done by comparing the Ca2+ profiles measured with mobile Ca2+ indicator dyes versus ribbon-tethered (immobile) Ca2+ indicator dyes. The Ca2+ imaging results provide a straightforward demonstration of Ca2+ gradients around the ribbon and validate their experimental strategy. This experimental work is complemented by a coherent, open-source, computational model that successfully describes changes in Ca2+ domains as a function of Ca2+ buffering. In addition, the authors try to demonstrate that there is heterogeneity among synaptic ribbons within an individual rbc terminal.

Strengths:

The study introduces a new set of tools for estimating Ca2+ concentration gradients at ribbon AZs, and the experimental results are accompanied by an open-source, computational model that nicely describes Ca2+ buffering at the rbc synaptic ribbon. In addition, the dissociated retinal preparation remains a valuable approach for studying ribbon synapses. Lastly, excellent EM.

Weaknesses:

Heterogeneity in the spatiotemporal dynamics of Ca2+ influx was not convincingly related to ribbon size, nor was the functional relevance of Ca2+ dynamics to rod bipolars demonstrated (e.g., exocytosis to different postsynaptic targets). In addition, the study would benefit from the inclusion of the Ca2+ currents that were recorded in parallel with the Ca2+ imaging.

Reviewer #3 (Public review):

Summary:

In this study, the authors have developed a new Ca indicator conjugated to the peptide, which likely recognizes synaptic ribbons, and have measured microdomain Ca near synaptic ribbons at retinal bipolar cells. This interesting approach allows one to measure Ca close to transmitter release sites, which may be relevant for synaptic vesicle fusion and replenishment. Though microdomain Ca at the active zone of ribbon synapses has been measured by Hudspeth and Moser, the new study uses the peptide recognizing synaptic ribbons, potentially measuring the Ca concentration relatively proximal to the release sites.

Strengths:

The study is in principle technically well done, and the peptide approach is technically interesting, which allows one to image Ca near the particular protein complexes. The approach is potentially applicable to other types of imaging.

Weaknesses:

Peptides may not be entirely specific, and the genetic approach tagging particular active zone proteins with fluorescent Ca indicator proteins may well be more specific. I also feel that "Nano-physiology" is overselling, because the measured Ca is most likely the local average surrounding synaptic ribbons. With this approach, nobody knows about the real release site Ca or the Ca relevant for synaptic vesicle replenishment. It is rather "microdomain physiology" which measures the local Ca near synaptic ribbons, relatively large structures responsible for fusion, replenishment, and recycling of synaptic vesicles.

Author response:

The following is the authors’ response to the original reviews

Reviewer #1 (Public review):

This paper describes technically-impressive measurements of calcium signals near synaptic ribbons in goldfish bipolar cells. The data presented provides high spatial and temporal resolution information about calcium concentrations along the ribbon at various distances from the site of entry at the plasma membrane. This is important information. Important gaps in the data presented mean that the evidence for the main conclusions is currently inadequate.

Thank you very much for this positive evaluation of our work. We would like to respectfully point out to the Reviewer that our current study was conducted using zebrafish as a model and not goldfish. We have revised the paper to eliminate any gaps in the data presentation.

Strengths

(1) The technical aspects of the measurements are impressive. The authors use calcium indicators bound to the ribbon and high-speed line scans to resolve changes with a spatial resolution of ~250 nm and a temporal resolution of less than 10 ms. These spatial and temporal scales are much closer to those relevant for vesicle release than previous measurements.

(2) The use of calcium indicators with very different affinities and different intracellular calcium buffers helps provide confirmation of key results.

Thank you very much for this positive evaluation of our work.

Weaknesses

(1) Multiple key points of the paper lack statistical tests or summary data from populations of cells. For example, the text states that the proximal and distal calcium kinetics in Figure 2A differ. This is not clear from the inset to Figure 2A - where the traces look like scaled versions of each other. Values for time to half-maximal peak fluorescence are given for one example cell but no statistics or summary are provided. Figure 8 shows examples from one cell with no summary data. This issue comes up in other places as well.

Thank you for this feedback. We have addressed this in our revised manuscript where possible. We now include the results of paired-t-tests to compare the amplitudes of proximal vs. distal calcium signals shown in Fig. 2A & C, Fig. 3C & D, Fig. 4 C & D, Fig. 5A-D, and Fig. 8E&F. Because proximal and distal calcium signals were obtained from the same ribbons within 500-nm distances, as the Reviewer pointed out, “the traces look like scaled versions of each other”. For experiments where we make comparisons across cells or different calcium indicators, as shown in Fig.3 E&F, Fig.5E, and Fig. 8B&C, we now include the results of an unpaired t-test. We have now included the t-test statistics information in the respective figure legends in the revised version.

Regarding the Reviewer’s concern that “values for time to half-maximal peak fluorescence are given for one example cell, but no statistics or summary are provided,” we estimated the fluorescence rise times by only fitting the average traces to compare the overall qualitative behavior of the corresponding calcium indicator fluorescence. We did attempt to analyze the uncertainty for the rise-time estimates, but the simultaneous fitting of the rise- and decay-behavior of time traces is notoriously sensitive to noise, and therefore, a much higher signal-to-noise ratio would be required to provide reliable uncertainty estimation for the corresponding rise-time and decay-time characteristics. This is now explicitly explained in the corresponding Methods subsection.

In Figure 8, we now show example fluorescence traces from one cell at the bottom of the A and D panels, and the summary data is described in B-C and E-F, with statistics provided in the figure legends.

(2) Figure 5 is confusing. The figure caption describes red, green, and blue traces, but the figure itself has only two traces in each panel and none are red, green, or blue. It's not possible currently to evaluate this figure.

Thank you for pointing out this oversight. The figure shows the proximal and distal calcium signals, not the cytoplasmic ones. The figure caption was adjusted to correctly reflect what is shown in the figure.

(3) The rise time measurements in Figure 2 are very different for low and high-affinity indicators, but no explanation is given for this difference. Similarly, the measurements of peak calcium concentration in Figure 4 are very different from the two indicators. That might suggest that the high-affinity indicator is strongly saturated, which raises concerns about whether that is impacting the kinetic measurements.

We agree with the Reviewer and had mentioned in the text that we do believe that the high-affinity version of the dye is at least partially saturated. This will be especially a problem for strong depolarizations and signals near the membrane. We slightly changed the corresponding description of results on page 6 to acknowledge this point: “However, it should be noted that Cal520HA will be at least partially saturated at the Ca2+ levels expected in Ca2+ microdomains relevant for vesicle exocytosis, affecting both the amplitude and the kinetics of the fluorescence signal”.

Recommendations:

(1) It would be good to describe the location of calcium channels relative to the ribbon in the introduction.

We have provided this information in the discussion (please see p. 19: “The faster, smaller, and more spatially confined Ca2+ signals that are insensitive to the application of high concentrations of exogenous Ca2+ buffers, referred to here as ribbon proximal Ca2+ signals, could be due to Ca2+ influx through Cav channel clusters beneath the synaptic ribbon”). We have now provided this information in the last paragraph of the introduction as well.

(2) The introduction is quite technical and would benefit from a more complete description of the findings of the paper (e.g. expanding the last sentence to a full paragraph).

We have updated the last paragraph of the introduction as per the reviewer’s advice.

(3) It is not clear that the capacitance measurements in Figure 1 are needed (I did not see them used anywhere else in the paper).

We have removed the capacitance measurements from the figure.

(4) Please add legends in the figures themselves defining different line colors and weights so that a reader does not need to search for them in the figure caption.

We agree that such figure improvements facilitate reading. We have added legends in the figures themselves, where appropriate.

(5) The insets with the expanded traces in many cases are too small - e.g. Figure 1F.

We have enlarged the insets in applicable figures as much as possible to facilitate visualization. These changes can be seen in Figures 1, 2, 3, 4, 5, and 8, as well as Supplementary Figure 3.

(6) Page 5, statistics for amplitude of calcium changes. Is p < 0.001 really correct here? The SEMs indicate an overlap of the two distributions of mean amplitudes - and later data for which you give p = 0.001 has much less overlap.

Since the two data sets in question come from paired recordings, with a high Pearson correlation coefficient of 0.93, the p-values are in fact, correct despite this significant overlap. We conducted paired-t-tests to compare proximal vs. distal calcium signals obtained from a single calcium indicator shown in Fig. 2A & C, Fig. 3C & D, Fig.4 C & D, Fig.5A-D, and Fig. 8E&F. For experiments where we make comparisons across cells or across different calcium indicators, as shown in Fig.3 E&F, Fig.5E, and Fig. 8B&C, we performed an unpaired t-test. In response to the Reviewer’s comment, we now provide details on t-statistics in the respective figure legends in the revised version.

(7) The text on page 6 describing Figure 3 appears to repeat several technical aspects of the measurements that have already been described in Figure 1. I would reduce that overlap as it is confusing for a reader.

Since Fig.1 describes calcium measurements with free calcium indicator, whereas Fig.3 describes bound calcium indicator, we would prefer to keep the information for the sake of completeness, despite some small amount of repetition.

(8) Figure 4A needs to be described in more detail.

We have provided the vesicle pool details in the Supplementary Fig. 1.

(9) The text in Figure 7 is too small.

We have redone Fig. 7 and Supplemental Fig. 4 to ensure that the tick labels and other text are sufficiently large.

(10) Are the units (nM) in Figure 8 correct?

Thank you for pointing that out. The units were supposed to be µM and have been corrected in the figure.

Reviewer #2 (Public review):

Summary:

The study introduces new tools for measuring intracellular Ca2+ concentration gradients around retinal rod bipolar cell (rbc) synaptic ribbons. This is done by comparing the Ca2+ profiles measured with mobile Ca2+ indicator dyes versus ribbon-tethered (immobile) Ca2+ indicator dyes. The Ca2+ imaging results provide a straightforward demonstration of Ca2+ gradients around the ribbon and validate their experimental strategy. This experimental work is complemented by a coherent, open-source, computational model that successfully describes changes in Ca2+ domains as a function of Ca2+ buffering. In addition, the authors try to demonstrate that there is heterogeneity among synaptic ribbons within an individual rbc terminal.

Strengths:

The study introduces a new set of tools for estimating Ca2+ concentration gradients at ribbon AZs, and the experimental results are accompanied by an open-source, computational model that nicely describes Ca2+ buffering at the rbc synaptic ribbon. In addition, the dissociated retinal preparation remains a valuable approach for studying ribbon synapses. Lastly, excellent EM.

Thank you very much for this appreciation of our work.

Weaknesses:

Heterogeneity in the spatiotemporal dynamics of Ca2+ influx was not convincingly related to ribbon size, nor was the functional relevance of Ca2+ dynamics to rod bipolars demonstrated (e.g., exocytosis to different postsynaptic targets). In addition, the study would benefit from the inclusion of the Ca2+ currents that were recorded in parallel with the Ca2+ imaging.

Thank you for this critique. We agree that our data do not establish the relationship between ribbon size and Ca2+ signal. By analogy to the hair cell literature, we believe that it is a reasonable hypothesis, but more studies will be necessary to definitively determine whether the signal relates to ribbon size or synaptic signaling. This will be addressed in future experiments.

We have included the calcium current recorded in parallel with calcium imaging in Fig.1, when we show a single example. We now do the same for individual examples shown in Fig. 8 A and D, bottom. The calcium imaging data shown in Figs. 2-5 and Supp. Fig. 3 is the average trace, thus we have provided the averages of the peak calcium current and statistics. Since in Figure 8D-F some ribbons only have one reading, we have not conducted statistical analysis in this case.

Recommendations:

The major conclusion of the work is that within bipolar cells, heterogeneity exists between Ca2+ microdomains formed at synaptic ribbons, which is supported by the results; however, what causes this is not clear. Most of the comments below are suggestions that hopefully help the authors strengthen the association of Ca2+ domain heterogeneity with features of ribbon AZs or at least offer additional options for the authors to communicate their work.

(1) In the current study, anatomical segregation of SRs by size does not appear to exist across the ZF rod bipolar terminal, nor has this been reported for mouse rod bipolars. In the absence of this, the current study lacks the fortuitous attributes, and thus reasoning, utilized in the hair cell (HC) studies (those cited in the current MS). Namely, the HC studies utilized the following anatomical features to compare EM, IF, and physio results: a) identified differences in ribbon synapses along a tonotopic gradient (basal to apical cochlea), b) compared ribbons on different sides of an inner HC (pillar vs. modiolar), or c) examined age-dependent changes in HC ribbons.

Thank you for this comment. We agree that we do not show any interesting systematic relationships between ribbon size and cell position or other large-scale morphological features. We added text on page 19 to stress this (“However, in comparing our findings with studies of ribbon size heterogeneity in hair cell…”). However, to our knowledge, diversity in ribbon size has never been reported in bipolar cells.

(2) In the absence of intrinsic topographical segregation in ribbon size within rod bipolars, then a) the imaging data attained from dissoc cells needs to be internally as sound as possible, and b) the parameters used to define ribbon dimensions in light (LM) and electron microscopy should be as communicative/interchangeable as possible.

Thank you for this comment. Our confocal images show a moderate correlation between ribbon size measured as fluorescence of ribeye binding peptide vs. calcium hot spots. Similarly, SBF-SEM images demonstrate that the ribbon active zone length vs width show a moderate correlation. We have summarized these findings in Figure 11. Thus, as the Reviewer pointed out, our confocal and SBF-SEM findings support each other.

(3) It is not entirely clear how the authors distinguish rod bipolars (a subset of On-bipolars) from all other ON-bipolars? The two different preparations: dissoc or intact retina, present distinct challenges. In the example presented in Supplementary Figure 2B, the PKCalpha stained bipolar has an axon that is approx. 25 um long, but the expected length should be approx. 50um based on ZF retinal anatomy and recent study on rbc1/2 (Hellevik et al BioRxiv 2023). One could argue rather that the enzymatic treatment or mechanical shear forces caused the axon to shrink. If that is the line of reasoning, then present a low mag field of view with an assortment of dissoc bipolars stained for PKCalpha, zoom in, and describe cell morphologies and their assignment as PKCa + or -. Then you can summarize how axon terminal size, axon length, and PKC staining are or aren't correlated. Based on the results, one might have to perform IF on each dissoc cell that was assayed under LM (Ca2+ imaging) and ephys to verify it's a rod bipolar. In the case of the EM, the authors refer to the terminals analyzed as rbcs because they have larger terminals and less branching than the cbs. Since these are really nice EM images, data-rich, with better resolution than I have ever seen for retinal SBF-EM, do due diligence by tracing the terminals of neighboring bcs (ignoring details within terminals just outline terminals) and make a visual presentation that illustrates that those you selected as rbs have larger terminals than cbs (this can also give of sense of the density distribution of terminal types). Is there a published ephysio on the ZF rbcs which has been correlated with morphology? The Hellevik et al BioRxiv 2023 study shows light responses but not necessary rbcs distinguished from other On-bcs.

We have quantified the number of rod bipolar cells obtained from our isolation procedure using two approaches: 1. To fix the isolated bipolar cells and perform immunofluorescence with PKC alpha. 2. To isolate bipolar cells from Tg(vsx1: memCerulean)q19 transgenic zebrafish, labeling rod bipolar cell type 1 (RBC1) that we recently obtained from Dr. Yoshimatsu (Hellevik et al., 2024). Of note, the circuitry of RBC1 has been shown to be similar to the mammalian rod bipolar cell pathway (Hellevik et al., 2024). Below, we list our findings:

The average terminal size of fixed bipolar cells labeled with PKC alpha was 5.9 ± 0.2 mm, whereas the freshly isolated living bipolar cells used for our physiology experiments had an average terminal size of 6.3 ± 0.2 mm, and the rod bipolar cells from the Tg(vsx1: memCerulean)q19 line had an average terminal size of 6.9 ± 0.2 mm. We also measured terminal size for fixed bipolar cells, unlabeled with PKC alpha: 3.3 ± 0.2 mm, and unlabeled cells from Tg(vsx1: memCerulean)q19 cells: 4.0± 0.2 mm.

In addition, we also pay attention to the soma shape and dendrites, as the primary dendrite of the RBC is thick and short. Connaughton and Nelson have done a thorough analysis of morphological classification. But no measurements were given. https://onlinelibrary.wiley.com/doi/10.1002/cne.20261. Since the axon length is not retained during the isolation procedure, we do not use it as an identification marker for rod bipolar cells in our experiments.

We re-imaged vsx1 with the DIC channel to compare the terminal sizes of fluorescently labeled RBC1 terminals with those of other BPCs in the DIC channel. Below are the images that can give a sense of the density distribution of terminal types and measurements.

Tracing all neighboring terminals in SBF-SEM is laborious and beyond the scope of this manuscript, but we will do full reconstructions in a future publication.

(4) How to strengthen the description of heterogeneity within the dissoc measurements? There are two places in the LM data where heterogeneity may be relevant. The first point here is that Ribbon size (TAMRA- Ribeye binding peptide) and active zone size (Cal520HA/LA-RBP) measurements depend on labelling the ribbon/Ribeye; thus, Ribbon size and AZ size should be correlated on this basis alone. I would expect Pearson's r value to show a stronger association (r > 0.7) than what is reported in Figure 11B/C (r: 0.52 or 0.32). I would interpret a moderate to weak correlation (r < 0.5 to 0.3) as an indication that ribbons are heterogeneous (variability in Ca influx per unit ribbon size). Now to the second point, in Figure 8 and Supplementary Figure 5 there is time-signal amplitude heterogeneity. >>> My curiosity is whether signal amplitude is heterogeneous in space (ribbon size, my speculation) and in time (complex, but compare ribeye bound and free Ca2+ indicator)? It seems like the data in Figure 8 and 11 should cross over and possibly offer the authors more to say.

We appreciate the Reviewer’s insightful observation and added a sentence at the very end of the Results section reflecting the Reviewer’s argument (“we note that a large correlation between the inferred ribbon size and active zone size…”)

The Reviewer’s second point about the connection between heterogeneity of signal amplitude in space and in time is an interesting one as well and could be grounds for an additional investigation in the future.

(5) As the authors know, a very powerful tool for exploring Ca microdomain dynamics is to exploit the Voltage dependence of Cavs (as exemplified in the numerous HC studies that are cited). An I-V protocol would provide a valuable means to illustrate different rates of saturating the LA and HA Ca indicators. More generally, the Ca currents and associated patch clamp parameters (Gm, leak...) can tell us much about the health of the cell and provide an added metric to assess normal variability between cells. A few places in the MS currents are mentioned yet this data is missing (Figure S5 , last line: Amplitude variability between two cells with similar Ca currents.).

Thank you for the valuable suggestion. We will include I-V protocol across several ribbons in future experiments. We have included the calcium currents for all the calcium transient traces. We have also included the statistics to compare those currents across conditions.

Technical comments

(6) Since the Ribeye-Ca2+ indicator covers the entire ribbon, it will contribute to a signal gradient. The proximal signal is assumed to be closest to the base of the ribbon where presumably the Cav channels are located, and the distal signal will originate from the top (apex) of the ribbon some 200 nm from the base of the ribbon. Have you tried to measure "ribbon lengths and widths" with the HA and LA Ca indicators? My guess would be that the LA will show a gradient, and give you a better indication of the base of the ribbon; whereas the HA signal will have dimensions similar to the TAMRA-peptide.

Due to the point spread function limitation in the light microscopy, we obtained all ribbon measurements from the SBF-SEM images only.

As a surrogate for size in the light microscopy, we used ribbon fluorescence, which we expect should scale with the number of ribeye molecules in the ribbon (Figure 11B)

(7) Normalize proximal and distal LM data to highlight kinetic differences (Fig 2-5, 8), and when describing temporal heterogeneity please use a better description that includes time, such as time-to-pk, and decay1, decay 2....

In the current manuscript, we only focus on the amplitude as it provides the information about the number of calcium channels. We used the rise time measurements to compare the time to reach the peak amplitude at the proximal vs. distal locations, demonstrating that proximal calcium signals reach the peak faster since the calcium channels are located beneath the ribbon.

We tried to perform fittings to the individual traces. Since they are too noisy to pick out true kinetic differences between ribbons, we would need to average several traces from each ribbon. We plan to apply our high-resolution approach established in this paper to a longer stimulus and perform the fittings as per the Reviewer’s advice for a future paper.

We now describe on pages 6-7 the two decay components for data in Figs. 2 and 3.

(8) Why not measure ribbon length in EM as done in confocal and then compare lengths from LM and EM. In Figure S8, you have made a nice presentation of AZ Area from EM. Make similar plots for EM ribbon length (and width?), and compare the distributions to Figure 11 LM data. Maybe use other statistical descriptions like Coeff of Var or look for different populations by using multi-distribution fits. If the differences in length or area (EM data) can be segregated into short and long distances, then a similar feature might arise from the LM data. If no such morphological segregation exists, then the heterogeneity in Ca microdomains may arise from variable Cav channel density or gating, Ca buffer, etc.

Due to the point spread function limitation in light microscopy, the size of the ribbon dimensions in light microscopy cannot be reliably measured. As a surrogate, we used total fluorescence of the ribbon, which should correlate with the number of ribeye molecules in the ribbon. To obtain ribbon dimensions, we used measurements from the SBF-SEM images only. We summarized the distribution of ribbon width and length in Figures 11C and 11D. The distribution of the active zone size is summarized in Supplementary Figure 8. Pearson’s correlation coefficients are positive, but a weak correlation, suggesting multiple mechanisms likely to contribute to heterogeneity in the local calcium signals as the Reviewer pointed out.

(9) Again, the quality of the EM data is great, and sufficient to make the assignment of SVs to different pools, as you have done in Fig S1. My only complaint is that the Ultrafast pool as indicated in the schematic of S1A seems to have a misassignment with respect to the green SV that is 15 nm from the PM. In the original Mennerick and Matthews 1996 study, the UF pool emptied in ~1msec. The morphological correlate for the UF has been assumed to be SVs touching the plasma membrane. 15 nm away is about 14 nm too far to be in the UF.

Thank you for pointing that out. We have updated the vesicles labeling in Supplementary Figure 1 and Main Figure 4.

Reviewer #3 (Public review):

Summary:

In this study, the authors have developed a new Ca indicator conjugated to the peptide, which likely recognizes synaptic ribbons, and have measured microdomain Ca near synaptic ribbons at retinal bipolar cells. This interesting approach allows one to measure Ca close to transmitter release sites, which may be relevant for synaptic vesicle fusion and replenishment. Though microdomain Ca at the active zone of ribbon synapses has been measured by Hudspeth and Moser, the new study uses the peptide recognizing synaptic ribbons, potentially measuring the Ca concentration relatively proximal to the release sites.

Thank you very much for this positive evaluation of our work.

Strengths:

The study is in principle technically well done, and the peptide approach is technically interesting, which allows one to image Ca near the particular protein complexes. The approach is potentially applicable to other types of imaging.

Thank you very much for this appreciation.

Weaknesses:

Peptides may not be entirely specific, and the genetic approach tagging particular active zone proteins with fluorescent Ca indicator proteins may well be more specific. I also feel that "Nano-physiology" is overselling, because the measured Ca is most likely the local average surrounding synaptic ribbons. With this approach, nobody knows about the real release site Ca or the Ca relevant for synaptic vesicle replenishment. It is rather "microdomain physiology" which measures the local Ca near synaptic ribbons, relatively large structures responsible for fusion, replenishment, and recycling of synaptic vesicles.

The peptide approach has been used fairly extensively in the ribbon synapse field and the evidence that it efficiently labels the ribbon is well established, however, we do acknowledge that the peptide is in equilibrium with a cytoplasmic pool. Thus, some of the signal arises from this cytoplasmic pool. The alternative of a genetically encoded Ca-indicator concatenated to a ribbon protein would not have this problem, but would be more limited in flexibility in changing calcium indicators. We believe both approaches have their merits, each with separate advantages and disadvantages.

As for the nano vs. micro argument, we certainly do not want to suggest that we are measuring the same nano-domains, on the spatial scale of 10s of nanometers, that drive neurotransmitter release, but we do believe we are in the sub-micrometer -- 100s of nm -- range. We chose the term based on the usage by other authors to describe similar measurements (Neef et al., 2018; https://doi.org/10.1038/s41467-017-02612-y), but we see the reviewer’s point.

Recommendations:

I have no recommendation for additional experiments. However, the statement of "nanophysiology" is too much, and the authors should tone done the ms recognizing some caveats.

As we mention above, we chose the term based on the usage by other authors to describe similar measurements, and we do believe that we achieve resolution of a few hundred nanometers, and therefore would prefer to keep the current title of the manuscript. For example, Figure 5E shows that, with ribeye-bound low-affinity calcium indicator, the proximal calcium signals were preserved in the presence of BAPTA, rising and decaying abruptly, as expected for a nanodomain Ca2+ elevation. Thus, we believe that this measurement in particular describes a nanodomain-scale signal. However, we acknowledge that we are not currently able to resolve the spatial distribution of Ca2+ signals with a spatial resolution of 10s of nanometers.

Version 2

Abstract

Abstract

Rapid and high local calcium (Ca2+) signals are essential for triggering neurotransmitter release from presynaptic terminals. In specialized bipolar ribbon synapses of the retina, these local Ca2+ signals control multiple processes, including the priming, docking, and translocation of vesicles on the ribbon before exocytosis, endocytosis, and the replenishment of release-ready vesicles to the fusion sites for sustained neurotransmission. However, our knowledge about Ca2+ signals along the axis of the ribbon active zone is limited. Here, we used fast confocal quantitative dual-color ratiometric line-scan imaging of a fluorescently labeled ribbon binding peptide and Ca2+ indicators to monitor the spatial and temporal aspects of Ca2+ transients of individual ribbon active zones in zebrafish retinal rod bipolar cells (RBCs). We observed that a Ca2+ transient elicited a much greater fluorescence amplitude when the Ca2+ indicator was conjugated to a ribeye-binding peptide than when using a soluble Ca2+ indicator, and the estimated Ca2+ levels at the ribbon active zone exceeded 26 μM in response to a 10-millisecond stimulus, as measured by a ribbon-bound low-affinity Ca2+ indicator. Our quantitative modeling of Ca2+ diffusion and buffering is consistent with this estimate and provides a detailed view of the spatiotemporal [Ca2+] dynamics near the ribbon. Importantly, our data demonstrates that the local Ca2+ levels may vary between ribbons of different RBCs and within the same cells. The variation in local Ca2+ signals is correlated to ribbon size, which in turn correlates with active zone extent, as serial electron microscopy provides new information about the heterogeneity in ribbon size, shape, and area of the ribbon in contact with the plasma membrane.

Introduction

Sensory synapses in the retina rely on the proper function of a specialized organelle, the synaptic ribbon 1–5. Retinal bipolar cells serve as the major conduit for transmitting visual information across the vertebrate retina. Rod bipolar cells (RBCs) can release brief bursts of neurotransmitters to signal a change in contrast or sustain the continuous release of neurotransmitters in a graded manner to provide an analog read-out of luminance 6. To maintain this ability, the RBCs must exert dynamic control over neurotransmitter release rate and facilitate efficient recruitment of release-ready vesicles to fusion sites near the synaptic ribbon. It is established that the elevation of presynaptic Ca2+ in RBCs regulates both dynamic changes in the release rate and accelerates the rate of vesicle replacement 6–14. However, the spatiotemporal properties of the Ca2+ signals that control neurotransmitter release and the molecular entities that regulate the interplay between Ca2+ signals and synaptic vesicle dynamics to sustain kinetically distinct neurotransmitter release components remain poorly understood. Here, we begin to address this lack of knowledge by measuring local Ca2+ signals at positions along the synaptic ribbon at different distances from the active zone in retinal RBCs.

Resolving local Ca2+ signals is technically challenging as it requires information about the spatiotemporal properties of Ca2+ signals specific to ribbon sites. Previous studies of Ca2+ dynamics in goldfish and mammalian RBCs focused primarily on the terminal as a whole and used quantitative methods to examine bulk Ca2+ levels, which are significantly lower and slower than those occurring at the active zone 15–18. Our previous studies estimated that the Ca2+ signals at a single ribbon active zone in zebrafish RBCs likely exceed sub-micromolar concentration level within 3 milliseconds after the opening of voltage-gated Ca2+ (Cav) channels located beneath the synaptic ribbon 19. However, these studies were based on estimates using soluble Ca2+ indicators to estimate the Ca2+ signals at the plasma membrane, which are free to diffuse and thus cause the spread of the signal. Measurements of Ca2+ signals along the axis of the ribbon have not been attempted, even though these signals likely control replenishment and priming of vesicles for slower phases of neurotransmitter release. To determine Ca2+ signals along the axis of zebrafish RBC ribbons, we establish a nanophysiology ratiometric approach that measures the spatial and temporal properties of Ca2+ transients by targeting the Ca2+ indicator to the ribbon via conjugation to a ribbon binding peptide (RBP) 20. Resolving Ca2+ signals in the immediate vicinity of the active zone is not currently possible due to the finite spatiotemporal resolution of optical imaging. Thus, we employ a quantitative model using Ca2+ diffusion and buffering to resolve the RBC Ca2+ signals associated with the active zone and along the ribbon and provide a detailed description of the spatiotemporal Ca2+ dynamics across zebrafish RBC ribbons. Our nanophysiology approach for measuring local Ca2+ transients at a single ribbon with high spatiotemporal resolution provides the first evidence of heterogeneity of Ca2+ signals at zebrafish RBC ribbon synapses. Heterogeneity in local Ca²⁺ signals persisted in some ribbons even after the application of high concentrations of exogenous Ca²⁺ buffers, suggesting that the variability in the faster, smaller, and more spatially confined Ca²⁺ microdomains originates from Ca²⁺ influx through synaptic Ca²⁺ channel clusters at the base of the synaptic ribbon. Using serial block-face scanning electron microscopy (SBF-SEM), we found substantial variability in synaptic ribbon size, shape, and particularly the area of the ribbon adjacent to the plasma membrane in the zebrafish bipolar cell terminal. Thus, the observed heterogeneity of the Ca2+ microdomain is likely due to variability in the number of Ca2+ channels near each ribbon. Since local Ca2+ signals control kinetically distinct neurotransmitter release, heterogeneity in local Ca2+ signal may alter the rate of vesicle release and allow them to function independently, adding a new mechanism for increasing dynamic range of RBC.

Results

Ca2+ concentrations at single ribbon locations measured using high– and low-affinity diffusible indicators

Synaptic vesicles in RBCs are distributed among at least four distinct pools based on their fusion kinetics, which are assumed to reflect the average proximity of vesicles to Cav channels and the state of vesicle preparedness for Ca2+-triggered fusion 1,6,8,11–13,21–25 (see Supplementary Fig. 1). To visualize and measure Ca2+ signals near the RBC synaptic ribbon controlling such kinetically distinct components of neurotransmitter release, we established a quantitative nanophysiology approach shown in Fig.1 (see also Materials and Methods). In this approach, both the ribbon location and the spatiotemporal Ca2+ signal profile were simultaneously measured by dialyzing the zebrafish RBC terminals with both TAMRA (tetramethyl rhodamine)-labeled RIBEYE-binding peptide (RBP) 26 (to label synaptic ribbons; Fig. 1Aa) and a high-affinity Ca2+ indicator Cal-520® (Cal520HA, effective KD 795 nM; see Materials and Methods) using a whole-cell patch pipette placed directly at the cell terminal. A rapid x-t line scan was taken perpendicular to the plasma membrane across a ribbon, extending from the extracellular space to the cytoplasmic region beyond the ribbon (Fig. 1Ab). The spatial resolution was limited by the point spread function (PSF) of the microscope to approximately 270 nm (Supplementary Fig.2A). In our previous studies, line scans have been applied primarily to measure the temporal properties of Ca2+ signals 19. However, x-t raster plots obtained at a ribbon active zone labeled with RBP allow us to characterize the spatial localization of Ca2+ transients relative to the synaptic ribbon and plasma membrane (Fig.1C-D) in addition to characterizing its temporal aspects (Fig.1E-F). We previously used a similar approach for localizing and tracking single synaptic vesicles before and during fusion at a single ribbon 27, and to measure the kinetics of clearance of fused synaptic vesicle membrane in zebrafish RBC 28. We found that depolarization-evoked Ca2+ influx caused a rapid increase in Ca2+ signals at ribbon locations (Fig.1C, cyan, white horizontal arrowhead) and increased more slowly and less dramatically in the cytoplasm (Fig.1C, cyan, white vertical arrow) in zebrafish RBC. The Sigmoid-Gaussian function fitting of the x-t scans horizontal profile scans (see Materials and Methods) show that the local Ca2+ signals increased rapidly during stimuli (Fig.1D, cyan line) and then decreased immediately after the end of depolarization (Fig.1D, black line) approaching the spatial profile corresponding to the resting Ca2+ levels (Fig.1D, gray line). As expected, the centroid position of Ca2+ signals during depolarization (Fig.1D, cyan x0) is closer to the plasma membrane (Fig.1D, magenta x1/2) than the centroid position of RBP (Fig.1D magenta, x0), since Cav channels are located at the membrane 19,26,29–42.

Nanophysiology approach unveils spatiotemporal properties of local Ca2+ signaling in retinal RBC terminals.

A. Left panel. Single projection from a series of confocal optical sections through a zebrafish RBC synaptic terminal. A synaptic terminal was voltage-clamped using a whole-cell pipette with an internal solution containing TAMRA-RBP (magenta) to label synaptic ribbons (Ai, magenta). RBP fluorescence was concentrated at ribbons and also filled the entire terminal, allowing visualization of the terminal border. Experiments were carried out on ribbons that could be distinguished from adjacent ribbons (white asterisks). Scale, 5 μm. Right panel, Close-up view of a single synaptic ribbon. The outside of the cell is to the right, and x-t scan lines (dotted lines) were positioned perpendicular to the plasma membrane, extending from the intracellular side of the ribbon to the extracellular space. A rapid x-t line scan was taken at a ribbon location perpendicular to the plasma membrane across a ribbon (Aii) with sequential dual laser scanning performed at rates of 1.51 milliseconds per line per channel (3.02 milliseconds per line for both channels). The resulting x-t raster plots were used to measure the fluorescence intensity profiles of RBP (magenta) and the Ca2+ transient (cyan; Cal520HA) shown in panels C and E. Scale, 1.6 μm.

B. Voltage-clamp recording of a RBC terminal. Terminals were held at –65 mV and stepped to 0 mV (t0) for 10 ms (red) to evoke a brief Ca2+ current (black). A typical experiment began with a voltage command (VH= –65 mV), and a TTL pulse generated by the Patch Master software (t1) triggers image acquisition (t2). t0 is the time of depolarization.

C. Illustration of the approach used to obtain the spatial location of Ca2+ signals with respect to the ribbon. Example of an x-t raster plot that is oriented to illustrate the x-axis intensity profiles of RBP (magenta) and Ca2+ signal (cyan) fluorescence during a brief depolarization. Sequential dual laser scanning was performed at 1.51 milliseconds per line for one channel (3.02 milliseconds per line for both channels).

D. Fluorescence intensity profiles along the x-axis for RBP (magenta) and Cal520HA before stimuli (gray line), three pixels along the time-axis. RBP (magenta) and Ca2+ signals during (light cyan) were fit with a Sigmoid-Gaussian function 1. The centroid (x-axis position) of the RBP (magenta) and Ca2+ signals during (cyan) were taken as the peak of the Gaussian fit (x0). The parameter x1/2 (dotted magenta line) from the Sigmoid fit to the RBP fluorescence (magenta trace) was used to estimate the location of the plasma membrane.

E. The x-t raster plot shown was from the same recording as in panel C but re-oriented to demonstrate the t-axis intensity profiles of RBP (magenta) and Ca2+ transient (cyan).

F. Spatially averaged Cal520HA fluorescence as a function of time at ribbon proximal (light cyan), distal (dark cyan), and cytoplasmic (gray) locations from the single ribbon shown in D, upper panel.

To characterize the temporal Ca2+ profile at different distances from the plasma membrane, we analyzed the x-t line scans along the time-axis at three distinct distances (Fig. 1E-F) based on the spatial profile of the ribbon described in Fig.1C-D (see also Materials and Methods). We will refer to the corresponding three measurements as ribbon-proximal, ribbon-distal, and cytoplasmic (see Materials and Methods). We found that at the onset of the stimulus (Fig. 1F, black arrow), the fluorescence at the location near the ribbon proximal to the membrane (ribbon-proximal) rose more rapidly and to a higher level (Fig. 1F, light cyan trace) than that at the ribbon-distal and cytoplasmic locations (Fig. 1F, dark cyan and gray traces, respectively). To quantify the kinetics of ribbon-proximal and distal Ca²⁺ signals, we averaged multiple x-t scans acquired with Cal520HA under the same imaging conditions (Fig. 2A). The rise time of the Ca2+ signals produced in response to brief 10 milliseconds stimuli, defined here as the time for the averaged Ca2+ transient measured with Cal520HA to reach half-maximal peak fluorescence amplitude from the resting level (Fig. 2A), is faster for the proximal (∼2.8 milliseconds) than distal (∼3.4 millisecond) signals (Fig. 2A). The corresponding maximum values of trial-averaged ΔF/Frest (changes in the Cal520HA fluorescence during depolarization normalized to background level before depolarization) was significantly higher at ribbon-proximal locations (nearest the Cav channels) than those at ribbon-distal locations, with mean ± SEM of 1.9 ± 0.3 and 1.5 ± 0.25, respectively (Fig. 2A; p< 0.001, N=24). These findings suggest that the spatial resolution of our Ca2+ imaging using Cal520HA is sufficient to resolve differences in the smaller Ca2+ signals at ribbon proximal and distal locations. However, it should be noted that Cal520HA will be at least partially saturated at the Ca2+ levels expected in Ca2+ microdomains relevant for vesicle exocytosis 43, affecting both the amplitude and kinetics of the fluorescence signal. Therefore, we repeated the x-t line scan analysis with a lower-affinity soluble Ca2+ indicator Cal520LA (KD 90 μM; Fig. 2B) allowing us to better define the typical Ca2+ signals controlling distinct neurotransmitter release components corresponding to locations proximal and distal relative to the synaptic ribbon (Fig. 2B). We found that the ribbon-proximal signals detected with Cal520LA (Fig. 2B, light cyan) showed a sharper rise and decay at the onset and termination of the stimulus (Fig. 2B, black arrow), when compared to ribbon-distal signals (Fig. 2B, dark cyan), as one would expect for nanodomain Ca2+ elevations 44–46. The relationship between rise times for the proximal (4.6 millisecond) and distal (7.1 millisecond) signals were similar to those we found with Cal520HA, with both proximal Ca2+ signals being faster than the distal signals. In twenty-one similar experiments, the peak ΔF/Frest at the membrane after 10-ms depolarization was significantly larger for proximal than distal signals (ΔF/Frest: 3.1 ± 0.4 and 1.9 ± 0.2 respectively P=0.001, N= 21). As expected, the ratio of proximal to distal signals measured with Cal520LA (1.6) was significantly higher than that measured with Cal520HA (1.3). The decay phase of all fluorescence transients was well fit by a sum of two exponential functions, as described in Methods; for the Cal520HA recording (Fig. 2A) the two decay time constants and their relative magnitudes were 11ms (60%) and 203ms (40%) for the ribbon-proximal recording, vs. 16ms (61%) and 208ms (39%) for the distal site. For the low affinity Cal520LA dye (Fig. 2B), the fluorescence decay components were faster due to faster Ca2+ unbinding from the dye, at 9.3ms (69%) and 175ms (31%) for the ribbon-proximal location, vs. 9ms (41%) and 135ms (59%) for the distal location.

Kinetics of Ca2+ transients in response to brief stimuli recorded with freely diffusible indicators.

A. Spatially averaged Cal520HA fluorescence as a function of time at ribbon proximal (light cyan) and distal (dark cyan) locations from a single ribbon, as shown in Fig.1 (n=24 ribbons from 7 different RBCs). The corresponding maximum values of trial-averaged ΔF/Frest was significantly higher at ribbon-proximal locations than those at ribbon-distal locations, with mean ± SEM of 1.9 ± 0.3 and 1.5 ± 0.25, respectively (paired t-test, p< 0.001, N=24).

A inset. The temporal profile of events between 20-60 ms is shown in an expanded view for better visualization. Scale bars: vertical, 0.5 (βF/Frest); horizontal,10 ms.

B. Spatially averaged Cal520LA fluorescence as a function of time at ribbon proximal (light cyan) and distal (dark cyan) locations. Data points show the average intensity (± SEM) in each horizontal row of 5 pixels for three 10 ms depolarizations at distinct ribbon locations (see Materials and Methods and Fig.1). Fluorescence intensity is normalized with respect to the baseline fluorescence before stimulation, and averaged over all pixels (i.e., over space and time). The arrow indicates the onset of the 10-ms depolarizing stimulus. (n=21 ribbons from 4 different RBCs). The peak ΔF/Frest at the membrane after 10-ms depolarization was significantly larger for proximal than distal signals (paired t-test, ΔF/Frest: 3.1 ± 0.4 and 1.9 ± 0.2, respectively, P=0.001, N= 21). The currents were not significantly different between the Cal520HA and Cal520LA conditions (unpaired t-test, Cal520HA average current = 45.1 ± 4.5 pA; Cal520LA average current = 51.4 ± 4.4 pA; p = 0.33).

B inset. Temporal profile of events between 20-60 ms was expanded for better visualization. Scale bars: vertical, 0.5 (βF/Frest); horizontal, 10 ms.

Ca2+ concentrations at single ribbon locations measured using ribbon-bound indicators

Although Ca²⁺-sensitive fluorescent chemical dyes have been used previously 7,20,29,36,42,43,45–57, the visualization of signals within smaller domains using freely diffusible Ca²⁺ reporters is limited by the resolution of light microscopy and the spread of the indicator by diffusion 58. Diffusible Ca2+ indicators report space-averaged Ca2+ concentrations, and their intracellular diffusion inherently broadens the spatial resolution of Ca2+ nanodomains. To partially overcome this problem, we targeted the Ca2+ indicators to the ribbon by fusing them to the RBP, as described previously 20. RBP-conjugated Ca2+ indicators Cal520HA-RBP or Cal520LA-RBP were introduced to the RBC terminal together with fluorescently labeled RBP via whole-cell voltage clamp by placing the patch pipette directly at the terminal while imaging the terminal using laser scanning confocal microscopy. For two-color imaging, ribbons were labeled with TAMRA-RBP that did not interfere with Cal520-RBP fluorescence, and both channels were scanned sequentially to prevent possible bleed-through. Under these conditions, we used the spots detected by TAMRA-RBP to define the locus of the synaptic ribbon (Fig 3A&B, magenta). We also found punctate regions with both Cal520HA-RBP (Fig 3A, cyan) and Cal520LA-RBP (Fig 3B, cyan) at the same location, on a dimmer fluorescent background of the synaptic terminal, which correspond to the Ca2+ indicator-peptide complexes that are bound and not bound to the ribbon, respectively 20. The overall changes in ΔF/Frest were averaged over several trials for proximal and distal Ca2+ signals measured with Cal520HA-RBP (Fig. 3C, light cyan vs. dark cyan) and Cal520LA-RBP (Fig. 3D, light cyan vs. dark cyan). When compared to their distal location counterparts, the Ca2+ signals proximal to the membrane showed a sharper rise and decay at both the onset and termination of stimuli (Figs. 3C and 3D, light vs. dark cyan trace), as expected when comparing nano– and microdomain Ca2+ profiles. Our data shows that the amplitude differences between ribbon-proximal and ribbon-distal Ca2+ signals were well resolvable using the ribbon-bound Cal520HA-RBP indicator (Fig. 3C, light vs. dark cyan traces: ΔF/Frest = 3 ± 0.4 vs. 1.9 ± 0.3, respectively, p=0.001) and Cal520LA-RBP indicator (Fig. 3D, light vs. dark cyan traces: ΔF/Frest = 5.5 ± 0.9 vs. 3.3 ± 0.8, respectively, p=0.003). The amplitudes of ribbon-proximal Ca2+ signals were higher when measured with Cal520LA-RBP than with Cal520LA-free (Fig. 3E, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray): ΔF/Frest = 5.5 ± 0.9, n=30 vs. 3 ± 0.4, n=19, p=0.04) but this was not the case for distal Ca2+ signals (Fig. 3F, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray): ΔF/Frest = 3.3 ± 0.8, n=30 vs. 1.9 ± 0.2, n=21, p=0.15). Notably, the Ca2+ signals at distal sites measured with Cal520LA-RBP reached their peak amplitude earlier (Fig. 3F inset, dark cyan) than those measured with Cal520LA-free (Fig. 3F inset, light gray). The two decay time constants and their relative magnitudes for the Cal520HA-RBP fluorescence (Fig. 3C) were 11ms (52.5%) and 182ms (47.5%) for the ribbon-proximal site, vs. 24ms (47%) and 192ms (53%) for the site distal to the ribbon. For the low affinity ribbon-bound indicator dye (Fig. 3D), the fluorescence decay components were faster, at 5.7ms (83%) and 129ms (17%) for the proximal site, vs. 7.5ms (76%) and 158ms (24%) for the distal location. Together, these findings strongly suggest that conjugating the Cal520LA indicator to RBP provides a more accurate, promising approach for measuring the distinct local Ca2+ signals at ribbon-proximal vs. ribbon-distal locations.

Temporal properties of Ca2+ transients recorded with free and ribeye bound Ca2+ indicators.

A-B. Confocal images of the isolated RBCs that were whole-cell voltage-clamped using an internal solution containing the TAMRA-RBP (magenta) and either (A) Cal520HA-RBP (cyan) or (B) Cal520LA-RBP (cyan). Note prominent spots in both TAMRA-RBP and ribeye-bound Ca2+ indicators (A) Cal520HA-RBP (cyan) or (B) Cal520LA-RBP, showing the location of the ribbon. Scale bars, 2 μm.

C-D. Spatially averaged fluorescence of (C) Cal520HA-RBP (n=19) or (D) Cal520LA-RBP (n=30) as a function of time at ribbon proximal (light cyan), and distal (dark cyan) locations. Data points show the average intensity (± SEM) in each horizontal row of five pixels for 10 ms depolarizations at distinct ribbon locations. Fluorescence intensity at the onset of the 10 ms depolarizing stimulus (arrow) was normalized to the baseline fluorescence before stimulation and averaged over all pixels (i.e., over space and time). The amplitude differences between ribbon-proximal and ribbon-distal Ca2+ signals were well-resolvable using the ribbon-bound Cal520HA-RBP indicator (Fig. 3C, light vs. dark cyan traces, paired t-test: ΔF/Frest = 3.0 ± 0.4 vs. 1.9 ± 0.3, respectively, p=0.001) and Cal520LA-RBP indicator (Fig. 3D, light vs. dark cyan traces, paired t-test: ΔF/Frest = 5.5 ± 0.9 vs. 3.3 ± 0.8, respectively, p=0.003). The current amplitudes were not significantly different between Cal520HA-RBP and Cal520LA-RBP readings (mean current amplitudes: Cal520HA-RBP = 49.8 ± 2.2 pA, Cal520LA-RBP = 49.9 ± 2.9 pA; p = 0.99).

C-D inset. Temporal profile of events between 10-50 ms were expanded for better visualization. Scale bars: vertical, 1 (βF/Frest, C inset) or 2 (βF/Frest, D inset); horizontal, 10 ms.

E-F. Average fluorescence intensity of (E) proximal and (F) distal Ca2+ signals obtained with Cal520LA-RBP (light cyan and dark cyan, respectively) and Cal520L-free (grey). The amplitudes of ribbon-proximal Ca2+ signals were higher when measured with Cal520LA-RBP than with Cal520LA-free (Fig. 3E, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray), unpaired t-test: ΔF/Frest = 5.5 ± 0.9, n=30 vs. 3.0 ± 0.4, n=19, p=0.04) but this was not the case for distal Ca2+ signals (Fig. 3F, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray), unpaired t-test: ΔF/Frest = 3.3 ± 0.8, n=30 vs. 1.9 ± 0.2, n=21, p=0.15).

E-F inset. Events between 10-50 ms were expanded for better visualization. Scale bars: vertical, 2 (βF/Frest, E inset) and 1 (βF/Frest, F inset); horizontal, 10 ms.

This method also enables ratiometric measurements with RBP-conjugated Ca2+ indicator by normalizing its fluorescence to that of RBP-peptides conjugated to Ca2+-insensitive fluorophores (TAMRA-RBP) to provide an estimate of Ca2+ concentration in RBC synaptic ribbons. We obtained the effective K1/2 (Keff) by measuring the Cal520HA-RBP/ TAMRA-RBP fluorescence ratio in buffered Ca2+ solutions and using the Grynkiewicz equation53. For Cal520HA-RBP, we found the Keff to be ∼795 nM, which is higher than the value of 320 nM reported by the manufacturer for Cal520HA. The differences between the in-cell measurements and the manufacturer’s values are likely to arise from differences in the cellular buffering capabilities, changes in dye properties due to interactions with the molecules inside the cell 36,59–61, and possible differences in the binding properties of the peptide-conjugated Ca2+ indicators when bound to synaptic ribbons. Since the in-cell approach most closely matched the experimental conditions, we used this value (Keff) for all further calculations.

We first measured the ribbon-proximal and ribbon-distal Ca2+ concentrations using Cal520HA-RBP with 0.2 mM EGTA in the patch pipette as it allowed to resolve the gradient between the two signals (Fig. 3C). Under these conditions, we found a maximum ribbon-proximal Ca2+ concentration produced in response to a brief 10-ms pulse of 3.7 µM, and a maximum ribbon-distal Ca2+ concentration of 0.7 µM. These apparent Ca2+ concentration amplitudes are well below levels required to trigger exocytosis of the ultrafast releasable pool (UFRP) and readily releasable pool (RRP) 43 and, as discussed earlier, likely represent the lower bounds of the Ca2+ concentration at a single ribbon location due to local saturation of the high-affinity indicator and/or due to nanodomains that are smaller than the resolution attainable using light microscopy. It should also be noted that unlike freely diffusing Ca2+ indicators, RBP-conjugated indicators that slowly unbind from the ribbon are not readily replaced by free Ca2+ indicators, rendering them even more prone to saturation. Nevertheless, our data demonstrate that ribbon-bound indicators may better report local Ca2+ concentrations due to their localization, albeit still subject to the limitations of light microscopy.

To test the contribution of local saturation and to better estimate ribbon Ca2+ concentrations, we repeated our measurements using the low-affinity Ca2+ indicator Cal520LA conjugated to the RBP peptide. Because it was difficult to perform in-cell measurements to determine the Keff for Cal520LA-RBP given the large amounts of Ca2+ required to calibrate the Cal 520LA indicator (see Methods and Materials), we used the K1/2 value provided by the manufacturer for free Cal520LA (90 µM). However, we expect that in-cell measurements of Keff, are likely to be different due to the cellular buffering properties reported previously for Keff, measurements of Oregon Green BAPTA-5N in inner hair cells 36.

Bipolar cells release neurotransmitters primarily from ribbon active zones, although some release also occurs at non-ribbon sites (referred to as NR, Fig. 4A) 62–64. To reveal the Ca2+ signaling at and away from the ribbon, we performed whole-cell patch clamping and x-t line scans at ribbon (Fig. 4Ba, R) and non-ribbon (Fig. 4Bb, NR) sites perpendicular to the plasma membrane using TAMRA-RBP and Cal520LA-RBP. A brief (10-ms) depolarizing voltage-clamp pulse evoked rapid high Ca2+ signals (Fig. 4Ba, cyan raster plot) at ribbon locations (Fig. 4Ba, magenta raster plot) but not at non-ribbon locations (Fig. 4Bb, cyan raster plot). The amplitude of Ca2+ signals elicited by brief depolarization and detected with Cal520LA-RBP at the ribbon-proximal site (Fig. 4C, light cyan traces) were significantly higher than those at ribbon-distal (Fig. 4C, dark cyan traces) and non-ribbon (Fig. 4C, blue traces) sites. The Ca2+ concentration gradient along the ribbon is summarized in Fig. 4D. We found that the average Ca2+ concentration at the proximal side of the ribbon (26.4 ± 3.1 µM, n=26) was significantly different from that at the ribbon-distal (15.6 ± 1.5 µM, n=26) and non-ribbon (10.4 ± 0.4 µM, n=15) sites, and that Ca2+ concentrations at ribbon-distal sites is higher than those at non-ribbon sites (Fig. 4D). These measurements display large heterogeneity across distinct ribbons and distinct cells, with the coefficient of variation of about 60% for [Ca2+] measurements at locations proximal to the ribbon, compared to a 10% CV for multiple depolarizations for the same ribbon. The Ca2+ signals at distal sites also reached their peak amplitude earlier (Fig. 4C inset, dark cyan arrowhead) than those at the non-ribbon sites (Fig. 4C inset, blue arrowhead). These findings are consistent with non-ribbon vesicle release being governed by cytoplasmic residual Ca2+ or Ca2+ influx via clustered Cav channels at ribbon sites rather than non-ribbon active zones with clusters of Cav channels and are also in agreement with previous reports regarding the temporal delay and sensitivity of non-ribbon exocytosis to EGTA 63,65.

Ca2+ signals at synaptic ribbon at different distances from the plasma membrane.

A. Ultrastructure of a zebrafish RBC with kinetically distinct vesicle pools, as described in Supplementary Fig. 1 UFRP (green vesicles), and RRP (yellow vesicles) are primarily released via ribbon sites (R) at the cytomatrix of the active zone (arrowhead). Recycling pool (RP, orange vesicles) in the cytoplasm (C), likely to be released via non-ribbon (NR) sites. Scale bar: 500 nm.

B. Representative x-t plots show the fluorescence intensity of Cal520LA-RBP (cyan) and TAMRA-RBP (magenta) as a function of distance (vertical axis) and time (horizontal axis) at (Ba) ribbon sites and (Bb) non-ribbon (NR) locations. The darker region at the upper edge of each plot is the extracellular space and the arrowheads show the timing of depolarization. Scale bars: vertical, 1.6 μm and horizontal, 75 ms.

C. Spatially averaged Cal520LA-RBP as a function of time at ribbon proximal (light cyan), distal (dark cyan), and non-ribbon locations (blue) (n=26 (proximal ribbons), 26 (distal ribbons), and 15 (non-ribbon) from 5∼8 different RBCs, respectively).

C inset. Temporal profile of events between 0-100 ms were expanded for better visualization. Scale bars: vertical, 0.2 (βF/Frest); horizontal, 20 ms.

D. Ca2+ measurements along the ribbon axis using the nanophysiology approach demonstrate the proximal Ca2+ signals can go as high as 26.4 ± 3.1 µM (light cyan, N=26) and distal as 15.6 ± 1.5 µM (dark cyan, N=26), and non-ribbon 10.4 ± 0.4 µM (NR, blue, N=15), respectively, in response to 10 ms stimuli. Error bars show standard errors. All conditions were significantly different from each other as assessed by paired t-test when comparing proximal vs. Distal and unpaired t-test when comparing non-ribbon to proximal or distal (proximal vs. distal p = 0.004; proximal vs. non-ribbon p < 0.001; distal vs. non-ribbon p = 0.002). The currents were not significantly different between conditions (Mean current: 0.2 mM EGTA Cal520L-RBP proximal and distal = 51.8 ± 3.2 pA, 0.2 mM EGTA Cal520L-RBP non-ribbon = 47.3 ± 3.0 pA; p = 0.35).

Sensitivity of microdomain Ca2+ to exogenous buffers

Previous work has shown that the rate of vesicle replacement in RBCs is accelerated by elevated Ca2+ levels at sites along the ribbon, and that while millimolar levels of EGTA have little effect on the fast exocytosis component near Cav channels, they selectively block sustained exocytosis, likely by preventing Ca2+ from reaching the distal locations on the ribbon 11,66–69. These findings raise the possibility that Ca2+ has two sites of action, one near the Cav channels that trigger vesicle release and one further away that replenishes the supply of releasable vesicles. Burrone et al. (2002) proposed that endogenous Ca2+ buffers regulate the size of the RRP by limiting the spatial spread of Ca2+ ions and could suppress vesicle release at the periphery of the active zone in bipolar cells 7. However, the Ca2+ gradient at different locations along the ribbon controlling RRP release has never been examined. Thus, we used our higher resolution approach (Fig.1) to measure the Ca2+ gradient along the synaptic ribbon under buffering conditions that differentially modulate vesicle release and resupply. To determine the spatiotemporal properties of Ca2+ signals under these conditions, we performed rapid x-t line scans at a single ribbon location in the presence of ribbon-bound Cal520LA-RBP and under varying concentrations of exogenous buffer in the patch pipette solution, including (EGTA at 0.2 (Fig. 5A), 2 (Fig. 5B), and 10 mM (Fig. 5C) or BAPTA at 2 mM (Fig. 5D)). We found that the ratio between proximal and distal Ca2+ signal amplitudes was similar in 0.2 and 2 mM EGTA (proximal-to-distal ratio ∼1.7, Supplementary Table 1) but was enhanced with 10 mM EGTA (proximal-to-distal ratio ∼1.9, Supplementary Table 1) and further enhanced with 2 mM BAPTA (proximal-to-distal ratio ∼4.5, Supplementary Table 1). These findings once again emphasize the reliable measurement of resolving ribbon-proximal vs. ribbon distal Ca2+ signals using our nano-physiology approach. Our experimental findings of the increase in the proximal-to-distal Ca2+ concentration ratio with increasing EGTA concentration are consistent with our simulation results (Fig. 7), although the corresponding ratios are greater in the simulation than in the experiment due to the large size of the microscope’s point-spread function.

Effect of exogenous Ca2+ buffers on spatiotemporal properties of Ca2+ microdomains in RBC terminal recorded with low-affinity ribbon-bound dye.

A-D. Average temporal fluorescence intensity (normalized to βF/Frest) of proximal (light cyan) and distal (dark cyan) Ca2+ signals with Cal520LA-RBP as a function of time at distinct ribbon locations with pipette solution containing (A) 0.2 mM EGTA, (B) 2 mM EGTA, (C) 10 mM EGTA, or (D) 2 mM BAPTA. Proximal measurements were significantly higher than distal measurements in all conditions as assessed using paired t-tests (0.2 mM EGTA: p = 0.0027, n = 29; 2 mM EGTA: p = 0.034, n = 21; 10 mM EGTA: p < 0.001, n = 42; 2 mM BAPTA: p = 0.0073, n = 21). The currents between conditions were not significantly different from each other (mean current amplitudes in 0.2 mM EGTA: 50.6 ± 3.0 pA, 2 mM EGTA: 49.7 ± 3.1 pA, 10 mM EGTA: 47.3 ± 3.1 pA, 2 mM BAPTA: 56.1 ± 2.6 pA; 0.2 mM EGTA vs 2 mM EGTA: p = 0.84, 0.2 mM EGTA vs 10 mM EGTA: p = 0.46, 0.2 mM EGTA vs 2 mM BAPTA: p = 0.19). Inset. The temporal profiles of events between 0-50 ms were expanded for better visualization. Scale bars: vertical, 2 (βF/Frest; panels A and B) and 1 (βF/Frest; panels C and D); horizontal, 10 ms.

E. Average temporal fluorescence intensity (normalized to βF/Frest) of proximal Ca2+ signals measured with Cal520LA-RBP (cyan) and Cal520LA-free (gray) as a function of time with pipette solution containing 2 mM BAPTA. The corresponding maximum values of trial-averaged ΔF/Frest was significantly higher with 2mM Cal520L bound BAPTA than 2 mM Cal520L free BAPTA, with mean ± SEM of 3.6 ± 0.9 and 1.3 ± 0.2, respectively (paired t-test, p = 0.015; Cal520L-free: n = 20; Cal520L-bound: n = 21). The currents between conditions were not significantly different from each other (mean current amplitudes in 2 mM BAPTA Cal520L-RBP: 56.1 ± 2.6 pA; 2 mM BAPTA Cal520L-Free: 53.3 ± 2.2 pA; p = 0.42).

Geometry of the computational model of intra-terminal Ca2+ dynamics.

Simulation domain is a box with dimensions (1.28 x 1.28 x 1.1) µm3, approximating the fraction of synaptic terminal volume per single ribbon. Ca2+ ions enter near the base of the ellipsoidal ribbon at four locations marked by black disks representing Ca2+ channels or their clusters. The semi-transparent gray coordinate plane Y=0 corresponds to the section used for the pseudo-color 2D [Ca2+] plots in Figure 7. Ca2+ is extruded on all surfaces of this domain, simulating combined clearance by pumps and exchangers on the plasmalemmal as well as internal endoplasmic reticulum and mitochondrial membranes.

Simulation of the effect of an endogenous immobile buffer of different concentrations on [Ca2+] dynamics in response to a 10ms pulse.

(A1-C2) In each sub-plot, the left-hand panels show the [Ca2+] time course in response to a 10 ms constant current pulse (total current of 1 pA) at five select locations marked by colored circles in the right panel. The right-hand panels show a pseudo-color plot of [Ca2+] in a 2D section of the 3D simulation volume in Fig. 6, at a fixed point in time corresponding to the end of the current pulse. Concentration values for each level curve are indicated in the color bar. Endogenous buffer is immobile, with a concentration of 200 µM in panels A1-C1 (resting buffering capacity 100 µM), vs. 1.44 mM in panels A2-C2 (resting buffering capacity 720). Exogenous buffer concentrations were 0.2 mM EGTA (panels A1, A2), 10 mM EGTA (panels B1, B2), 2 mM BAPTA (panels C1, C2).

Though BAPTA significantly abolished the spread of Ca2+ signals, it is impressive that a substantial amount of ribbon-proximal Ca2+ concentration measured with ribbon-bound indicators was still present with 2 mM BAPTA. One reason for this observation could be that ribbon-bound Ca2+ indicators are likely to measure Ca2+ signals very close to Cav channels without diffusing away, which makes it impossible for BAPTA to intercept Ca2+ ions. If so, similar experiments conducted with free Ca2+ indicators should report a significantly lower proximal Ca2+ signal since the measurement by a diffusible dye would effectively spread the signal over a larger volume. Indeed, this is what we found. As shown in Fig. 5E, the apparent proximal Ca2+ signal in response to a 10-ms brief pulse measured with Cal520LA-free in the presence of 2 mM BAPTA was 0.9 ± 0.2 (n=9), 4-fold lower than proximal Ca2+ signals measured with Cal520LA-RBP (3.6 ± 1, n=20). The latter value is closer to the true measure of the Ca2+ concentration in the vicinity of the ribbon base. These findings emphasize that the reliable measurement of ribbon-proximal Ca2+ signals in the vicinity of Cav channels greatly benefits from the increased spatial resolution of our nano-physiology approach. We also measured the spatiotemporal properties of local Ca2+ signals using Cal520HA-RBP as we have done above for Cal520LA-RBP (Supplementary Fig. 3). We found that the ratio between proximal and distal Ca2+ signals was similar in 0.2 and 2 mM EGTA (proximal-to-distal ratio ∼1.5, Supplementary Table 2) but was enhanced with 10 mM EGTA (proximal-to-distal ratio ∼2.6, Supplementary Table 2) and further enhanced with 2 mM BAPTA (proximal-to-distal ratio ∼2.8, Supplementary Table 2). Cal520 acts as a buffer that binds Ca2+ and carries it away by diffusion. Thus, the higher affinity indicator Cal520HA will bind and shuttle away more Ca2+ than Cal520LA. If the Cal520HA indicator is saturated near the Cav channels, it may be underreporting the fast Ca2+ transients that occur in those locations. However, Cal520LA and Cal520HA show similar increase in the ribbon proximal-to-distal ratios with increasing concentrations of exogenous buffer.

Computational models of Ca2+ signals along the axis of the RBC synaptic ribbon

Since we could identify the position of the ribbon and we used RBP-fused Ca2+ indicator-RBP, we were able to measure and distinguish ribbon-proximal vs. ribbon-distal Ca2+ signals that drive the release of UFRP and RRP. However, several factors, including finite spatiotemporal resolution of optical imaging, spatial diffusion, dye saturation, and dye binding kinetics, may limit our ability to achieve optimal resolution. These limitations, however, can be overcome by the use of quantitative models. Thus, we developed a model based on our data with published information on Ca2+ diffusion and buffering to estimate more accurately the [Ca2+]I gradient along the ribbon at distinct distances from the plasma membrane.

Combining whole-terminal Ca2+ current measurements and our estimate for the number of synaptic ribbons per terminal allowed us to infer the magnitude of Ca2+ current per single ribbon, which we used in our model to determine the spatiotemporal [Ca2+] dynamics near the ribbon. Fig. 7 shows the results of the simulation of [Ca2+] at various distances from the ribbon during and after a depolarizing pulse of 10ms duration in the presence of different concentrations of exogenous and endogenous buffers, replicating the conditions used in our experiments. The geometry of the simulation domain box is shown in Fig. 6; it represents the fraction of total synaptic terminal volume per single ribbon (see Methods, Supplementary Table 3 and Supplementary Fig.4A.). Fig. 7 shows the time-dependence of [Ca2+] during and after the depolarization pulse at 5 specific locations both near and distant from the ribbon (left-hand panels in each subplot). These five spatial locations are marked by circles of different colors in the right-hand panel of each subplot, which shows [Ca2+] at the end of the Ca2+ current pulse in pseudo-color in the entire planar cross-section of the simulation domain cutting through the middle of the ribbon, as shown in Fig. 6. We assumed a highly simplified Cav channel arrangement into four clusters forming a square with a side length of 80nm. The closest of the five spatial locations (red curves and circles in Fig. 7) is X=20 nm away from the base of the ribbon center-line, Z=10 nm above the ribbon, about 45 nm away from the closest Cav channel cluster (see Fig. 6). We assumed that this location was within the Ca2+ microdomain that triggered the release of the UFRP. Vesicles were not included in the simulation since the total exclusion volume attributed to them represented only a small fraction of the total inter-terminal volume and, therefore, did not significantly impact [Ca2+] at the qualitative resolution level we are interested in. A couple of simulations with vesicles included were performed to confirm this statement, requiring much finer spatial resolution and longer computational time. We note also that [Ca2+] for short pulse durations considered here is relatively insensitive to our assumptions on the membrane Ca2+ extrusion mechanisms, listed in Supplementary Table 3.

Fig. 7A1-B2 reveals that [Ca2+] could rise above 50 µM within the microdomain near the base of the ribbon in the presence of 0.2 mM up to 10 mM EGTA. Given the size of the microscope point-spread function, this is in qualitative agreement with the 26 μM estimate of ribbon-proximal [Ca2+] that we recorded using the ribeye-bound low-affinity indicator. In addition, this concentration level is reached very soon after the channel opening event due to the rapid formation of the Ca2+ microdomain 70,71, and therefore, this estimate is expected to hold for shorter pulses as well. However, [Ca2+] decayed rapidly with distance, with the rate of decay significantly increasing when EGTA concentration was increased to 10mM (Fig. 7B1, B2). Note that a 7-fold change in the capacity of the immobile endogenous Ca2+ buffer (cf. Fig. 7A1 vs. 7A2) had a relatively modest effect on the [Ca2+] level up to about 150 nm distance from the ribbon base. This effect of increasing endogenous buffer concentration was further reduced in 10 mM EGTA (cf. Fig. 7B1 vs. 7B2), due to strong competition of endogenous buffer with large concentrations of EGTA. This agrees with the expectation that immobile buffers primarily slow down Ca2+ signals but do not affect the rapidly forming quasi-steady-state Ca2+ microdomains in the immediate vicinity of the channel 70–72. Finally, Fig. 7C1-C2 shows that 2 mM BAPTA had a much greater effect on localizing the Ca2+ signal to the immediate vicinity of the ribbon base.

Since a native unpatched cell may well contain mobile rather than immobile buffers, it was interesting to examine the effect of increasing the concentration of mobile endogenous buffer on Ca2+ dynamics in the ribbon vicinity, as compared to the corresponding effect of immobile buffer. Supplemental Figure 4B-C shows that a 7-fold increase in mobile buffer concentration reduces the microdomain Ca2+ by a factor of 2 and greatly reduces the size of the microdomain. This contrasts with the effect of increasing the concentration of the immobile buffer, which has a much more subtle effect (cf. Fig. 7A1, A2). We note that 0.2 mM EGTA was absent in the simulation with mobile buffer shown in Supplemental Figure 4B-C; in general, the effect of modest concentrations of EGTA is expected to be negligible due to its slow Ca2+ binding speed compared to the endogenous buffer.

Variability of local Ca2+ signals across the RBC synaptic ribbons

Bipolar cells release neurotransmitters primarily from ribbon active zones 1,6,8,11–13,21–25. However, the factors that shape the synaptic ribbon microdomains at retinal ribbon synapses have not been examined. We wondered whether all of the 30∼50 synaptic ribbons at an RBC terminal release glutamate in a similar fashion or whether there is some heterogeneity in glutamate release from different ribbon active zones. In particular, we asked what underlying mechanisms could differentiate the Ca2+ signals between ribbons of the same RBC terminal and across different RBCs. We first found evidence for such variability in local Ca2+ signals at single-ribbon locations of different cells using the freely diffusible high-affinity indicator Cal520HA, observing high variability in the Ca2+ transient amplitudes, even for RBCs with similar depolarization-evoked Ca2+ current amplitudes (Supplementary Figs. 5A vs. B). However, we did not observe such variability when we obtained multiple line scans across the same ribbon (Supplementary Figs. 5A vs. B, gray traces). We wondered whether the observed variability between cells could be due to distinct subtypes of zebrafish RBCs 73, which might have different subtypes or numbers of Cav channels near the ribbon. Recent work in zebrafish retina identified two distinct RBC subtypes: RBC1 and RBC2 73. The wiring pattern of amacrine cells postsynaptic to RBC1 closely resembles the circuitry of mammalian RBCs, whereas RBC2 forms distinct pathways. Furthermore, RBC1 specifically labels for the known marker of mammalian RBCs, PKC-α 73. Due to these similarities, RBC1 in zebrafish is classified as analogous to mammalian RBCs. Moreover, RBC1s have expected morphological characteristics, in particular, the shape and size of soma and the presence of a single synaptic terminal 74. Immunolabeling of isolated RBC preparation used in these experiments showed primarily intact RBCs, which are PKC-α-specific RBC1 (Supplementary Fig. 2B). For simplicity, and because this study focuses exclusively on RBC1, we will refer to RBC1 as ‘RBC’ throughout this work. Thus, we attribute the variability in the RBC Ca2+ transients to the variability between the Ca2+ microdomains within the same cell type.

To identify mechanisms that contribute to the heterogeneity in the local Ca2+ signals we have reported in this study, we began by asking whether differences in local Ca2+ buffering could account for this heterogeneity. To examine the role of Ca2+ buffering in the variability of local Ca2+ signals across different cells and different ribbons (Figs. 8A and 8D, respectively), we compared proximal (Figs. 8B and 8E) and distal (Figs. 8C and 8F) Ca2+ signals in 10 mM EGTA by averaging the Ca2+ recordings from all the ribbons of a given cell (Supplementary Figs. 6 and 7). When the Ca2+ signals were restricted to the ribbon sites with 10 mM EGTA in the pipette solution (Supplementary Tables 1 and 2), the amplitude of the ribbon-proximal and ribbon-distal Ca2+ signals averaged over all ribbons of a given cell exhibited significant variation across different cells (Figs. 8B, 8C, and Supplementary Fig. 6). We next compared variability in Ca2+ signals in the presence of 10 mM EGTA at individual ribbons within the same cell (Fig. 8D and Supplementary Fig. 7) at proximal (Fig. 8E) and distal (Fig. 8F) locations. For the cell described in Figs. 8D-8F, the proximal Ca2+ signals were significantly different across all ribbons examined, and there were considerable differences in distal Ca2+ signals between ribbons, with the exception of ribbons numbered 2 and 5. These findings suggest that the heterogeneity in proximal and distal Ca2+ signals at distinct ribbons within the same cell may result from different underlying mechanisms, for example, heterogeneity in Cav expression or subtype and Ca2+ handling mechanisms. These findings suggest that exogenous Ca2+ buffering has a negligible effect on experimentally observed heterogeneity and variability of the proximal Ca2+ signals and that local Ca2+ signals at RBC ribbons are dominated by Ca2+ in regions close to the ribbon base where Cav channels are located.

Heterogeneity in the spatiotemporal properties of Ca2+ microdomains in RBC terminal.

A. Top panel. Cartoon of two representative RBCs (cell A and cell B), each containing differently colored ribbons. Bottom panel. Ribbon-to-ribbon variability was measured by recording local Ca2+ signals near different ribbons (yellow and purple traces) for each RBC. If a cell had multiple readings for a single ribbon, averages were obtained for comparisons as described in Supplementary Fig. 6. Bottom panel inset: sample Ca2+ currents for the cells from which the Ca2+ signal sample traces mentioned above were obtained (purple and yellow traces). Currents were similar across the different cells. Vertical scale = 80 pA, horizontal scale = 5 ms.

B-C. Variability of Ca2+ signals in different ribbons across different cells in the presence of 10 mM EGTA, recorded using Cal520LA-RBP. Proximal Ca2+ amplitude values were significantly different between cells A and B (p = 0.022), cells A and D (p = 0.023), and cells B and D (p = 0.005) but similar between A and C (n = 4 cells, 4 fish), as assessed by unpaired t-tests. (C) In distal locations, ribbon amplitude values were significantly different between cells A and B (p = 0.031), but similar across all other cell comparisons (n = 4 cells, 4 fish), as assessed by unpaired t-tests. The currents were not significantly different across the different cells, as assessed by unpaired t-tests (mean currents in cell A: 47.7 ± 2.9 pA, cell B: 43.7 ± 4.0 pA, cell C: 43.4 ± 2.0 pA, cell D: 40.9 ± 2.4 pA; cell A vs. cell B: p = 0.45, cell A vs. cell C: p = 0.22, cell A vs. cell D: p = 0.14, cell B vs. cell C: p = 0.93, cell B vs. cell D: p = 0.54, cell C vs. cell D: p = 0.52).

D. Top panel. Illustration of a RBC containing three ribbons (numbered 1-3). Bottom panel. Ca2+ signal to-ribbon variability within each RBC, as described in Supplementary Fig. 7. Bottom panel inset: sample Ca2+ currents for the cells from which the Ca2+ signal sample traces mentioned above were obtained (black, gray, and blue traces). Currents were similar across the different cells. Vertical scale = 80 pA, horizontal scale = 5 ms.

E-F. Box plot illustrating [Ca2+] across various ribbons of an individual RBC. Ribbon variability within individual cells was measured with 10 mM EGTA using Cal520LA-RBP at (E) proximal and (F) distal locations. (E) Proximal Ca2+ amplitude values were significantly different among all ribbons (paired t-test, p < 0.001) (n = 5 ribbons, 1 RBCs, 1 fish). (F) Distal Ca2+ amplitudes were significantly different among all ribbon comparisons (paired t-test, p < 0.001) except for ribbons 2 and 5 (n = 5 ribbons, 1 RBCs, 1 fish). Similar analyses were conducted in two more cells and found similar observations (data not shown). The currents were similar across all ribbons since these were readings from the same cell. Given that some ribbons only have one reading, it is not possible to conduct a paired t-test to statistically compare them; however, the average current ± standard error for the cell shown was 47.7 ± 2.9 pA.

The ultrastructure of the zebrafish RBC terminal reveals diversity in the size of the synaptic ribbons across the terminal

In hair cells, Ca2+ microdomain signaling varies with ribbon size, reflecting larger patches of Cav channels aligned with larger ribbons 51. However, the number, size, and shape of ribbon active zones per terminal have not been established in our experimental system, the synaptic terminals of zebrafish RBCs. To provide quantitative measurement of ribbon microdomains, we examined the ultrastructure of zebrafish RBC synaptic ribbons by serial block face scanning electron microscopy (SBF-SEM) and reconstructed three RBCs from serial section electron micrograph (Fig. 9A). We analyzed the bipolar cell terminals that are closest to the ganglion cell layer with a morphology similar to mammalian RBCs and zebrafish RBC1s 73. Our reconstruction of three RBCs revealed 30-41 ribbons within RBC terminals (Fig. 9A), similar to what was previously reported in goldfish giant ON-type mixed RBCs (range 45-60 75).

Distribution of measured distance between synaptic ribbons in three zebrafish RBCs.

A. Reconstruction of three RBC terminals closest to the ganglion cell layer resembling the shape and size of the mammalian RBC1. The ribbons are shown in magenta. Note that the total number of ribbons includes the “floating” ribbons detached from the plasma membrane. RBC II contained 7 floating ribbons. Scale bar, 5 µm. RBCs, rod bipolar cells.

B. Overview of the distance measurements. Top. 3D rendering of a single RBC terminal is shown in light green with ribbons in magenta. Black lines show an example of the different SBF-SEM layers that are cut to obtain individual images. Bottom. Four sample SBF-SEM layers are shown with two example ribbons (magenta) located near each other but in different layers. If two ribbons were on the same layer, their linear distance was taken, whereas if they were located in different layers, their linear distance and height difference were used to calculate their actual distance using the Pythagorean Theorem. Each SBF-SEM layer has a thickness of 50 nm. A 3D volume movie of the synaptic terminal of a zebrafish bipolar neuron synaptic ribbon distribution and measurement is provided in Video 1.

C. Histograms showing the ribbon distance for the three RBCs. x0 shows the mean of the distribution.

D. Histogram of the distances between each ribbon and its five nearest ribbons for all ribbons contained in the three RBCs. x0 shows the mean of the distribution. Please note that the y-axis for B and C have different sizes, given that they represent the number of occurrences of each event.

We examined the distribution of these 30–41 synaptic ribbons within the zebrafish RBC terminals. The distribution of RBC ribbons as estimated by the distance between ribbon sites (Fig. 9B and Video 1; see Materials and Methods) revealed a wide distribution in synaptic ribbon distance to its nearest neighbor across the three reconstructed RBCs (means of 2.48 ± 0.12 µm, 1.84 ± 0.06 µm and 1.98 ± 0.08 µm; Fig. 9C). The distance between RBC synaptic ribbons ranged from 141 nm to 6 µm, with a mean of 2 ± 0.05 µm (Fig. 9D; across all three reconstructed RBCs). Of the 510 zebrafish RBC synaptic ribbons, only 4 (0.8%) were separated by distances smaller than our confocal microscope resolution limit of 270 nm.

To reveal whether heterogeneity in synaptic Ca2+ signals correlates with active zone size, we compared the shape and size of the ribbon active zones across the three reconstructed RBCs. Our results show significant variations in the shape and size of zebrafish RBC synaptic ribbons in RBCs and associated active zones (Fig. 10). On average, individual ribbons spanned 2–5 consecutive sections, with some located within the axons or closer to the terminal. The SBF-SEM images and 3D projections of ribbon structures (colored in magenta) revealed considerable variability in shape and size across all dimensions, as illustrated in Figs. 10A–C. The number of Cav channels per active zone in hair cells from chicken, frog, and turtle varies with the size of the synaptic ribbon 76. We thus measured the area of the ribbon facing the plasma membrane where the Cav channels are known to be located to estimate the number of Cav channels per active zone and to determine whether variation in active zone size might plausibly contribute to the heterogeneity in Ca2+ signaling. The EM images and 3D projection of the plasma membrane (yellow) and the area of the ribbon facing the plasma membrane (cyan), representing the active zone, are shown in Figs. 10A-C. We observed large variability in the area of the ribbon facing the plasma membrane or active zone within the RBCs, with substantial variability in their average distribution across the three RBCs (Figs. 10C, cyan, 10E, Video 2 and Supplementary Fig. 9), suggesting that the number of Cav channels per RBC active zone could plausibly be heterogeneous.

Serial block-face scanning electron microscopy analysis reveals heterogenous RBC ribbon shape, size and area of the ribbon facing the plasma membrane.

A-C. EM images of zebrafish RBC ribbon structures (A) and their respective 3D reconstructions to illustrate different shapes and sizes of synaptic ribbons (A, B, and C, magenta), plasma membrane (A and B, yellow), and the area of the ribbon facing the plasma membrane (A, B, and C, cyan). Each of the 5 rows of images illustrates one ribbon synapse from a zebrafish retinal RBC. A 3D reconstruction of the RBC synaptic terminal and ribbon from SBF-SEM stacks is provided in Video 2.

D. 3D reconstruction of the RBC terminal closest to the ganglion cell layer resembles the shape and the size of the mammalian RBC1 2. The ribbons are colored magenta.

E. Summary of the active zone size, the area of the ribbon associated with the plasma membrane measured in serial sections across the three RBCs from Fig 9A. The individual distribution of the three RBCs active zone sizes is provided in Supplementary Fig. 8. The z-size of the SBF-SEM sections is 50 nm, and each ribbon spans 2-5 consecutive sections. The solid cyan circles in the violin plots show individual synaptic ribbon measurements from three RBCs, with the average measurements shown in the solid black circle.

The size of the active zone and maximal Ca2+ influx correlate with the size of the synaptic ribbon

To reveal whether larger ribbons display stronger maximal Ca2+ influx, we measured Ca2+ signals in response to a series of 200-ms depolarizations. We first imaged TAMRA-RBP and Cal520LA-RBP in sequential scans to obtain ribbon location and resting Ca2+ levels. To maximize the capture of Ca2+ signals during brief stimuli, we image the Cal520LA-RBP channel, followed by TAMRA-RBP and Cal520LA-RBP sequential scans to confirm the ribbon locations. We found that the maximum amplitude of depolarization-evoked Ca2+ signals increased with an increase in ribeye fluorescence (Fig. 11A & B and Video 3; r = 0.51, N = 121 ribbons, p< 0.001), consistent with findings reported in cochlear inner hair cells 51,56. Since ribeye fluorescence correlates with the number of ribeye molecules per ribbon, and, therefore, ribbon size 26,51,56,77, our findings suggest that larger ribbons display stronger Ca2+ signals as active zone size scales with ribbon size 56. Given the diverse shapes of RBC synaptic ribbons (Fig. 10A & B, magenta), we measured the longest length and width of ribbons from EM images and compared this with the active zone size. We found that RBC ribbon length and width have a moderate positive correlation (Fig. 11C, r = 0.47, N = 102 ribbons, p< 0.001), and both dimensions have a moderate positive correlation with active zone size, albeit the ribbon width (Fig. 11D, r = 0.52, N = 102 ribbons, p< 0.001) has a stronger correlation with active zone size than the ribbon length (Fig. 11D, r = 0.32, N = 102 ribbons, p< 0.001). These results suggest that heterogeneity in synaptic Ca2+ signals correlates with ribbon dimensions and that active zone and the number of Cav channels 51,56 scale with ribbon size.

Heterogeneity of Ca2+ microdomains in RBC terminals.

Larger ribbons have stronger maximal Ca2+ influx and larger CAZ. A. Images show voltage-clamped RBC filled with a solution containing TAMRA-RBP to label ribbons before depolarization (left, magenta) and Cal520LA-RBP to measure the amplitude of Ca2+ influx (middle, cyan) during depolarization and superimposed of the two (right) to compare the size or ribbon vs. maximal Ca2+ influx. A time-lapse movie of the synaptic terminal of a zebrafish bipolar neuron during Ca2+ influx is provided in Video 3. B. Scatter plot of maximal Ca2+ influx (F/Frest) vs. TAMRA-RBP. Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. C. Scatter plot of maximal synaptic ribbon length vs. maximal synaptic ribbon width as estimated from SBF-SEM images. Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. D. Scatter plot of area of AZ vs. ribbon width (filled black circles) and of area of AZ vs. ribbon length (filled grey circles). Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. AZ, Active Zone.

We note that a large correlation between the inferred ribbon size and active zone size would be expected if the size heterogeneity is partially due to the differences in the degree of labeling by the two ribbon-binding peptides. Therefore, the observed significant but moderate degree of correlation (r=0.32 or r = 0.52) between the active zone area and the ribbon length or width provides an additional indication of large heterogeneity in the calcium influx between ribbons of similar size, increasing further the heterogeneity in neurotransmitter release by different ribbons.

Discussion

A deep look at the active zone Ca2+ microdomain in RBC terminals

In this study, we developed a quantitative nanophysiology approach using ribeye-bound Ca2+ indicators and fluorescently labeled RBP, combined with ratiometric dual-color imaging, to measure changes in Ca2+ concentration at and along the ribbon axis. By targeting Ca2+ indicators to the ribbon combined with fluorescently labeled RBP, we determined the Ca2+ concentration along the ribbon at different distances from active zone and found that near the ribbon base the Ca2+ concentration could increase to an average of 26 μM upon depolarization. Given the localization of our indicator on the ribbon and minimal sensitivity to exogenous buffers, we believe that our ribbon-proximal signal represents Ca2+ concentrations on the ribbon in locations very near release sites, likely some of those that contribute to vesicle resupply. Notably, the [Ca2+] i levels obtained from our higher resolution approach using confocal microscopy are similar to those that were reported for measurements in hair cells using Stimulated Emission Depletion (STED) lifetime measurements 36. Previous work on RBCs found slowing of recovery from paired-pulse depression by EGTA, suggesting that Ca2+ levels at distal sites on the ribbon might be important for the recruitment of new vesicles following depletion of the RRP and UFRP 11,24. Here, we measured directly the effect of EGTA on the spread of Ca2+ signals and found that the Ca2+ signal at the distal half of the ribbon was two-fold lower than that at the location proximal to the membrane under conditions of low Ca2+ buffering. It should be noted, however, that diffraction of light during the optical imaging procedure is likely to underestimate spatiotemporal [Ca2+] along the ribbon. Therefore, we used our measurement of the Ca2+ current per ribbon to develop a detailed computational model that allowed us to resolve the expected profile of Ca2+ microdomains governing UFRP fusion. The model results shown in Fig. 7 reveal, on a finer spatial scale, the distribution of Ca2+ around a ribbon in the presence of various quantities of immobile endogenous buffer and added exogenous buffers. Our finding of high proximal [Ca2+] levels on the order of 26 µM recorded using a ribeye-bound low-affinity indicator dye is in qualitative agreement with the simulated high concentration of Ca2+ in the microdomain at the base of the ribbon. Simulations also show that, as expected, the peak Ca2+ microdomain amplitude is relatively insensitive to the concentration of endogenous immobile buffer and added EGTA, but its decay with distance is accelerated by increasing the concentrations of exogenous and endogenous buffers. In contrast, the addition of 2 mM BAPTA causes a profound reduction of the Ca2+ microdomain amplitude and localizes the signal to a small volume around the ribbon. We also compared the impact of mobile vs immobile endogenous buffers on the Ca2+ distribution, showing the much greater effect of mobile buffers on the microdomain amplitude and distance dependence. Although these results are generally expected 70–72, the simulation results estimate the [Ca2+] levels in the vicinity of the ribbon with greater spatial and temporal resolution. As expected, the ratios of Ca2+ concentration values at the base of the ribbon vs. further away from the ribbon are much greater in the model simulation than our experimentally determined proximal-to-distal Ca2+ ratio values due to the large size of the microscope PSF, which precludes precise localization of the corresponding Ca2+ signals. Further, simulations show the effect of exogenous buffers with greater spatial resolution. However, the simulations fully agree with the overall increase of the proximal-to-distal Ca2+ concentration ratios by the application of exogenous Ca2+ buffers, especially BAPTA. Despite the technical challenges posed by the limited spatiotemporal resolution of optical imaging, the spatial resolution attained with the nanophysiology approach developed in our study is sufficient to differentiate Ca²⁺ signals associated with distal sites on the ribbon from those localized near the plasma membrane. Conceivably, the nanophysiology approaches established here could generally be applied to other classes of ribbon-containing cells, such as rods, cones and hair cells.

Heterogeneity of Ca2+ microdomains across RBCs terminals

Our quantitative nanophysiology approach provides evidence for the heterogeneity in synaptic Ca2+ signals across zebrafish RBCs, suggesting variability in the Ca2+ microdomains as has been previously reported for hair cells 51,56 and cone photoreceptors. Retinal bipolar neurons contain many ribbon active zones. This ranges from 45-60 in adult goldfish, as quantified by EM 75, or 25-45 in RBCs of adult living zebrafish, as estimated from laser scanning confocal microscopy 78. Our analyses of ribbon number across zebrafish RBCs through SBF-SEM arrived at an estimate of 31-40. The differences in the number of ribbons as determined through confocal microscopy compared to SBF-SEM could be explained by the resolution limit encountered by the confocal image analyses. Our analyses of zebrafish RBC terminal SBF-SEM images and 3D projections provide the first evidence of diversity in synaptic ribbon shape and sizes across zebrafish RBC terminals. Further, our quantitative measurements of SBF-SEM images reveal that synaptic ribbon width and length correlate with the active zone size, consistent with previous findings in hair cell ribbon synapses 76,77 and cone photoreceptor ribbon synapses52. Our observation that maximal active zone Ca2+ influx correlates with ribeye fluorescence in live imaging is consistent with findings previously reported in hair cells 56. Thus, we expect larger active zones, with more Ca2+ influx, to have more Cav channel expression, as demonstrated for hair cell ribbon synapses 56. Heterogeneity among ribbon active zones has been previously reported in hair cells where synaptic Ca2+ microdomains varied substantially in amplitude and voltage-dependence within individual inner hair cells 51,56. Studies in aquatic tiger salamander cone photoreceptors revealed that the amplitude and midpoint activation voltage of Ca²⁺ signals varied across individual ribbons within the same cone. Additionally, local Ca²⁺ signal dynamics at cone photoreceptor ribbons were found to be independently regulated 52. However, in comparing our findings with studies of ribbon size heterogeneity in hair cells we note that we do not observe any systematic variation of ribbon size with larger-scale morphological features like the cell position, as compared to the reported interesting tonotopic variation of ribbon size with position along the cochlea 51,56,

The faster, smaller, and more spatially confined Ca2+ signals that are insensitive to the application of high concentrations of exogenous Ca2+ buffers, referred to here as ribbon proximal Ca2+ signals, could be due to Ca2+ influx through Cav channel clusters beneath the synaptic ribbon 19,26,29–42. However, the variability in Ca2+ signals at the distal ribbon, away from the plasma membrane, may result from the spread between two closely spaced microdomains; thus, if proximal Ca2+ signals are variable, the distal Ca2+ signals are also variable between ribbons. Alternatively, the variability in distal Ca2+ signals could be due to additional mechanisms, for example, Ca2+ influx from internal stores. Future studies should aim to elucidate the specific mechanisms underlying the observed heterogeneity in distal Ca2+ signals. For instance, exploring differences in Ca2+ release from intracellular stores or variations in Ca2+ sequestration, as reported in goldfish, could reveal key contributing factors 79. Although the potential impact of these intracellular Ca2+ stores on Ca2+ microdomain heterogeneity in hair cells has not been reported,51 previous studies in goldfish bipolar cell terminals have documented spatially restricted Ca2+ oscillations in voltage-clamped retinal bipolar cells, occurring independently of membrane potential 20. To our knowledge, the sources of Ca2+ for these oscillations remain unknown. However, previous studies in goldfish bipolar cells suggest that the endoplasmic reticulum79 and mitochondria18 may act as internal Ca2+ stores. Notably, IP₃ receptors have been localized to retinal bipolar cell terminals in some species.80,81

In the current study, we investigated possible mechanisms for heterogeneity in proximal Ca2+ signals by measuring the ribbon size, particularly the area of the ribbon adjacent to the plasma membrane, where Cav channel clusters are located. Our SBF-SEM shows substantial variability in synaptic ribbon size, shape, and the area of the ribbon facing the plasma membrane where Cav channel clusters are tethered. Since both Ca2+ signals and the area of the synaptic ribbon facing the plasma membrane are heterogeneous, we propose that the larger ribbons may anchor more channels, leading to larger Ca2+ microdomains. Since local Ca2+ signals control kinetically distinct neurotransmitter release components, heterogeneity in local Ca2+ signals may alter the rate of vesicle release and allow them to function independently. Indeed, heterogeneity in neurotransmitter release kinetics has been proposed for the observed diversity in excitatory postsynaptic current amplitudes and kinetics reported from paired recordings of goldfish bipolar cells82 and ON-type mixed RBCs 83. The heterogeneity in the RBC Ca2+ microdomains, synaptic ribbon shape, size, and active zone area reported in this study may contribute to regulating the dynamic range of RBC ribbon synapses22,84,85 – a hypothesis that needs to be tested in future studies.

Methods

Rearing of zebrafish

Male and female zebrafish (Danio rerio; 16 ∼20 months) were raised under a 14 h light/10 h dark cycle and housed according to NIH guidelines and the University of Tennessee Health Science Center (UTHSC) Guidelines for Animals in Research. All procedures were approved by the UTHSC Institutional Animal Care and Use Committee (IACUC; protocol # 23-0459).

Isolation of zebrafish retinal RBCs

Dissociation of RBCs was performed using established procedures 86. Briefly, retinas were dissected from zebrafish eyes and incubated in hyaluronidase (1100 units/ml) for 20 minutes. The tissue was washed with a saline solution containing 120 mM NaCl, 2.5 mM KCl, 0.5 mM CaCl2, 1 mM MgCl2, 10 mM glucose, and 10 mM HEPES, pH = 7.4] before being cut into quadrants. Each quadrant was incubated at room temperature for 25 to 40 min in the same saline solution, to which was added DL-cysteine and papain (20-30 units/ml; Sigma Millipore, St. Louis, MO) and triturated using a fire-polished glass Pasteur pipette. Individual cells were transferred to glass-bottomed dishes, allowed to attach for 30 minutes, and washed with saline solution before being used for experiments.

Ribeye binding peptides

As a means of localizing the ribbons, custom peptides containing the ribbon binding sequence fused to TAMRA (tetramethylrhodamine; TAMRA: GIDEEKPVDLTAGRRAG) dye were synthesized, purified, and purchased from LifeTein (>95% purity, LifeTein LLC, NJ).

Ca2+ indicators

Free Ca2+ indicators

The potassium salts of high and low-affinity Ca2+ indicators Cal-520® (high affinity, KD 320 nM) and Cal-520N™ AM (low affinity, KD 90 µM), referred to as Cal520HA and Cal520LA, respectively were purchased from AAT Bioquest.

Direct conjugation of Ca2+ indicator to cysteine-containing ribbon-binding peptides

To target Ca2+ indicators to the ribbon, custom-made cysteine-containing ribeye binding peptides NH2-CIEDEEKPVDLTAGRRAC-COOH were synthesized and purchased from LifeTein to directly fuse with the fluorogenic 520® maleimide (purchased from AAT Bioquest) for high affinity (HA) and low affinity (LA Ca2+ indicator dyes). Each peptide at one mM concentration was mixed and incubated with two mM Cal-520® maleimide (20 mM stock solution in DMSO purchased from AAT Bioquest) for one hour at room temperature, then overnight at 4°C. Calibration of Ca2+ indicator dyes was performed as described 19 and as detailed below. The conjugated Ca²⁺ indicators were stored at –20°C in smaller aliquots, each sufficient for a single day’s experiments.

Measurement of dissociation constants (KD) for Ca2+ indicator peptides

The effective KD (Keff) was obtained by measuring the fluorescence of Cal520HA-RBP and TAMRA-RBP in buffered Ca2+ solutions, determining the ratio between them 19, and using the Grynkiewicz equation to determine the Ca2+ concentration [Ca2+] from this ratio 53. However, this was not possible for the low-affinity indicator Cal520LA due to the large Ca2+ levels required to calibrate it. Thus, Keff for Cal520LA-RBP could be larger than the K1/2 provided by the manufacturer for Cal520LA (KD, 90 μM), as reported previously with Keff, measurements of OB-5N in inner hair cells 36. Thus, our estimates of local Ca2+ concentrations obtained using Cal520LA represent the lower bounds of the underlying true values. As noted in the Results section, the same may be true for the high-affinity Cal520HA despite its accurate Keff estimate, due to potential dye saturation effects.

RBC voltage clamp recording

Whole-cell patch-clamp recordings were made from isolated RBCs, as described previously 19,27,87. Briefly, a patch pipette containing pipette solution (120 mM Cs-gluconate, 10 mM tetraethyl-ammonium-Cl, 3 mM MgCl2, 0.2 mM N-methyl-d-glucamine-EGTA, 2 mM Na2ATP, 0.5 mM Na2GTP, 20 mM HEPES, pH = 7.4) was placed on the synaptic terminal, as described previously. The patch pipette solution also contained a fluorescently-labeled RBP peptide (TAMRA-RBP) to mark the positions of the ribbons and either 1) free Ca2+ indicators Cal520HA-free (Figs. 1 and 2) and Cal520LA-free (Fig. 2, 3 and 5) to demonstrate our nanophysiological approach or 2) ribeye-bound Ca2+ indicators Cal520HA-RBP (Figs. 3 and Supplementary Figs. 3 and 5) and Cal520LA-RBP (Figs. 3, 4, 5, and 8) to measure local ribbon-associated Ca2+ signals. Current responses from the cell membrane were recorded under a voltage clamp with a holding potential (VH) of –65 mV that was stepped to 0 mV (t0) for 10 milliseconds. These responses were recorded with a patch clamp amplifier running PatchMaster software (version v2×90.4; HEKA Instruments, Inc., Holliston, MA). Membrane capacitance, series conductance, and membrane conductance were measured via the sine DC lock-in extension in PatchMaster and a 1600 Hz sinusoidal stimulus with a peak-to-peak amplitude of 10 mV centered on the holding potential 88.

Acquisition of confocal images

Confocal images were acquired using an Olympus model IX 83 motorized inverted FV3000RS laser-scanning confocal microscopy system (Olympus, Shinjuku, Tokyo, Japan) running FluoView FV31S-SW software (Version 2.3.1.163; Olympus, Center Valley, PA) equipped with a 60 X silicon objective (NA 1.3), all diode laser combiner with five laser lines (405, 488, 515, 561 & 640 nm), a true spectral detection system, a hybrid galvanometer, and a resonant scanning unit. Fluorescently labeled ribeye binding peptide (RBP) 26 and Ca2+ indicator were delivered to RBC via a whole-cell patch pipette placed directly at the cell terminal. We waited for 30 s after break-in to allow Cal520HA to reach equilibrium with the patch-pipette before obtaining the first fluorescence image. Rapid x-t line scans at the ribbon location were performed to localize synaptic ribbons (Fig.1Aii) and to monitor local changes in Ca2+ concentration at a single ribbon, as we demonstrated previously, to estimate the Ca2+ levels at the plasma membrane and to track a single synaptic vesicle at ribbon locations 19,20. The z-projection from a series of confocal optical sections through the synaptic terminal (Fig. 1Ai) illustrates ribbon labeling (magenta spots). RBP fluorescence was used to localize a synaptic ribbon and to define a region for placing a scan line perpendicular to the plasma membrane, extending from the extracellular space to the cytoplasmic region beyond the ribbon to monitor changes in the Ca2+ concentration along the ribbon axis (Fig. 1Aii). The focal plane of the TAMRA-RBP signal was carefully adjusted for sharp focus to avoid potential errors arising from the high curvature near the top of the terminal and the plane of membrane adherence to the glass coverslip at the bottom of the terminal. Sequential dual laser scanning was performed at rates of 1.51 milliseconds per line. Two-color laser scanning methods allowed observation of Ca2+ signals (Fig.1C, cyan) throughout the full extent of the ribbon in voltage-clamped synaptic terminals, while the ribbon and cell border were imaged with a second fluorescent label. The Exchange of TTL (transistor-transistor logic) pulses between the patch-clamp and imaging computers synchronized the acquisition of electrophysiological and imaging data. The precise timing of imaging relative to voltage-clamp stimuli was established using PatchMaster software to digitize horizontal-scan synced pulses from the imaging computer in parallel with the electrophysiological data (Fig.1B). Acquisition parameters, such as pinhole diameter, laser power, PMT gain, scan speed, optical zoom, offset, and step size were kept constant between experiments. Sequential line scans were acquired at 1-2 millisecond/line and 10 μs/pixel with a scan size of 256 × 256 pixels.

Bleed-through between the channels was confirmed with both lasers using the imaging parameters we typically use for experiments. To test bleed-through from the RBP channel (TAMRA-RBP) to the Ca2+ indicator (Cal-520HA and Cal-520LA), whole-cell recordings from RBC terminals were performed with patch pipette solution that contained TAMRA-RBP or the aforementioned Ca2+ indicators, and line-scan images were collected and analyzed using the same procedures used for experimental samples.

Point spread function

The lateral and axial point spread function (PSF) was obtained as described previously 19,28,86. Briefly, an XYZ scan was performed through a single 27 nm bead and the maximum projection in the xy-plane was fit to the Gaussian function using Igor Pro software. We obtained the full width at half maximum (FWHM) values for x and y-width of 268 and 273 nm, respectively, in the lateral (x-y plane) and for y-z width of 561 nm in the axial (y-z axis) resolution.

Photobleaching

We minimized photobleaching and phototoxicity during live-cell scanning by using fast scan speed (10 microsec/pixel), low laser intensity (0.01–0.06% of maximum), and low pixel density (frame size, 256 × 256 pixels). We estimated photobleaching using x-t line scans of Cal520HA or Cal520LA and TAMRA-RBP in the absence of stimulation, with the same imaging parameters used for experimental samples.

Data analysis

Quantitative FluoView x-t and x-y scans were analyzed initially with ImageJ software (imagej.nih.gov) and subsequently with Igor Pro software (Wavemetrics, Portland OR) for curve fitting and production of the figures. Data from PatchMaster software were initially exported to Microsoft Excel (Version 16.81) for normalizing and averaging and exported from MS Excel to Igor Pro (Version 9.05) for curve fitting and production of the l figures.

Analysis of x-t scan data

X-axis profile

To determine the Ca2+ signals along the ribbon axis, we spatially averaged the x-axis profile intensity of RBP (i.e., a horizontal row of pixels, see Fig.1C, top) to determine the position of the center of the ribbon and estimate the location of the plasma membrane. The parameter x0 is the peak of the Gaussian fit, giving the x-position of the center of the ribbon. The x-axis profile is also used to obtain the spatial profile of Ca2+ signals before, during, and after with respect to the ribbon profile. To identify the Ca2+ signals specific to the ribbon location, we fit x-axis intensity profiles with the equation f(x) = s(x) + g(x), where s(x) is a sigmoid function that describes the transition from intracellular to extracellular background fluorescence at the edge of the cell, given by s(x) = b – c / (1– exp((x1/2 – x)/d)), and g(x) is a Gaussian function that represents the fluorescence of RBP, given by g(x) = a exp(−(x − x0)2/w2), as described 27. The parameters x1/2and x0 were taken as the x-axis positions of the plasma membrane and the fluorescence emitter, respectively. While parameter b is intracellular background fluorescence, c is extracellular background fluorescence, d is the slope of the sigmoid, a is the peak amplitude of emitter fluorescence, and w is √2 times the standard deviation of the Gaussian function, in practice, the latter parameters were highly constrained by the data or by the measured PSF, leaving only x1/2and x0 as free parameters in the fitting. Fig.1D, demonstrates that the peak of the Ca2+ signals (x$, cyan; Fig.1D) during the stimuli is proximal to the ribbon center (x0, magenta; Fig.1D) towards the plasma membrane (x1/2, magenta; Fig.1D), as expected for Ca2+ influx originating from Ca2+ channels localized in the plasma membrane 44–46.

T-axis profile

The temporal profiles of the Cal520, Cal520-RBP, and TAMRA-RBP signals were determined by analyzing the time-axis profile of the x-t line scan to obtain the kinetics of the Ca2+ transient with respect to the ribbon. We determined the baseline kinetics by averaging the fluorescence obtained immediately before depolarization. The timing of depolarization and the amplitude of the Ca2+ current was obtained from the PatchMaster software. The rising phase of the Ca2+ transient was fit with the sigmoid function, the peak of which is referred to as the peak amplitude Ca2+ transient.

We used this baseline profile of ribbon-proximal Ca2+ signals to distinguish signals proximal or distal to the ribbon despite the distance between these two signals being within the PSF. For example, the ribbon-proximal signals were obtained by averaging 5 pixels of the temporal profile of the Ca2+ signals (obtained with Cal520 or Cal520-RBP) between the x1/2 and x0 values obtained in the x-axis profile of TAMRA-RBP. The distal profile was obtained similarly but was 5-pixels after x0 towards the cytoplasm.

Quantifying the kinetics of the Ca2+ transients

The rising and decay phases of the Ca2+ fluorescence traces were fit independently, as described below, with the peak of the average transient defining the transition between these two phases. The rise-time of the fluoresence transients (Figs. 2A and 2B) was defined as the time to reach half-maximal value from the start of the stimulation, after subtracting its baseline. It was determined by fitting a hyperbolic function with an argument shifted by a constant value β, which serves as one of two free parameter of the fit, along with the argument scale factor α characterizing the steepness of the rising phase of the signal: The constant C = tanℎ(α ∗ β) ensures that frise ((0) = 0 at the start of the stimulation, t = 0. This fitting function reaches its peak value fpeak at t = tpeak, as determined from the averaged fluorescence trace. The sign of the argument shift β controls the shape of the signal near t = 0: β < 0 corresponds to a signal which is concave downward (i.e. very rapid onset), while β > 0 corresponds to a signal which is concave upward near t = 0 (i.e. gradual rise). Therefore, this function provides a more flexible functional relationship than the Hill function or the standard sigmoidal function. The decay phase of the fluorescence transient was fit using the standard bi-exponential sum with time constants τfast and τslow: where afast and aslow are the relative proportions of the two decay components, specified as percentages and satisfying afast + aslow = 100%. Both the rise and the decay parameters were found using a global differential evolution optimization algorithm 89, repeating the optimization 400 times starting with different initial parameters values, to ensure true global minimum of the fit error is reached.

We note that the rise time estimate is very sensitive to noise in the data, given only a small number of data points during the short rising phase of the signal, while the decay parameters are also sensitive to the duration of the recording. Therefore, much higher signal-to-noise ratio would be required for reliable uncertainty quantification of the corresponding kinetic fit parameters. Still, time-course fit of the averaged traces provides a valuable qualitative-level comparison of stimulus-evoked fluorescence transients recorded with different dyes at different locations with respect to the ribbon.

Analysis of x-y scan data

We analyzed the rate of loading the TAMRA-RBP and Ca2+ indicator into the RBC terminal with ImageJ software by placing a square region of interest (ROI; 5 x 5 μm) and using a Plot z-axis profile function to obtain spatially averaged TAMRA-RBP and Ca2+ indicator fluorescence as a function of time. The rising phase of TAMRA-RBP and Ca2+ indicator fluorescence was obtained by fitting the rising to the peak to the single exponential function using the curve fitting function in Igor Pro 9 software. The rate of the exponential function is defined as the rate of fluorescence loading to the terminal.

For analysis comparing the fluorescence intensity of ribbons and Ca2+ influx elicited by 100-ms stimuli (Fig. 11A & B), 10 images of both TAMRA-RBP and Cal520LA-RBP were collected in sequential, followed by imaged Cal520LA-RBP only during 200-ms depolarization, and the sequence ended with 10 images of both TAMRA-RBP and Cal520LA-RBP in sequential to confirm the ribbon locations. Two color imaging were obtained sequentially with a frame interval of 407 ms, and Cal520LA-RBP only during 100ms-stimuli was 205 ms. Images before and after were then averaged to obtain the ribbon location. Synaptic ribbon fluorescence was quantified as the ratio of TAMRA fluorescence to the fluorescence of the nearby RBC cytoplasm, measured approximately eight to nine pixels away (Fribbon/Fnear by), measuring the pixel with the highest intensity. The change in Cal520LA-RBP fluorescence, used as a proxy for active zone Ca2+ influx, was estimated as ΔF/Frest, where Frest is the fluorescence intensity before depolarization at −65 mV, and ΔF is the difference when depolarized to 0 mV. Ca2+ indicator intensity was calculated as the average fluorescence of nine pixels centered on the pixel with the greatest fluorescence increase.

Experimental design and statistical methods

We did not use any statistical methods to determine the sample size prior to experiments. Mean value variances were reported as ± the standard error of the mean (SEM). The statistical significance of differences in average amplitudes of Ca2+ current, capacitance, synaptic ribbon size and number, and Ca2+ transients were assessed using unpaired, two-tailed t-tests with unequal variance using R Studio (Version 2023.09.0+463) and Igor Pro software.

Amplitude analysis for Figure 8

Amplitude data were processed and analyzed using R Studio software (Version 2023.09.0+463). The data were divided into two groups, that of individual cells and of multiple cells, and analyzed as follows. Multiple individual cells from each condition (proximal 10 mM EGTA Cal520LA-RBP and distal 10 mM EGTA Cal520LA-RBP) were analyzed to compare the variability between the Ca2+ amplitude measurements of the ribbons from each cell. Having verified the normality of the data to ensure the validity of the statistical test, we used one-way ANOVA and Tukey’s honest significant difference test to assess the specific statistical differences between ribbons. To compare the variability between the amplitude measurements of different cells, we compared readings for various cells. If there were multiple measurements for a single ribbon in a specific cell, those readings were averaged. In this case, the normality of the data was again verified and Welch’s ANOVA was employed to assess the between-group differences as it does not assume homogeneity of variance, with the Games-Howell post-hoc test being used to understand the specific differences between cells.

Computational modeling of Ca2+ dynamics

Spatio-temporal Ca2+ dynamics is simulated in a volume shown in Fig. 9, which is a 3D box of dimensions (1.28 x 1.28 x 1.1) µm3. Assuming an approximately spherical bouton of diameter 5µm and an average of 36 ribbons per terminal, one obtains synaptic volume per ribbon of 1.81 µm3, which matches the volume of our box domain. Ribbon serves as an obstacle to Ca2+ and buffer diffusion, and has an ellipsoidal shape with semi-axis of 190nm along the Y– and Z-axes (for a total height of 380nm), and a semi-axis of 70 nm along the X-axis. It is attached to the plasma membrane by a thin ridge (arciform density) of dimensions 60×30×30 nm. Ca2+ ions enter through 4 channels (or clusters of channels) at the base of the ribbon indicated by black disks in Fig. 6, forming a square with a side length of 80 nm; this highly simplified channel arrangement is sufficient to capture the level of detail in spatial [Ca2+] distribution that we seek to resolve. Each of the four clusters admits a quarter of the total Ca2+ current. Ca2+ current values and current pulse durations are listed in Fig. 7. Although the actual distribution of Cav channels is more heterogeneous, channels close to the ribbon play a greater role in neurotransmitter release 2.

The simulation includes a single dominant Ca2+ buffer with molecules possessing a single Ca2+ ion binding site. We assume that only immobile buffer is present in a patch-clamped cell, since any mobile buffer would diffuse into recording pipette. Simulations with mobile buffer in an un-patched cell are also provided (Supplementary Fig. 4B&C). Buffer parameter values were set to values reported by 7, namely dissociation constant of 2 µM, and total concentration of 1.44 mM (total buffering capacity at rest of 720). We also examine the impact of the assumption on buffer concentration by repeating the simulation with a lower buffering capacity of 100 (total buffer concentration of 200 µM). The Ca2+ ion diffusivity is set to 0.22 µm2/ms 90.

To minimize the number of undetermined parameters, we assumed the same parameters for the Ca2+ clearance on all surfaces, with one high-capacity but lower affinity process simulating Na+/Ca2+ exchangers, and one lower capacity but higher affinity process simulating Ca2+ extrusion by ATPase PMCA and SERCA pumps. We used the Hill coefficient of m=2 for the Ca2+ sensitivity of the SERCA pumps on all surfaces except the bottom surface 91, whereas Hill coefficient of m=1 corresponds to PMCA pumps on the bottom surface 92. Thus, the flux of [Ca2+] across the surface (ions extruded per unit area per unit time) is described by where C is the Ca2+ concentration, DC is the Ca2+ diffusivity, and n ⋅ ∇c is the gradient of [Ca2+] normal to the boundary. The constant leak term ΦLeak ensures that the flux is zero at the resting value [Ca2+]rest = C0 = 0.1 µM. The Ca2+ clearance rate and affinity parameters for the pumps and exchangers are listed in Supplementary Table 3. We note however that the density of Na+-Ca2+ exchangers and PMCA/SERCA pumps are rough estimates and have a significant effect on results only for longer stimulation durations of over 100-200 ms. Reaction-diffusion equations for [Ca2+] and buffer are solved numerically using Calcium Calculator software version 6.10.7 93,94.

Serial block-face scanning electron microscopy (SBF-SEM)

The eyecups of adult (16 ∼20-month-old) zebrafish were dissected in bicarbonate-buffered Ames’ medium and, after removing the lens, the eyecups were fixed in 4% glutaraldehyde in 0.1 M sodium-cacodylate buffer at RT for one hour, followed by overnight fixation at 4°C. Thereafter, samples were treated to generate blocks for serial block face scanning electron microscopy (SBF-SEM) according to the published protocol 95, in this case using Zeiss 3-View SBF-SEM. Multiple 35 µm montages were acquired at a 5 nm X-Y resolution and a 50nm Z resolution across the entire retina from the outer (photoreceptor) layer to the inner (ganglion) cell layer. Images were inspected visually and structures of interest were traced using the TrakEM plug-in in ImageJ software (NIH/Fiji), as described below.

Analysis of serial block-face scanning electron microscopy images

SBF-SEM image stacks of the zebrafish retina were imported, aligned, visualized, and analyzed using the TrakEM plugin (version 1.5h) in ImageJ software (NIH). RBC1 were identified based on their characteristic morphology and 4-7 μm terminal size, as previously described 74. RBCs, synaptic ribbons, pre-synaptic membranes, and the area of the ribbon touching the plasma membrane were traced, painted using the TrakEM brush features, and rendered in 3D for structural visualization. The area of the synaptic ribbon touching the RBC membrane was quantified using the measuring tools in ImageJ and images were prepared using Adobe Photoshop software (version 25.9). The distance between each ribbon from three different RBCs and its nearest five ribbons (Fig. 9B and Video 1) was measured as follows: Ribbons on the same plane had their linear distance measured, whereas the distance between ribbons that were close to each other but in different layers was calculated using the Pythagorean Theorem.

eLife Assessment

The study introduces new tools for measuring the intracellular calcium concentration close to transmitter release sites, which may be relevant for synaptic vesicle fusion and replenishment. This approach yields important new information about the spatial and temporal profile of calcium concentrations near the site of entry at the plasma membrane. This experimental work is complemented by a coherent, open-source, computational model that successfully describes changes in calcium domains. Some of the conclusions are strongly supported by the data, but a few gaps in the data presented mean that the evidence for other conclusions is incomplete.

Reviewer #1 (Public review):

This paper describes technically-impressive measurements of calcium signals near synaptic ribbons in goldfish bipolar cells. The data presented provides high spatial and temporal resolution information about calcium concentrations along the ribbon at various distances from the site of entry at the plasma membrane. This is important information. Important gaps in the data presented mean that the evidence for the main conclusions is currently inadequate.

Strengths

• The technical aspects of the measurements are impressive. The authors use calcium indicators bound to the ribbon and high speed line scans to resolve changes with a spatial resolution of ~250 nm and temporal resolution of less than 10 ms. These spatial and temporal scales are much closer to those relevant for vesicle release than previous measurements.

• The use of calcium indicators with very different affinities and of different intracellular calcium buffers helps provide confirmation of key results.

Weaknesses

• Multiple key points of the paper lack a statistical test or summary data from populations of cells. For example, the text states that the proximal and distal calcium kinetics in Figure 2A differ. This is not clear from the inset to Figure 2A - where the traces look like scaled versions of each other. Values for time to half-maximal peak fluorescence are given for one example cell but no statistics or summary are provided. Figure 8 shows examples from one cell with no summary data. This issue comes up in other places as well.

• The rise time measurements in Figure 2 are very different for low and high affinity indicators, but no explanation is given for this difference. Similarly, the measurements of peak calcium concentration in Figure 4 are very different with the two indicators. That might suggest that the high affinity indicator is strongly saturated, which raises concerns about whether that is impacting the kinetic measurements.

Reviewer #2 (Public review):

Summary:

The study introduces new tools for measuring intracellular Ca2+ concentration gradients around retinal rod bipolar cell (rbc) synaptic ribbons. This is done by comparing the Ca2+ profiles measured with mobile Ca2+ indicator dyes versus ribbon-tethered (immobile) Ca2+ indicator dyes. The Ca2+ imaging results provide a straightforward demonstration of Ca2+ gradients around the ribbon and validate their experimental strategy. This experimental work is complemented by a coherent, open-source, computational model that successfully describes changes in Ca2+ domains as a function of Ca2+ buffering. In addition, the authors try to demonstrate that there is heterogeneity among synaptic ribbons within an individual rbc terminal.

Strengths:

The study introduces a new set of tools for estimating Ca2+ concentration gradients at ribbon AZs, and the experimental results are accompanied by an open-source, computational model that nicely describes Ca2+ buffering at the rbc synaptic ribbon. In addition, the dissociated retinal preparation remains a valuable approach for studying ribbon synapses. Lastly, excellent EM.

Comments on revisions:

Specific minor comments:

(1) Rewrite the final sentence of the Abstract. It is difficult to understand.

(2) Add a definition in the Introduction (and revisit in the Discussion) that delineates between micro- and nano-domain. A practical approach would be to round up and round down. If you round up from 0.6 um, then it is microdomain which means ~ 1 um or higher. Likewise, round down from 0.3 um to nanodomain? If you are using confocal, or even STED, the resolution for Ca imaging will be in the 100 to 300 nm range. The point of your study is that your new immobile Ca2-ribbon indicator may actually be operating on a tens of nm scale: nanophysiology. The Results are clearly written in a way that acknowledges this point but maybe make such a "definition" comment in the intro/discussion in order to: 1) demonstrate the power of the new Ca2+ indicator to resolve signals at the base of the ribbon (effectively nano), and 2) (Discussion) to acknowledge that some are achieving nanoscopic resolution (50 to 100nm?) with light microscopy (as you ref'd Neef et al., 2018 Nat Comm).

(3) Suggested reference: Grabner et al. 2022 (Sci Adv, Supp video 13, and Fig S5). Here rod Cav channels are shown to be expressed on both sides the ribbon, at its base, and they are within nanometers from other AZ proteins. This agrees with the conclusions from your imaging work.

(4) In the Discussion, add a little more context to what is known about synaptic transmission in the outer and inner retina.. First, state that the postsynaptic receptors (for example: mGluR6-OnBCs vs KARs-Off-BCs, vs. AMPAR-HCs), and possibly the synaptic cleft (ground squirrel), are known to have a significant impact on signaling in the outer retina. In the inner retina, there are many more unknowns. For example, when I think of the pioneering Palmer JPhysio study, which you sight, I think of NMDAR vs AMPAR, and uncertainty in what type postsynaptic cell was patched (GC or AC....). Once you have informed the reader that the postsynapse is known to have a significant impact on signaling, then promote your experimental work that addresses presynaptic processes: "...the new tool and results allow us to explore release heterogeneity, ribbon by ribbon in dissociated preps, which we eventually plan to use at ribbon synapses within slices......to better understand how the presynapse shapes signaling......".

Reviewer #3 (Public review):

Summary:

In this study, the authors have developed a new Ca indicator conjugated to the peptide, which likely recognizes synaptic ribbons and have measured microdomain Ca near synaptic ribbons at retinal bipolar cells. This interesting approach allows one to measure Ca close to transmitter release sites, which may be relevant for synaptic vesicle fusion and replenishment. Though microdomain Ca at the active zone of ribbon synapses has been measured by Hudspeth and Moser, the new study uses the peptide recognizing synaptic ribbons, potentially measuring the Ca concentration relatively proximal to the release sites.

Strengths:

The study is, in principle, technically well done, and the peptide approach is technically interesting, which allows one to image Ca near the particular protein complexes. The approach is potentially applicable to other types of imaging.

Weaknesses:

Peptides may not be entirely specific, and genetic approach tagging particular active zone proteins with fluorescent Ca indicator proteins may well be more specific. Although the authors are aware of this and the peptide approach is generally used for ribbon synapses, the authors should be aware of this, when interpreting the results.

Author response:

The following is the authors’ response to the previous reviews

Reviewer #1 (Public review):

This paper describes technically-impressive measurements of calcium signals near synaptic ribbons in goldfish bipolar cells. The data presented provides high spatial and temporal resolution information about calcium concentrations along the ribbon at various distances from the site of entry at the plasma membrane. This is important information. Important gaps in the data presented mean that the evidence for the main conclusions is currently inadequate.

Strengths

The technical aspects of the measurements are impressive. The authors use calcium indicators bound to the ribbon and high speed line scans to resolve changes with a spatial resolution of ~250 nm and temporal resolution of less than 10 ms. These spatial and temporal scales are much closer to those relevant for vesicle release than previous measurements.

The use of calcium indicators with very different affinities and of different intracellular calcium buffers helps provide confirmation of key results.

Thank you very much for this positive evaluation of our work.

Weaknesses

Multiple key points of the paper lack a statistical test or summary data from populations of cells. For example, the text states that the proximal and distal calcium kinetics in Figure 2A differ. This is not clear from the inset to Figure 2A - where the traces look like scaled versions of each other. Values for time to half-maximal peak fluorescence are given for one example cell but no statistics or summary are provided. Figure 8 shows examples from one cell with no summary data. This issue comes up in other places as well.

Thank you for this fair and valuable feedback. Following also the suggestion by the Editor, we have now removed the rise-time kinetic fitting results from the manuscript and only retain the bi-exponential decay time constant values. Further, we explicitly detail the issues with kinetic fitting, and state that the precise quantitative conclusions should not be drawn from the differences in kinetic parameters (pages 7 and 2728).

We have included the results of paired-t-tests to compare the amplitudes of proximal vs. distal calcium signals shown in Fig. 2A & B, Fig. 3C & D, Fig. 4C & D, Fig. 5A-D, and Fig. 8E&F. Because proximal and distal calcium signals were obtained from the same ribbons within 500-nm distances, as the Reviewer pointed out, “the traces look like scaled versions of each other”. For experiments where we make comparisons across cells or different calcium indicators, as shown in Fig. 3E & F, Fig.5E, and Fig. 8B&C, we have included the results of an unpaired t-test. We have also included the t-test statistics information in the respective figure legends in the revised version.

In Figure 8, we have shown example fluorescence traces from two different cells at the bottom of the A panel, and example traces from different ribbons of RBC a in the D, and the summary data is described in B-C and E-F, with statistics provided in the figure legends.

The rise time measurements in Figure 2 are very different for low and high affinity indicators, but no explanation is given for this difference. Similarly, the measurements of peak calcium concentration in Figure 4 are very different with the two indicators. That might suggest that the high affinity indicator is strongly saturated, which raises concerns about whether that is impacting the kinetic measurements.

Yes, we do believe that the high-affinity indicator is partially saturated, and therefore, the measurement with the low-affinity indicator dye is a more accurate reflection of the measured Ca2+ signal. We now state this more explicitly in the text. Further, we note that the rise time values are no longer listed due to lack of statistical significance for such comparisons, as noted above.

Reviewer #2 (Public review):

Summary:

The study introduces new tools for measuring intracellular Ca2+ concentration gradients around retinal rod bipolar cell (rbc) synaptic ribbons. This is done by comparing the Ca2+ profiles measured with mobile Ca2+ indicator dyes versus ribbon-tethered (immobile) Ca2+ indicator dyes. The Ca2+ imaging results provide a straightforward demonstration of Ca2+ gradients around the ribbon and validate their experimental strategy. This experimental work is complemented by a coherent, open-source, computational model that successfully describes changes in Ca2+ domains as a function of Ca2+ buffering. In addition, the authors try to demonstrate that there is heterogeneity among synaptic ribbons within an individual rbc terminal.

Strengths:

The study introduces a new set of tools for estimating Ca2+ concentration gradients at ribbon AZs, and the experimental results are accompanied by an open-source, computational model that nicely describes Ca2+ buffering at the rbc synaptic ribbon. In addition, the dissociated retinal preparation remains a valuable approach for studying ribbon synapses. Lastly, excellent EM.

Thank you very much for this positive evaluation of our work.

Comments on revisions:

Specific minor comments:

(1) Rewrite the final sentence of the Abstract. It is difficult to understand.

Thank you for pointing that out. We have updated the final sentence of the Abstract.

(2) Add a definition in the Introduction (and revisit in the Discussion) that delineates between micro- and nano-domain. A practical approach would be to round up and round down. If you round up from 0.6 um, then it is microdomain which means ~ 1 um or higher. Likewise, round down from 0.3 um to nanodomain? If you are using confocal, or even STED, the resolution for Ca imaging will be in the 100 to 300 nm range. The point of your study is that your new immobile Ca2-ribbon indicator may actually be operating on a tens of nm scale: nanophysiology. The Results are clearly written in a way that acknowledges this point but maybe make such a "definition" comment in the intro/discussion in order to: 1) demonstrate the power of the new Ca2+ indicator to resolve signals at the base of the ribbon (effectively nano), and 2) (Discussion) to acknowledge that some are achieving nanoscopic resolution (50 to 100nm?) with light microscopy (as you ref'd Neef et al., 2018 Nat Comm).

Thank you for the valuable comments. We have now provided this information in the introduction and discussion.

(3) Suggested reference: Grabner et al. 2022 (Sci Adv, Supp video 13, and Fig S5). Here rod Cav channels are shown to be expressed on both sides the ribbon, at its base, and they are within nanometers from other AZ proteins. This agrees with the conclusions from your imaging work.

Thank you for the valuable suggestion. We have now provided this information in the introduction and discussion.

(4) In the Discussion, add a little more context to what is known about synaptic transmission in the outer and inner retina.. First, state that the postsynaptic receptors (for example: mGluR6-OnBCs vs KARs-OffBCs, vs. AMPAR-HCs), and possibly the synaptic cleft (ground squirrel), are known to have a significant impact on signaling in the outer retina. In the inner retina, there are many more unknowns. For example, when I think of the pioneering Palmer JPhysio study, which you sight, I think of NMDAR vs AMPAR, and uncertainty in what type postsynaptic cell was patched (GC or AC....). Once you have informed the reader that the postsynapse is known to have a significant impact on signaling, then promote your experimental work that addresses presynaptic processes: "...the new tool and results allow us to explore release heterogeneity, ribbon by ribbon in dissociated preps, which we eventually plan to use at ribbon synapses within slices......to better understand how the presynapse shapes signaling......".

Thank you for the valuable comments. We have now provided this information in the introduction and discussion.

Reviewer #3 (Public review):

Summary:

In this study, the authors have developed a new Ca indicator conjugated to the peptide, which likely recognizes synaptic ribbons and have measured microdomain Ca near synaptic ribbons at retinal bipolar cells. This interesting approach allows one to measure Ca close to transmitter release sites, which may be relevant for synaptic vesicle fusion and replenishment. Though microdomain Ca at the active zone of ribbon synapses has been measured by Hudspeth and Moser, the new study uses the peptide recognizing synaptic ribbons, potentially measuring the Ca concentration relatively proximal to the release sites.

Strengths:

The study is, in principle, technically well done, and the peptide approach is technically interesting, which allows one to image Ca near the particular protein complexes. The approach is potentially applicable to other types of imaging.

Thank you very much for this appreciation.

Weaknesses:

Peptides may not be entirely specific, and genetic approach tagging particular active zone proteins with fluorescent Ca indicator proteins may well be more specific. Although the authors are aware of this and the peptide approach is generally used for ribbon synapses, the authors should be aware of this, when interpreting the results.

We acknowledge the reviewer’s point and believe the peptides and genetic approaches to measure local calcium signals have their merits, each with separate advantages and disadvantages.

Reviewer #1 (Recommendations for the authors):

The revisions helped with some concerns about the original paper, but some issues were not adequately addressed. I have left two primary concerns in my public review. To summarize those:

The difference in kinetics of proximal and distal locations is emphasized and quantified in the paper, but the quantification consists of a fit to the average responses. This does not give an idea of whether the difference observed is significant or not. Without an estimate of the error across measurements the difference in kinetic quoted is not interpretable.

Thank you for this feedback. Since the kinetics information is a minor part of the manuscript, we have followed the Editor’s advice to significantly tone down the comparison of kinetic fit parameters (completely removing the rise-time comparisons), in order to put more focus on the better-documented conclusions. We also note that we did establish statistical significance of the differences in fluorescence signal amplitudes.

Somewhat relatedly, the difference in amplitude and kinetics of the calcium signals measured with low and high affinity indicators is quite concerning. The authors added one sentence stating that the high affinity indicator might be saturated. This is not adequate. Should we distrust the measurements using the high affinity indicator? The differences between the results using the low and high affinity indicators is in some cases large - e.g. larger than the differences cited as a key result between distal and proximal locations. This issue needs to be dealt with directly in the paper.

Thank you for this feedback. Yes, the measurements from high-affinity indicators cannot report the Ca2+ as accurately as low-affinity indicators. However, the value of HA indicators is in their ability to detect lowamplitude signals that lower-affinity indicators may miss due to lower signal-to-noise resolution. We added a sentence on page 12 to further stress this point.

Related to the point about statistics, it is not clear how to related the horizontal lines in Figure 8 to the actual measurements. It is critical for the evaluation of the conclusions from that figure to understand what is plotted and what the error bars are on the plotted data.

We apologize for the earlier ambiguity in Fig. 8. In this figure, we first compare proximal (panel B) and distal (panel C) calcium signals across several RBCs, labeled RBC-a through RBC-d. Each RBC contains multiple ribbons, and for each cell, we present the average calcium signals from multiple ribbons using box plots in panels B and C. In these box plots, the horizontal lines represent the average calcium signal for each cell, while the size of the error bars reflects the variability in proximal and distal calcium signals among the ribbons within that RBC.

For example, RBC-a had five identifiable ribbons. In panels D–F, we use RBC-a to illustrate the variability in calcium signals across individual ribbons. Specifically, we distinguished proximal and distal calcium signals from five ribbons (ribbons 1–5) within RBC-a. When feasible, we acquired multiple x–t line scans at a single ribbon, shown now as individual data points, to assess variability in calcium signals recorded from the same ribbon.

The box plots in panels E and F display the average calcium signal (horizontal lines) for each ribbon, based on multiple recordings. These plots demonstrate considerable variability between ribbons of RBC-a. Importantly, the lack of or minimal error bars for repeated measurements at the same ribbon indicates that the proximal and distal calcium signals are consistent within a ribbon. These findings emphasize that the observed variability among ribbons and among cells reflects true biological heterogeneity in local calcium domains, rather than experimental noise.

Version 3

Abstract

Abstract

Rapid and high local calcium (Ca2+) signals are essential for triggering neurotransmitter release from presynaptic terminals. In specialized bipolar ribbon synapses of the retina, these local Ca2+ signals control multiple processes, including the priming, docking, and translocation of vesicles on the ribbon before exocytosis, endocytosis, and the replenishment of release-ready vesicles to the fusion sites for sustained neurotransmission. However, our knowledge about Ca2+ signals along the axis of the ribbon active zone is limited. Here, we used fast confocal quantitative dual-color ratiometric line-scan imaging of a fluorescently labeled ribbon binding peptide and Ca2+ indicators to monitor the spatial and temporal aspects of Ca2+ transients of individual ribbon active zones in zebrafish retinal rod bipolar cells (RBCs). We observed that a Ca2+ transient elicited a much greater fluorescence amplitude when the Ca2+ indicator was conjugated to a ribeye-binding peptide than when using a soluble Ca2+ indicator, and the estimated Ca2+ levels at the ribbon active zone exceeded 26 μM in response to a 10-millisecond stimulus, as measured by a ribbon-bound low-affinity Ca2+ indicator. Our quantitative modeling of Ca2+ diffusion and buffering is consistent with this estimate and provides a detailed view of the spatiotemporal [Ca2+] dynamics near the ribbon. Importantly, our data demonstrates that the local Ca2+ levels may vary between ribbons of different RBCs and within the same cells. The variation in local Ca2+ signals is found to correlate with ribbon size and active zone extent. Our serial electron microscopy results provide new information about the heterogeneity in ribbon size, shape, and area of the ribbon in contact with the plasma membrane.

Introduction

Sensory synapses in the retina rely on the proper function of a specialized organelle, the synaptic ribbon 1–5. Retinal bipolar cells serve as the major conduit for transmitting visual information across the vertebrate retina. Rod bipolar cells (RBCs) can release brief bursts of neurotransmitters to signal a change in contrast or sustain the continuous release of neurotransmitters in a graded manner to provide an analog read-out of luminance 6. To maintain this ability, the RBCs must exert dynamic control over neurotransmitter release rate and facilitate efficient recruitment of release-ready vesicles to fusion sites near the synaptic ribbon. It is established that the elevation of presynaptic Ca2+ in RBCs regulates both dynamic changes in the release rate and accelerates the rate of vesicle replacement 6–14. Of note, nanodomain calcium signals refer to highly localized, steep calcium concentration gradients that occur within tens to a couple of hundred nanometers of an open calcium channel. They are typically involved in triggering fast, synchronous neurotransmitter release, especially in synapses where Ca²⁺ sensors are closely coupled to the channels 15–18. Microdomain Ca2+ signals extend over a larger spatial range and arise when multiple calcium channels open in proximity, leading to overlapping calcium plumes. These domains result in slower, more diffuse Ca2+ elevations, often associated with modulatory functions or asynchronous release, where the Ca2+ sensor is located farther from the channel cluster 18–20. However, the spatiotemporal properties of the Ca2+ signals that control neurotransmitter release and the molecular entities that regulate the interplay between Ca2+ signals and synaptic vesicle dynamics to sustain kinetically distinct neurotransmitter release components remain poorly understood. Here, we begin to address this lack of knowledge by measuring local Ca2+ signals at positions along the synaptic ribbon at different distances from the active zone in retinal RBCs.

Resolving local Ca2+ signals is technically challenging as it requires information about the spatiotemporal properties of Ca2+ signals specific to ribbon sites. Previous studies of Ca2+ dynamics in goldfish and mammalian RBCs focused primarily on the terminal as a whole and used quantitative methods to examine bulk Ca2+ levels, which are significantly lower and slower than those occurring at the active zone 21–24. Our previous studies estimated that the Ca2+ signals at a single ribbon active zone in zebrafish RBCs likely exceed sub-micromolar concentration levels within 3 milliseconds after the opening of voltage-gated Ca2+ (Cav) channels located beneath the synaptic ribbon 25. However, these studies were based on estimates using soluble Ca2+ indicators to estimate the Ca2+ signals at the plasma membrane, which are free to diffuse and thus cause the spread of the signal. Measurements of Ca2+ signals along the axis of the ribbon have not been attempted, even though these signals likely control replenishment and priming of vesicles for slower phases of neurotransmitter release. A practical way to conceptualize the distinction between micro- and nanodomain Ca2+ signals is by approximating spatial scales: signals extending from around 0.5 μm and above may be considered microdomains, while those below 0.3-0.4 μm fall within the nanodomain range 18. To determine Ca2+ signals along the axis of zebrafish RBC ribbons, we establish a nanophysiology ratiometric approach that measures the spatial and temporal properties of Ca2+ transients by targeting the Ca2+ indicator to the ribbon via conjugation to a ribbon binding peptide (RBP)26. Resolving Ca2+ signals in the immediate vicinity of the active zone is not currently possible due to the finite spatiotemporal resolution of optical imaging. Thus, we employ a quantitative model using Ca2+ diffusion and buffering to resolve the RBC Ca2+ signals associated with the active zone and along the ribbon and provide a detailed description of the spatiotemporal Ca2+ dynamics across zebrafish RBC ribbons. Our nanophysiology approach for measuring local Ca2+ transients at a single ribbon with high spatiotemporal resolution provides the first evidence of heterogeneity of Ca2+ signals at zebrafish RBC ribbon synapses. Heterogeneity in local Ca²⁺ signals persisted in some ribbons even after the application of high concentrations of exogenous Ca²⁺ buffers, suggesting that the variability in the faster, smaller, and more spatially confined Ca²⁺ microdomains originates from Ca²⁺ influx through synaptic Ca²⁺ channel clusters at the base of the synaptic ribbon. Using serial block-face scanning electron microscopy (SBF-SEM), we found substantial variability in synaptic ribbon size, shape, and particularly the area of the ribbon adjacent to the plasma membrane in the zebrafish bipolar cell terminal. Thus, the observed heterogeneity of the Ca2+ microdomain is likely due to variability in the number of Ca2+ channels near each ribbon. Since local Ca2+ signals control kinetically distinct neurotransmitter release, heterogeneity in local Ca2+ signal may alter the rate of vesicle release and allow them to function independently, adding a new mechanism for increasing dynamic range of RBC.

Results

Ca2+ concentrations at single ribbon locations measured using high- and low-affinity diffusible indicators

Synaptic vesicles in RBCs are distributed among at least four distinct pools based on their fusion kinetics, which are assumed to reflect the average proximity of vesicles to Cav channels and the state of vesicle preparedness for Ca2+-triggered fusion 1,6,8,11–13,27–31 (see Supplementary Fig. 1). To visualize and measure Ca2+ signals near the RBC synaptic ribbon controlling such kinetically distinct components of neurotransmitter release, we established a quantitative nanophysiology approach shown in Fig.1 (see also Materials and Methods). In this approach, both the ribbon location and the spatiotemporal Ca2+ signal profile were simultaneously measured by dialyzing the zebrafish RBC terminals with both TAMRA (tetramethyl rhodamine)-labeled RIBEYE-binding peptide (RBP) 32 (to label synaptic ribbons; Fig. 1Aa) and a high-affinity Ca2+ indicator Cal-520® (Cal520HA, effective KD 795 nM; see Materials and Methods) using a whole-cell patch pipette placed directly at the cell terminal. A rapid x-t line scan was taken perpendicular to the plasma membrane across a ribbon, extending from the extracellular space to the cytoplasmic region beyond the ribbon (Fig. 1Ab). The spatial resolution was limited by the point spread function (PSF) of the microscope to approximately 270 nm (Supplementary Fig.2A). In our previous studies, line scans have been applied primarily to measure the temporal properties of Ca2+ signals 25. However, x-t raster plots obtained at a ribbon active zone labeled with RBP allow us to characterize the spatial localization of Ca2+ transients relative to the synaptic ribbon and plasma membrane (Fig.1C-D) in addition to characterizing its temporal aspects (Fig.1E-F). We previously used a similar approach for localizing and tracking single synaptic vesicles before and during fusion at a single ribbon 33, and to measure the kinetics of clearance of fused synaptic vesicle membrane in zebrafish RBC 34. We found that depolarization-evoked Ca2+ influx caused a rapid increase in Ca2+ signals at ribbon locations (Fig.1C, cyan, white horizontal arrowhead) and increased more slowly and less dramatically in the cytoplasm (Fig.1C, cyan, white vertical arrow) in zebrafish RBC. The Sigmoid-Gaussian function fitting of the x-t scans horizontal profile scans (see Materials and Methods) show that the local Ca2+ signals increased rapidly during stimuli (Fig.1D, cyan line) and then decreased immediately after the end of depolarization (Fig.1D, black line) approaching the spatial profile corresponding to the resting Ca2+ levels (Fig.1D, gray line). As expected, the centroid position of Ca2+ signals during depolarization (Fig.1D, cyan x0) is closer to the plasma membrane (Fig.1D, magenta x1/2) than the centroid position of RBP (Fig.1D magenta, x0), since Cav channels are located at the membrane 25,32,35–48.

Nanophysiology approach unveils spatiotemporal properties of local Ca2+ signaling in retinal RBC terminals.

A. Left panel. Single projection from a series of confocal optical sections through a zebrafish RBC synaptic terminal. A synaptic terminal was voltage-clamped using a whole-cell pipette with an internal solution containing TAMRA-RBP (magenta) to label synaptic ribbons (a, magenta). RBP fluorescence was concentrated at ribbons and also filled the entire terminal, allowing visualization of the terminal border. Experiments were carried out on ribbons that could be distinguished from adjacent ribbons (white asterisks). Scale, 5 μm. Right panel, Close-up view of a single synaptic ribbon. The outside of the cell is to the right, and x-t scan lines (dotted lines) were positioned perpendicular to the plasma membrane, extending from the intracellular side of the ribbon to the extracellular space. A rapid x-t line scan was taken at a ribbon location perpendicular to the plasma membrane across a ribbon (b) with sequential dual laser scanning performed at rates of 1.51 milliseconds per line per channel (3.02 milliseconds per line for both channels). The resulting x-t raster plots were used to measure the fluorescence intensity profiles of RBP (magenta) and the Ca2+ transient (cyan; Cal520HA) shown in panels C and E. Scale, 1.6 μm.

B. Voltage-clamp recording of a RBC terminal. Terminals were held at -65 mV and stepped to 0 mV (t0) for 10 ms (red) to evoke a brief Ca2+ current (black). A typical experiment began with a voltage command (VH= -65 mV), and a TTL pulse generated by the Patch Master software (t1) triggers image acquisition (t2). t0 is the time of depolarization.

C. Illustration of the approach used to obtain the spatial location of Ca2+ signals with respect to the ribbon. Example of an x-t raster plot that is oriented to illustrate the x-axis intensity profiles of RBP (magenta) and Ca2+ signal (cyan) fluorescence during a brief depolarization. Sequential dual laser scanning was performed at 1.51 milliseconds per line for one channel (3.02 milliseconds per line for both channels).

D. Fluorescence intensity profiles along the x-axis for RBP (magenta) and Cal520HA before stimuli (gray line), during 10 ms depolarization (cyan line), and after stimuli (black line) depolarization were obtained by averaging three pixels along the time-axis. RBP (magenta) and Ca2+ signals during (light cyan) were fit with a Sigmoid-Gaussian function 1. The centroid (x-axis position) of the RBP (magenta) and Ca2+ signals during (cyan) were taken as the peak of the Gaussian fit (x0). The parameter x1/2 (dotted magenta line) from the Sigmoid fit to the RBP fluorescence (magenta trace) was used to estimate the location of the plasma membrane.

E. The x-t raster plot shown was from the same recording as in panel C but re-oriented to demonstrate the t-axis intensity profiles of RBP (magenta) and Ca2+ transient (cyan).

F. Spatially averaged Cal520HA fluorescence as a function of time at ribbon proximal (light cyan), distal (dark cyan), and cytoplasmic (gray) locations from the single ribbon shown in D, upper panel.

To characterize the temporal Ca2+ profile at different distances from the plasma membrane, we analyzed the x-t line scans along the time-axis at three distinct distances (Fig. 1E-F) based on the spatial profile of the ribbon described in Fig.1C-D (see also Materials and Methods). We will refer to the corresponding three measurements as ribbon-proximal, ribbon-distal, and cytoplasmic (see Materials and Methods). We found that at the onset of the stimulus (Fig. 1F, black arrow), the fluorescence at the location near the ribbon proximal to the membrane (ribbon-proximal) rose more rapidly and to a higher level (Fig. 1F, light cyan trace) than that at the ribbon-distal and cytoplasmic locations (Fig. 1F, dark cyan and gray traces, respectively). To quantify the kinetics of ribbon-proximal and distal Ca²⁺ signals, we averaged multiple x-t scans acquired with Cal520HA under the same imaging conditions (Fig. 2A). The maximum values of trial-averaged ΔF/Frest (changes in the Cal520HA fluorescence during depolarization normalized to background level before depolarization) were significantly higher at ribbon-proximal locations (nearest the Cav channels) than those at ribbon-distal locations, with mean ± SEM of 1.9 ± 0.3 and 1.5 ± 0.25, respectively (Fig. 2A; p< 0.001, N=24). These findings suggest that the spatial resolution of our Ca2+ imaging using Cal520HA is sufficient to resolve differences in the smaller Ca2+ signals at ribbon proximal and distal locations. However, it should be noted that Cal520HA will be partially saturated at the Ca2+ levels expected in Ca2+ microdomains relevant for vesicle exocytosis 49, affecting both the amplitude and kinetics of the fluorescence signal. Therefore, we repeated the x-t line scan analysis with a lower-affinity soluble Ca2+ indicator Cal520LA (KD 90 μM; Fig. 2B) allowing us to better define the typical Ca2+ signals controlling distinct neurotransmitter release components corresponding to locations proximal and distal relative to the synaptic ribbon (Fig. 2B). We found that the ribbon-proximal signals detected with Cal520LA (Fig. 2B, light cyan) showed a sharper decay at the termination of the stimulus (Fig. 2B), when compared to ribbon-distal signals (Fig. 2B, dark cyan), as one would expect for nanodomain Ca2+ elevations 50–52. In twenty-one similar experiments, the peak ΔF/Frest at the membrane after 10-ms depolarization was significantly larger for proximal than distal signals (ΔF/Frest: 3.1 ± 0.4 and 1.9 ± 0.2 respectively P=0.001, N= 21). As expected, the ratio of proximal to distal signals measured with Cal520LA (1.6) was significantly higher than that measured with Cal520HA (1.3). The decay phase of all fluorescence transients was fit by a sum of two exponential functions, as described in Methods; for the Cal520HA recording (Fig. 2A) the two decay time constants and their relative magnitudes were similar for the ribbon-proximal and distal recordings, namely 11ms (60%) and 203ms (40%) for the ribbon-proximal recording, vs. 16ms (61%) and 208ms (39%) for the distal site. For the low affinity Cal520LA dye (Fig. 2B), the fluorescence decay components appeared faster due to faster Ca2+ unbinding from the dye, at 9.3ms (69%) and 175ms (31%) for the ribbon-proximal location, vs. 9ms (41%) and 135ms (59%) for the distal location. We note, however, that bi-exponential data fits are known to be highly sensitive to measurement duration and noise. Therefore, precise quantitative conclusions should not be drawn from the best-fit decay time constant values.

Kinetics of Ca2+ transients in response to brief stimuli recorded with freely diffusible indicators.

A. Spatially averaged Cal520HA fluorescence as a function of time at ribbon proximal (light cyan) and distal (dark cyan) locations from a single ribbon, as shown in Fig.1 (n=24 ribbons from 7 different RBCs). The corresponding maximum values of trial-averaged ΔF/Frest were significantly higher at ribbon-proximal locations than those at ribbon-distal locations, with mean ± SEM of 1.9 ± 0.3 and 1.5 ± 0.25, respectively (paired t-test, p< 0.001, N=24).

A inset. The temporal profile of events between 20-60 ms is shown in an expanded view for better visualization. Scale bars: vertical, 0.5 (ΔF/Frest); horizontal,10 ms.

B. Spatially averaged Cal520LA fluorescence as a function of time at ribbon proximal (light cyan) and distal (dark cyan) locations. Data points show the average intensity (± SEM) in each horizontal row of 5 pixels for three 10 ms depolarizations at distinct ribbon locations (see Materials and Methods and Fig.1). Fluorescence intensity is normalized with respect to the baseline fluorescence before stimulation, and averaged over all pixels (i.e., over space and time). The arrow indicates the onset of the 10-ms depolarizing stimulus. (n=21 ribbons from 4 different RBCs). The peak ΔF/Frest at the membrane after 10-ms depolarization was significantly larger for proximal than distal signals (paired t-test, ΔF/Frest: 3.1 ± 0.4 and 1.9 ± 0.2, respectively, P=0.001, N=21). The currents were not significantly different between the Cal520HA and Cal520LA conditions (unpaired t-test, Cal520HA average current = 45.1 ± 4.5 pA; Cal520LA average current = 51.4 ± 4.4 pA; p = 0.33).

B inset. Temporal profile of events between 20-60 ms was expanded for better visualization. Scale bars: vertical, 0.5 (ΔF/Frest); horizontal, 10 ms.

Ca2+ concentrations at single ribbon locations measured using ribbon-bound indicators

Although Ca²⁺-sensitive fluorescent chemical dyes have been used previously 7,26,35,42,48,49,51–63, the visualization of signals within smaller domains using freely diffusible Ca²⁺ reporters is limited by the resolution of light microscopy and the spread of the indicator by diffusion 64. Diffusible Ca2+ indicators report space-averaged Ca2+ concentrations, and their intracellular diffusion inherently broadens the spatial resolution of Ca2+ nanodomains. To partially overcome this problem, we targeted the Ca2+ indicators to the ribbon by fusing them to the RBP, as described previously 26. RBP-conjugated Ca2+ indicator s Cal520HA-RBP or Cal520LA-RBP were introduced to the RBC terminal together with fluorescently labeled RBP via whole-cell voltage clamp by placing the patch pipette directly at the terminal while imaging the terminal using laser scanning confocal microscopy. For two-color imaging, ribbons were labeled with TAMRA-RBP that did not interfere with Cal520-RBP fluorescence, and both channels were scanned sequentially to prevent possible bleed-through. Under these conditions, we used the spots detected by TAMRA-RBP to define the locus of the synaptic ribbon (Fig 3A&B, magenta). We also found punctate regions with both Cal520HA-RBP (Fig 3A, cyan) and Cal520LA-RBP (Fig 3B, cyan) at the same location, on a dimmer fluorescent background of the synaptic terminal, which correspond to the Ca2+ indicator-peptide complexes that are bound and not bound to the ribbon, respectively 26. The overall changes in ΔF/Frest were averaged over several trials for proximal and distal Ca2+ signals measured with Cal520HA-RBP (Fig. 3C, light cyan vs. dark cyan) and Cal520LA-RBP (Fig. 3D, light cyan vs. dark cyan). When compared to their distal location counterparts, the Ca2+ signals proximal to the membrane showed a sharper decay at the termination of stimuli (Figs. 3C and 3D, light vs. dark cyan trace), as expected when comparing nano- and microdomain Ca2+ profiles. Our data shows that the amplitude differences between ribbon-proximal and ribbon-distal Ca2+ signals were well resolvable using the ribbon-bound Cal520HA-RBP indicator (Fig. 3C, light vs. dark cyan traces: ΔF/Frest = 3 ± 0.4 vs. 1.9 ± 0.3, respectively, p=0.001) and Cal520LA-RBP indicator (Fig. 3D, light vs. dark cyan traces: ΔF/Frest = 5.5 ± 0.9 vs. 3.3 ± 0.8, respectively, p=0.003). The amplitudes of ribbon-proximal Ca2+ signals were higher when measured with Cal520LA-RBP than with Cal520LA-free (Fig. 3E, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray): ΔF/Frest = 5.5 ± 0.9, n=30 vs. 3 ± 0.4, n=19, p=0.04) but this was not the case for distal Ca2+ signals (Fig. 3F, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray): ΔF/Frest = 3.3 ± 0.8, n=30 vs. 1.9 ± 0.2, n=21, p=0.15). Notably, the Ca2+ signals at distal sites measured with Cal520LA-RBP reached their peak amplitude earlier (Fig. 3F inset, dark cyan) than those measured with Cal520LA-free (Fig. 3F inset, light gray). The two decay time constants and their relative magnitudes for the Cal520HA-RBP fluorescence (Fig. 3C) were 11ms (52.5%) and 182ms (47.5%) for the ribbon-proximal site, vs. 24ms (47%) and 192ms (53%) for the site distal to the ribbon. For the low affinity ribbon-bound indicator dye (Fig. 3D), the fluorescence decay components appeared faster, at 5.7ms (83%) and 129ms (17%) for the proximal site, vs. 7.5ms (76%) and 158ms (24%) for the distal location. Together, these findings suggest that conjugating the Cal520LA indicator to RBP provides a more accurate, promising approach for measuring the distinct local Ca2+ signals at ribbon-proximal vs. ribbon-distal locations.

Temporal properties of Ca2+ transients recorded with free and ribeye-bound Ca2+ indicators.

A-B. Confocal images of the isolated RBCs that were whole-cell voltage-clamped using an internal solution containing the TAMRA-RBP (magenta) and either (A) Cal520HA-RBP (cyan) or (B) Cal520LA-RBP (cyan). Note prominent spots in both TAMRA-RBP and ribeye-bound Ca2+ indicators (A) Cal520HA-RBP (cyan) or (B) Cal520LA-RBP, showing the location of the ribbon. Scale bars, 2 μm.

C-D. Spatially averaged fluorescence of (C) Cal520HA-RBP (n=19) or (D) Cal520LA-RBP (n=30) as a function of time at ribbon proximal (light cyan), and distal (dark cyan) locations. Data points show the average intensity (± SEM) in each horizontal row of five pixels for 10 ms depolarizations at distinct ribbon locations. Fluorescence intensity at the onset of the 10 ms depolarizing stimulus (arrow) was normalized to the baseline fluorescence before stimulation and averaged over all pixels (i.e., over space and time). The amplitude differences between ribbon-proximal and ribbon-distal Ca2+ signals were well-resolvable using the ribbon-bound Cal520HA-RBP indicator (Fig. 3C, light vs. dark cyan traces, paired t-test: ΔF/Frest = 3.0 ± 0.4 vs. 1.9 ± 0.3, respectively, p=0.001) and Cal520LA-RBP indicator (Fig. 3D, light vs. dark cyan traces, paired t-test: ΔF/Frest = 5.5 ± 0.9 vs. 3.3 ± 0.8, respectively, p=0.003). The current amplitudes were not significantly different between Cal520HA-RBP and Cal520LA-RBP readings (mean current amplitudes: Cal520HA-RBP = 49.8 ± 2.2 pA, Cal520LA-RBP = 49.9 ± 2.9 pA; p = 0.99).

C-D inset. Temporal profile of events between 10-50 ms were expanded for better visualization. Scale bars: vertical, 1 (ΔF/Frest, C inset) or 2 (ΔF/Frest, D inset); horizontal, 10 ms.

E-F. Average fluorescence intensity of (E) proximal and (F) distal Ca2+ signals obtained with Cal520LA-RBP (light cyan and dark cyan, respectively) and Cal520L-free (grey). The amplitudes of ribbon-proximal Ca2+ signals were higher when measured with Cal520LA-RBP than with Cal520LA-free (Fig. 3E, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray), unpaired t-test: ΔF/Frest = 5.5 ± 0.9, n=30 vs. 3.0 ± 0.4, n=19, p=0.04) but this was not the case for distal Ca2+ signals (Fig. 3F, Cal520LA-RBP (light cyan) vs. Cal520LA-free (gray), unpaired t-test: ΔF/Frest = 3.3 ± 0.8, n=30 vs. 1.9 ± 0.2, n=21, p=0.15).

E-F inset. Events between 10-50 ms were expanded for better visualization. Scale bars: vertical, 2 (ΔF/Frest, E inset) and 1 (ΔF/Frest, F inset); horizontal, 10 ms.

This method also enables ratiometric measurements with RBP-conjugated Ca2+ indicator by normalizing its fluorescence to that of RBP-peptides conjugated to Ca2+-insensitive fluorophores (TAMRA-RBP) to provide an estimate of Ca2+ concentration in RBC synaptic ribbons. We obtained the effective K1/2 (Keff) by measuring the Cal520HA-RBP/ TAMRA-RBP fluorescence ratio in buffered Ca2+ solutions and using the Grynkiewicz equation59. For Cal520HA-RBP, we found the Keff to be ∼795 nM, which is higher than the value of 320 nM reported by the manufacturer for Cal520HA. The differences between the in-cell measurements and the manufacturer’s values are likely to arise from differences in the cellular buffering capabilities, changes in dye properties due to interactions with the molecules inside the cell 42,65–67, and possible differences in the binding properties of the peptide-conjugated Ca2+ indicators when bound to synaptic ribbons. Since the in-cell approach most closely matched the experimental conditions, we used this value (Keff) for all further calculations.

We first measured the ribbon-proximal and ribbon-distal Ca2+ concentrations using Cal520HA-RBP with 0.2 mM EGTA in the patch pipette, as it allowed to resolve the gradient between the two signals (Fig. 1A). Under these conditions, we found a maximum ribbon-proximal Ca2+ concentration produced in response to a brief 10- ms pulse of 3.7 µM, and a maximum ribbon-distal Ca2+ concentration of 0.7 µM. These apparent Ca2+ concentration amplitudes are well below levels required to trigger exocytosis of the ultrafast releasable pool (UFRP) and readily releasable pool (RRP) 49 and, as discussed earlier, likely represent the lower bounds of the Ca2+ concentration at a single ribbon location due to local saturation of the high-affinity indicator and/or due to nanodomains that are smaller than the resolution attainable using light microscopy. It should also be noted that unlike freely diffusing Ca2+ indicators, RBP-conjugated indicators that slowly unbind from the ribbon are not readily replaced by free Ca2+ indicators, rendering them even more prone to saturation. Nevertheless, our data demonstrate that ribbon-bound indicators may better report local Ca2+ concentrations due to their localization, albeit still subject to the limitations of light microscopy.

To test the contribution of local saturation and to better estimate ribbon Ca2+ concentrations, we repeated our measurements using the low-affinity Ca2+ indicator Cal520LA conjugated to the RBP peptide. Because it was difficult to perform in-cell measurements to determine the Keff for Cal520LA-RBP given the large amounts of Ca2+ required to calibrate the Cal 520LA indicator (see Methods and Materials), we used the K1/2 value provided by the manufacturer for free Cal520LA (KD 90 µM). However, we expect that in-cell measurements of Keff, are likely to be different due to the cellular buffering properties reported previously for Keff, measurements of Oregon Green BAPTA-5N in inner hair cells 42.

Bipolar cells release neurotransmitters primarily from ribbon active zones, although some release also occurs at non-ribbon sites (referred to as NR, Fig. 4A) 68–70. To reveal the Ca2+ signaling at and away from the ribbon, we performed whole-cell patch clamping and x-t line scans at ribbon (Fig. 4Ba, R) and non-ribbon (Fig. 4Bb, NR) sites perpendicular to the plasma membrane using TAMRA-RBP and Cal520LA-RBP. A brief (10-ms) depolarizing voltage-clamp pulse evoked rapid high Ca2+ signals (Fig. 4Ba, cyan raster plot) at ribbon locations (Fig. 4Ba, magenta raster plot) but not at non-ribbon locations (Fig. 4Bb, cyan raster plot). The amplitude of Ca2+ signals elicited by brief depolarization and detected with Cal520LA-RBP at the ribbon-proximal site (Fig. 4C, light cyan traces) were significantly higher than those at ribbon-distal (Fig. 4C, dark cyan traces) and non-ribbon (Fig. 4C, blue traces) sites. The Ca2+ concentration gradient along the ribbon is summarized in Fig. 4D. We found that the average Ca2+ concentration at the proximal side of the ribbon (26.4 ± 3.1 µM, n=26) was significantly different from that at the ribbon-distal (15.6 ± 1.5 µM, n=26) and non-ribbon (10.4 ± 0.4 µM, n=15) sites, and that Ca2+ concentrations at ribbon-distal sites is higher than those at non-ribbon sites (Fig. 4D). These measurements display large heterogeneity across distinct ribbons and distinct cells, with the coefficient of variation of about 60% for [Ca2+] measurements at locations proximal to the ribbon, compared to a 10% CV for multiple depolarizations for the same ribbon. The Ca2+ signals at distal sites also reached their peak amplitude earlier (Fig. 4C inset, dark cyan arrowhead) than those at the non-ribbon sites (Fig. 4C inset, blue arrowhead). These findings are consistent with non-ribbon vesicle release being governed by cytoplasmic residual Ca2+ or Ca2+ influx via clustered Cav channels at ribbon sites rather than non-ribbon active zones with clusters of Cav channels and are also in agreement with previous reports regarding the temporal delay and sensitivity of non-ribbon exocytosis to EGTA 69,71.

Ca2+ signals at synaptic ribbon at different distances from the plasma membrane.

A. Ultrastructure of a zebrafish RBC with kinetically distinct vesicle pools, as described in Supplementary Fig. 1, UFRP (green vesicles), and RRP (yellow vesicles) are primarily released via ribbon sites (R) at the cytomatrix of the active zone (arrowhead). Recycling pool (RP, orange vesicles) in the cytoplasm (C), likely to be released via non-ribbon (NR) sites. Scale bar: 500 nm.

B. Representative x-t plots show the fluorescence intensity of Cal520LA-RBP (cyan) and TAMRA-RBP (magenta) as a function of distance (vertical axis) and time (horizontal axis) at (Ba) ribbon sites and (Bb) non-ribbon (NR) locations. The darker region at the upper edge of each plot is the extracellular space and the arrowheads show the timing of depolarization. Scale bars: vertical, 1.6 μm and horizontal, 75 ms.

C. Spatially averaged Cal520LA-RBP as a function of time at ribbon proximal (light cyan), distal (dark cyan), and non-ribbon locations (blue) (n=26 (proximal ribbons), 26 (distal ribbons), and 15 (non-ribbon) from 5∼8 different RBCs, respectively).

C inset. Temporal profile of events between 0-100 ms were expanded for better visualization. Scale bars: vertical, 0.2 (ΔF/Frest); horizontal, 20 ms.

D. Ca2+ measurements along the ribbon axis using the nanophysiology approach demonstrate that the proximal Ca2+ signals can go as high as 26.4 ± 3.1 µM (light cyan, N=26) and distal as 15.6 ± 1.5 µM (dark cyan, N=26), and non-ribbon 10.4 ± 0.4 µM (NR, blue, N=15), respectively, in response to 10 ms stimuli. Error bars show standard errors. All conditions were significantly different from each other as assessed by paired t-test when comparing proximal vs. Distal and unpaired t-test when comparing non-ribbon to proximal or distal (proximal vs. distal p = 0.004; proximal vs. non-ribbon p < 0.001; distal vs. non-ribbon p = 0.002). The currents were not significantly different between conditions (Mean current: 0.2 mM EGTA Cal520L-RBP proximal and distal = 51.8 ± 3.2 pA, 0.2 mM EGTA Cal520L-RBP non-ribbon = 47.3 ± 3.0 pA; p = 0.35).

Sensitivity of microdomain Ca2+ to exogenous buffers

Previous work has shown that the rate of vesicle replacement in RBCs is accelerated by elevated Ca2+ levels at sites along the ribbon, and that while millimolar levels of EGTA have little effect on the fast exocytosis component near Cav channels, they selectively block sustained exocytosis, likely by preventing Ca2+ from reaching the distal locations on the ribbon 11,72–75. These findings raise the possibility that Ca2+ has two sites of action, one near the Cav channels that trigger vesicle release and one further away that replenishes the supply of releasable vesicles. Burrone et al. (2002) proposed that endogenous Ca2+ buffers regulate the size of the RRP by limiting the spatial spread of Ca2+ ions and could suppress vesicle release at the periphery of the active zone in bipolar cells 7. However, the Ca2+ gradient at different locations along the ribbon controlling RRP release has never been examined. Thus, we used our higher resolution approach (Fig.1) to measure the Ca2+ gradient along the synaptic ribbon under buffering conditions that differentially modulate vesicle release and resupply. To determine the spatiotemporal properties of Ca2+ signals under these conditions, we performed rapid x-t line scans at a single ribbon location in the presence of ribbon-bound Cal520LA-RBP and under varying concentrations of exogenous buffer in the patch pipette solution, including (EGTA at 0.2 (Fig. 5A), 2 (Fig. 5B), and 10 mM (Fig. 5C) or BAPTA at 2 mM (Fig. 5D). In these experiments, we chose Cal520LA-RBP to study the varying contributions of exogenous buffer, as Cal530HA-RBP measurements are less accurate. However, it should be noted that for briefer depolarizations and lower amplitude depolarizations, where the signal for the Cal520LA-RBP is small, the Cal520HA-RBP could be more useful. We found that the ratio between proximal and distal Ca2+ signal amplitudes was similar in 0.2 and 2 mM EGTA (proximal-to-distal ratio ∼1.7, Supplementary Table 1) but was enhanced with 10 mM EGTA (proximal-to-distal ratio ∼1.9, Supplementary Table 1) and further enhanced with 2 mM BAPTA (proximal-to-distal ratio ∼4.5, Supplementary Table 1). These findings once again emphasize the reliable measurement of resolving ribbon-proximal vs. ribbon distal Ca2+ signals using our nano-physiology approach. Our experimental findings of the increase in the proximal-to-distal Ca2+ concentration ratio with increasing EGTA concentration are consistent with our simulation results (Fig. 7), although the corresponding ratios are greater in the simulation than in the experiment due to the large size of the microscope’s point-spread function.

Effect of exogenous Ca2+ buffers on spatiotemporal properties of Ca2+ microdomains in RBC terminal recorded with low-affinity ribbon-bound dye.

A-D. Average temporal fluorescence intensity (normalized to ΔF/Frest) of proximal (light cyan) and distal (dark cyan) Ca2+ signals with Cal520LA-RBP as a function of time at distinct ribbon locations with pipette solution containing (A) 0.2 mM EGTA, (B) 2 mM EGTA, (C) 10 mM EGTA, or (D) 2 mM BAPTA. Proximal measurements were significantly higher than distal measurements in all conditions as assessed using paired t-tests (0.2 mM EGTA: p = 0.0027, n = 29; 2 mM EGTA: p = 0.034, n = 21; 10 mM EGTA: p < 0.001, n = 42; 2 mM BAPTA: p = 0.0073, n = 21). The currents between conditions were not significantly different from each other (mean current amplitudes in 0.2 mM EGTA: 50.6 ± 3.0 pA, 2 mM EGTA: 49.7 ± 3.1 pA, 10 mM EGTA: 47.3 ± 3.1 pA, 2 mM BAPTA: 56.1 ± 2.6 pA; 0.2 mM EGTA vs 2 mM EGTA: p = 0.84, 0.2 mM EGTA vs 10 mM EGTA: p = 0.46, 0.2 mM EGTA vs 2 mM BAPTA: p = 0.19).

Inset. The temporal profiles of events between 0-50 ms were expanded for better visualization. Scale bars: vertical, 2 (ΔF/Frest; panels A and B) and 1 (ΔF/Frest; panels C and D); horizontal, 10 ms.

E. Average temporal fluorescence intensity (normalized to ΔF/Frest) of proximal Ca2+ signals measured with Cal520LA-RBP (cyan) and Cal520LA-free (gray) as a function of time with pipette solution containing 2 mM BAPTA. The corresponding maximum values of trial-averaged ΔF/Frest was significantly higher with 2mM Cal520L bound BAPTA than 2 mM Cal520L free BAPTA, with mean ± SEM of 3.6 ± 0.9 and 1.3 ± 0.2, respectively (paired t-test, p = 0.015; Cal520L-free: n = 20; Cal520L-bound: n = 21). The currents between conditions were not significantly different from each other (mean current amplitudes in 2 mM BAPTA Cal520L-RBP: 56.1 ± 2.6 pA; 2 mM BAPTA Cal520L-Free: 53.3 ± 2.2 pA; p = 0.42).

Though BAPTA significantly abolished the spread of Ca2+ signals, it is impressive that a substantial amount of ribbon-proximal Ca2+ concentration measured with ribbon-bound indicators was still present with 2 mM BAPTA. One reason for this observation could be that ribbon-bound Ca2+ indicators are likely to measure Ca2+ signals very close to Cav channels without diffusing away, which makes it impossible for BAPTA to intercept Ca2+ ions. If so, similar experiments conducted with free Ca2+ indicators should report a significantly lower proximal Ca2+ signal since the measurement by a diffusible dye would effectively spread the signal over a larger volume. Indeed, this is what we found. As shown in Fig. 5E, the apparent proximal Ca2+ signal in response to a 10-ms brief pulse measured with Cal520LA-free in the presence of 2 mM BAPTA was 0.9 ± 0.2 (n=9), 4-fold lower than proximal Ca2+ signals measured with Cal520LA-RBP (3.6 ± 1, n=20). The latter value is closer to the true measure of the Ca2+ concentration in the vicinity of the ribbon base. These findings emphasize that the reliable measurement of ribbon-proximal Ca2+ signals in the vicinity of Cav channels greatly benefits from the increased spatial resolution of our nano-physiology approach. We also measured the spatiotemporal properties of local Ca2+ signals using Cal520HA-RBP as we have done above for Cal520LA-RBP (Supplementary Fig. 3). We found that the ratio between proximal and distal Ca2+ signals was similar in 0.2 and 2 mM EGTA (proximal-to-distal ratio ∼1.5, Supplementary Table 2) but was enhanced with 10 mM EGTA (proximal-to-distal ratio ∼2.6, Supplementary Table 2) and further enhanced with 2 mM BAPTA (proximal-to-distal ratio ∼2.8, Supplementary Table 2). Cal520 acts as a buffer that binds Ca2+ and carries it away by diffusion. Thus, the higher affinity indicator Cal520HA will bind and shuttle away more Ca2+ than Cal520LA. If the Cal520HA indicator is saturated near the Cav channels, it may be underreporting the fast Ca2+ transients that occur in those locations. However, Cal520LA and Cal520HA show similar increase in the ribbon proximal-to-distal ratios with increasing concentrations of exogenous buffer.

Computational models of Ca2+ signals along the axis of the RBC synaptic ribbon

Since we could identify the position of the ribbon and we used RBP-fused Ca2+ indicator-RBP, we were able to measure and distinguish ribbon-proximal vs. ribbon-distal Ca2+ signals that drive the release of UFRP and RRP. However, several factors, including finite spatiotemporal resolution of optical imaging, spatial diffusion, dye saturation, and dye binding kinetics, may limit our ability to achieve optimal resolution. These limitations, however, can be overcome by the use of quantitative models. Thus, we developed a model based on our data with published information on Ca2+ diffusion and buffering to estimate more accurately the [Ca2+]I gradient along the ribbon at distinct distances from the plasma membrane.

Combining whole-terminal Ca2+ current measurements and our estimate for the number of synaptic ribbons per terminal allowed us to infer the magnitude of Ca2+ current per single ribbon, which we used in our model to determine the spatiotemporal [Ca2+] dynamics near the ribbon. Fig. 7 shows the results of the simulation of [Ca2+] at various distances from the ribbon during and after a depolarizing pulse of 10ms duration in the presence of different concentrations of exogenous and endogenous buffers, replicating the conditions used in our experiments. The geometry of the simulation domain box is shown in Fig. 6; it represents the fraction of total synaptic terminal volume per single ribbon (see Methods, Supplementary Table 3 and Supplementary Fig.4A.). Fig. 7 shows the time-dependence of [Ca2+] during and after the depolarization pulse at 5 specific locations both near and distant from the ribbon (left-hand panels in each subplot). These five spatial locations are marked by circles of different colors in the right-hand panel of each subplot, which shows [Ca2+] at the end of the Ca2+ current pulse in pseudo-color in the entire planar cross-section of the simulation domain cutting through the middle of the ribbon, as shown in Fig. 6. We assumed a highly simplified Cav channel arrangement into four clusters forming a square with a side length of 80nm. The closest of the five spatial locations (red curves and circles in Fig. 7) is X=20 nm away from the base of the ribbon center-line, Z=10 nm above the ribbon, about 45 nm away from the closest Cav channel cluster (see Fig. 6). We assumed that this location was within the Ca2+ microdomain that triggered the release of the UFRP. Vesicles were not included in the simulation since the total exclusion volume attributed to them represented only a small fraction of the total inter-terminal volume and, therefore, did not significantly impact [Ca2+] at the qualitative resolution level we are interested in. A couple of simulations with vesicles included were performed to confirm this statement, requiring much finer spatial resolution and longer computational time. We note also that [Ca2+] for short pulse durations considered here is relatively insensitive to our assumptions on the membrane Ca2+ extrusion mechanisms, listed in Supplementary Table 3.

Geometry of the computational model of intra-terminal Ca2+ dynamics.

Simulation domain is a box with dimensions (1.28 × 1.28 × 1.1) µm3, approximating the fraction of synaptic terminal volume per single ribbon. Ca2+ ions enter near the base of the ellipsoidal ribbon at four locations marked by black disks representing Ca2+ channels or their clusters. The semi-transparent gray coordinate plane Y=0 corresponds to the section used for the pseudo-color 2D [Ca2+] plots in Figure 7. Ca2+ is extruded on all surfaces of this domain, simulating combined clearance by pumps and exchangers on the plasmalemmal as well as internal endoplasmic reticulum and mitochondrial membranes.

Simulation of the effect of an endogenous immobile buffer of different concentrations on [Ca2+] dynamics in response to a 10ms pulse.

(A1-C2) In each sub-plot, the left-hand panels show the [Ca2+] time course in response to a 10 ms constant current pulse (total current of 1 pA) at five select locations marked by colored circles in the right panel. The right-hand panels show a pseudo-color plot of [Ca2+] in a 2D section of the 3D simulation volume in Fig. 6, at a fixed point in time corresponding to the end of the current pulse. Concentration values for each level curve are indicated in the color bar. Endogenous buffer is immobile, with a concentration of 200 µM in panels A1-C1 (resting buffering capacity 100 µM), vs. 1.44 mM in panels A2-C2 (resting buffering capacity 720). Exogenous buffer concentrations were 0.2 mM EGTA (panels A1, A2), 10 mM EGTA (panels B1, B2), 2 mM BAPTA (panels C1, C2).

Fig. 7A1-B2 reveals that [Ca2+] could rise above 50 µM within the microdomain near the base of the ribbon in the presence of 0.2 mM up to 10 mM EGTA. Given the size of the microscope point-spread function, this is in qualitative agreement with the 26 μM estimate of ribbon-proximal [Ca2+] that we recorded using the ribeye-bound low-affinity indicator. In addition, this concentration level is reached very soon after the channel opening event due to the rapid formation of the Ca2+ microdomain 76,77, and therefore, this estimate is expected to hold for shorter pulses as well. However, [Ca2+] decayed rapidly with distance, with the rate of decay significantly increasing when EGTA concentration was increased to 10mM (Fig. 7B1, B2). Note that a 7-fold change in the capacity of the immobile endogenous Ca2+ buffer (cf. Fig. 7A1 vs. 7A2) had a relatively modest effect on the [Ca2+] level up to about 150 nm distance from the ribbon base. This effect of increasing endogenous buffer concentration was further reduced in 10 mM EGTA (cf. Fig. 7B1 vs. 7B2), due to strong competition of endogenous buffer with large concentrations of EGTA. This agrees with the expectation that immobile buffers primarily slow down Ca2+ signals but do not affect the rapidly forming quasi-steady-state Ca2+ microdomains in the immediate vicinity of the channel 76–78. Finally, Fig. 7C1-C2 shows that 2 mM BAPTA had a much greater effect on localizing the Ca2+ signal to the immediate vicinity of the ribbon base.

Since a native unpatched cell may well contain mobile rather than immobile buffers, it was interesting to examine the effect of increasing the concentration of mobile endogenous buffer on Ca2+ dynamics in the ribbon vicinity, as compared to the corresponding effect of immobile buffer. Supplemental Figure 4B-C shows that a 7-fold increase in mobile buffer concentration reduces the microdomain Ca2+ by a factor of 2 and greatly reduces the size of the microdomain. This contrasts with the effect of increasing the concentration of the immobile buffer, which has a much more subtle effect (cf. Fig. 7A1, A2). We note that 0.2 mM EGTA was absent in the simulation with mobile buffer shown in Supplemental Figure 4B-C; in general, the effect of modest concentrations of EGTA is expected to be negligible due to its slow Ca2+ binding speed compared to the endogenous buffer.

Variability of local Ca2+ signals across the RBC synaptic ribbons

Bipolar cells release neurotransmitters primarily from ribbon active zones 1,6,8,11–13,27–31. However, the factors that shape the synaptic ribbon microdomains at retinal ribbon synapses have not been examined. We wondered whether all of the 30∼50 synaptic ribbons at an RBC terminal release glutamate in a similar fashion or whether there is some heterogeneity in glutamate release from different ribbon active zones. In particular, we asked what underlying mechanisms could differentiate the Ca2+ signals between ribbons of the same RBC terminal and across different RBCs. We first found evidence for such variability in local Ca2+ signals at single-ribbon locations of different cells using the freely diffusible high-affinity indicator Cal520HA, observing high variability in the Ca2+ transient amplitudes, even for RBCs with similar depolarization-evoked Ca2+ current amplitudes (cf. Supplementary Figs. 5A vs. 5B). However, we did not observe such variability when we obtained multiple line scans across the same ribbon (Supplementary Figs. 5A vs. 5B, gray traces). We wondered whether the observed variability between cells could be due to distinct subtypes of zebrafish RBCs 79, which might have different subtypes or numbers of Cav channels near the ribbon. Recent work in zebrafish retina identified two distinct RBC subtypes: RBC1 and RBC2 79. The wiring pattern of amacrine cells postsynaptic to RBC1 closely resembles the circuitry of mammalian RBCs, whereas RBC2 forms distinct pathways. Furthermore, RBC1 specifically labels for the known marker of mammalian RBCs, PKC-α 79. Due to these similarities, RBC1 in zebrafish is classified as analogous to mammalian RBCs. Moreover, RBC1s have expected morphological characteristics, in particular, the shape and size of soma and the presence of a single synaptic terminal 80. Immunolabeling of isolated RBC preparation used in these experiments showed primarily intact RBCs, which are PKC-α-specific RBC1 (Supplementary Fig. 2B). For simplicity, and because this study focuses exclusively on RBC1, we will refer to RBC1 as ‘RBC’ throughout this work. Thus, we attribute the variability in the RBC Ca2+ transients to the variability between the Ca2+ microdomains within the same cell type.

To identify mechanisms that contribute to the heterogeneity in the local Ca2+ signals we have reported in this study, we began by asking whether differences in local Ca2+ buffering could account for this heterogeneity. To examine the role of Ca2+ buffering in the variability of local Ca2+ signals across different RBC and different ribbons (Figs. 8A and 8D, respectively), we compared proximal (Figs. 8B and 8E) and distal (Figs. 8C and 8F) Ca2+ signals in 10 mM EGTA by averaging the Ca2+ recordings from all the ribbons of a given RBC (Supplementary Figs. 6 and 7). Here, we first compare proximal (panel B) and distal (panel C) calcium signals across several RBCs, labeled RBC-a through RBC-d. Each RBC contains multiple ribbons, and for each cell, we present the average calcium signals from multiple ribbons using box plots in panels B and C. In these box plots, the horizontal lines represent the average calcium signal for each cell, while the size of the error bars reflects the variability in proximal and distal calcium signals among the ribbons within that RBC. For example, RBC-a had five identifiable ribbons. When the Ca2+ signals were restricted to the ribbon sites with 10 mM EGTA in the pipette solution (Supplementary Tables 1 and 2), the amplitude of the ribbon-proximal and ribbon-distal Ca2+ signals averaged over all ribbons of a given cell exhibited significant variation across different RBCs (Figs. 8B, 8C, and Supplementary Fig. 6). In panels D–F (Fig. 8), we use RBC-a to illustrate the variability in Ca2+ signals across individual ribbons. Specifically, we distinguished proximal and distal Ca2+ signals from five ribbons (ribbons 1–5) within RBC-a. We next compared variability in Ca2+ signals in the presence of 10 mM EGTA at individual ribbons within the same cell (Fig. 8D and Supplementary Fig. 7) at proximal (Fig. 8E) and distal (Fig. 8F) locations. The box plots in panels E and F (Fig. 8) display the average Ca2+ signal (horizontal lines) for each ribbon, based on multiple recordings. For the cell described in Figs. 8D-8F, the proximal Ca2+ signals were significantly different across all ribbons examined, and there were considerable differences in distal Ca2+ signals between ribbons, with the exception of ribbons numbered 2 and 5. Importantly, the lack of or minimal error bars for repeated measurements at the same ribbon indicates that the proximal and distal calcium signals are consistent within a ribbon. These findings emphasize that the observed variability among ribbons and among cells reflects true biological heterogeneity in local calcium domains, rather than experimental noise. The heterogeneity in proximal and distal Ca2+ signals at distinct ribbons within the same cell may result from different underlying mechanisms, for example, heterogeneity in Cav expression or subtype and Ca2+ handling mechanisms. These findings suggest that exogenous Ca2+ buffering has a negligible effect on experimentally observed heterogeneity and variability of the proximal Ca2+ signals and that local Ca2+ signals at RBC ribbons are dominated by Ca2+ in regions close to the ribbon base where Cav channels are located.

Heterogeneity in the spatiotemporal properties of Ca2+ microdomains in RBC terminal.

A. Top panel. Cartoon of two representative RBCs (cell A and cell B), each containing differently colored ribbons. Bottom panel. Ribbon-to-ribbon variability was measured by recording local Ca2+ signals near different ribbons (yellow and purple traces) for each RBC. If an RBC had multiple readings for a single ribbon, averages were obtained for comparisons as described in Supplementary Fig. 6. Bottom panel inset: sample Ca2+ currents for the cells from which the Ca2+ signal sample traces mentioned above were obtained (purple and yellow traces). Currents were similar across the different cells. Vertical scale = 80 pA, horizontal scale = 5 ms.

B-C. Variability of Ca2+ signals in different ribbons across different cells in the presence of 10 mM EGTA, recorded using Cal520LA-RBP. Proximal Ca2+ amplitude values were significantly different between cells A and B (p = 0.022), cells A and D (p = 0.023), and cells B and D (p = 0.005) but similar between A and C (n = 4 cells, 4 fish), as assessed by unpaired t-tests. (C) In distal locations, ribbon amplitude values were significantly different between cells A and B (p = 0.031), but similar across all other cell comparisons (n = 4 cells, 4 fish), as assessed by unpaired t-tests. The currents were not significantly different across the different cells, as assessed by unpaired t-tests (mean currents in RBC a: 47.7 ± 2.9 pA, RBC b: 43.7 ± 4.0 pA, RBC c: 43.4 ± 2.0 pA, RBC d: 40.9 ± 2.4 pA; RBC a vs. RBC b: p = 0.45, RBC a vs. RBC c: p = 0.22, RBC a vs. RBC d: p = 0.14, RBC b vs. RBC c: p = 0.93, RBC b vs. RBC d: p = 0.54, RBC c vs. RBC d: p = 0.52).

D. Top panel. Illustration of a RBC containing three ribbons (numbered 1-3). Bottom panel. Ca2+ signal measurements from three distinct ribbons (black, gray, and blue traces) were compared to determine the ribbon-to-ribbon variability within each RBC, as described in Supplementary Fig. 7. Bottom panel inset: sample Ca2+ currents for the cells from which the Ca2+ signal sample traces mentioned above were obtained (black, gray, and blue traces). Currents were similar across the different cells. Vertical scale = 80 pA, horizontal scale = 5 ms.

E-F. Box plot illustrating [Ca2+] across various ribbons of an individual RBC, which is shown as RBC a. Ribbon variability within individual cells was measured with 10 mM EGTA using Cal520LA-RBP at (E) proximal and (F) distal locations. (E) Proximal Ca2+ amplitude values were significantly different among all ribbons (paired t-test, p < 0.001) (n = 5 ribbons, 1 RBCs, 1 fish). (F) Distal Ca2+ amplitudes were significantly different among all ribbon comparisons (paired t-test, p < 0.001) except for ribbons 2 and 5 (n = 5 ribbons, 1 RBCs, 1 fish). Similar analyses were conducted in two more cells and found similar observations (data not shown). The currents were similar across all ribbons since these were readings from the same cell. Given that some ribbons only have one reading, it is not possible to conduct a paired t-test to statistically compare them; however, the average current ± standard error for the cell shown was 47.7 ± 2.9 pA.

The ultrastructure of the zebrafish RBC terminal reveals diversity in the size of the synaptic ribbons across the terminal

In hair cells, Ca2+ microdomain signaling varies with ribbon size, reflecting larger patches of Cav channels aligned with larger ribbons 57. However, the number, size, and shape of ribbon active zones per terminal have not been established in our experimental system, the synaptic terminals of zebrafish RBCs. To provide quantitative measurement of ribbon microdomains, we examined the ultrastructure of zebrafish RBC synaptic ribbons by serial block face scanning electron microscopy (SBF-SEM) and reconstructed three RBCs from serial section electron micrograph (Fig. 9A). We analyzed the bipolar cell terminals that are closest to the ganglion cell layer with a morphology similar to mammalian RBCs and zebrafish RBC1s 79. Our reconstruction of three RBCs revealed 30-41 ribbons within RBC terminals (Fig. 9A), similar to what was previously reported in goldfish giant ON-type mixed RBCs (range 45-60 81).

Distribution of measured distance between synaptic ribbons in three zebrafish RBCs.

A. Reconstruction of three RBC terminals closest to the ganglion cell layer resembling the shape and size of the mammalian RBC1. The ribbons are shown in magenta. Note that the total number of ribbons includes the “floating” ribbons detached from the plasma membrane. RBC II contained 7 floating ribbons. Scale bar, 5 µm. RBCs, rod bipolar cells.

B. Overview of the distance measurements. Top. 3D rendering of a single RBC terminal is shown in light green with ribbons in magenta. Black lines show an example of the different SBF-SEM layers that are cut to obtain individual images. Bottom. Four sample SBF-SEM layers are shown with two example ribbons (magenta) located near each other but in different layers. If two ribbons were on the same layer, their linear distance was taken, whereas if they were located in different layers, their linear distance and height difference were used to calculate their actual distance using the Pythagorean Theorem. Each SBF-SEM layer has a thickness of 50 nm. A 3D volume movie of the synaptic terminal of a zebrafish bipolar neuron synaptic ribbon distribution and measurement is provided in Video 1.

C. Histograms showing the ribbon distance for the three RBCs. x0 shows the mean of the distribution.

D. Histogram of the distances between each ribbon and its five nearest ribbons for all ribbons contained in the three RBCs. x0 shows the mean of the distribution. Please note that the y-axis for B and C have different sizes, given that they represent the number of occurrences of each event.

We examined the distribution of these 30–41 synaptic ribbons within the zebrafish RBC terminals. The distribution of RBC ribbons as estimated by the distance between ribbon sites (Fig. 9B and Video 1; see Materials and Methods) revealed a wide distribution in synaptic ribbon distance to its nearest neighbor across the three reconstructed RBCs (means of 2.48 ± 0.12 µm, 1.84 ± 0.06 µm and 1.98 ± 0.08 µm; Fig. 9C). The distance between RBC synaptic ribbons ranged from 141 nm to 6 µm, with a mean of 2 ± 0.05 µm (Fig. 9D; across all three reconstructed RBCs). Of the 510 comparisons of zebrafish RBC synaptic ribbons, only 4 ribbon pairs (0.8%) were separated by distances smaller than our confocal microscope resolution limit of 270 nm.

To reveal whether heterogeneity in synaptic Ca2+ signals correlates with active zone size, we compared the shape and size of the ribbon active zones across the three reconstructed RBCs. Our results show significant variations in the shape and size of zebrafish RBC synaptic ribbons in RBCs and associated active zones (Fig. 10). On average, individual ribbons spanned 2–5 consecutive sections, with some located within the axons or closer to the terminal. The SBF-SEM images and 3D projections of ribbon structures (colored in magenta) revealed considerable variability in shape and size across all dimensions, as illustrated in Figs. 10A–C. The number of Cav channels per active zone in hair cells from chicken, frog, and turtle varies with the size of the synaptic ribbon 82. We thus measured the area of the ribbon facing the plasma membrane where the Cav channels are known to be located to estimate the number of Cav channels per active zone and to determine whether variation in active zone size might plausibly contribute to the heterogeneity in Ca2+ signaling. The EM images and 3D projection of the plasma membrane (yellow) and the area of the ribbon facing the plasma membrane (cyan), representing the active zone, are shown in Figs. 10A-C. We observed large variability in the area of the ribbon facing the plasma membrane or active zone within the RBCs, with substantial variability in their average distribution across the three RBCs (Figs. 10C, cyan, 10E, Video 2 and Supplementary Fig. 9), suggesting that the number of Cav channels per RBC active zone could plausibly be heterogeneous.

Serial block-face scanning electron microscopy analysis reveals heterogenous RBC ribbon shape, size and area of the ribbon facing the plasma membrane.

A-C. EM images of zebrafish RBC ribbon structures (A) and their respective 3D reconstructions to illustrate different shapes and sizes of synaptic ribbons (A, B, and C, magenta), plasma membrane (A and B, yellow), and the area of the ribbon facing the plasma membrane (A, B, and C, cyan). Each of the 5 rows of images illustrates one ribbon synapse from a zebrafish retinal RBC. A 3D reconstruction of the RBC synaptic terminal and ribbon from SBF-SEM stacks is provided in Video 2.

D. 3D reconstruction of the RBC terminal closest to the ganglion cell layer resembles the shape and the size of the mammalian RBC1 2. The ribbons are colored magenta.

E. Summary of the active zone size, the area of the ribbon associated with the plasma membrane measured in serial sections across the three RBCs from Fig 9A. The individual distribution of the three RBCs active zone sizes is provided in Supplementary Fig. 8. The z-size of the SBF-SEM sections is 50 nm, and each ribbon spans 2-5 consecutive sections. The solid cyan circles in the violin plots show individual synaptic ribbon measurements from three RBCs, with the average measurements shown in the solid black circle.

The size of the active zone and maximal Ca2+ influx correlate with the size of the synaptic ribbon

To reveal whether larger ribbons display stronger maximal Ca2+ influx, we measured Ca2+ signals in response to a series of 200-ms depolarizations. We first imaged TAMRA-RBP and Cal520LA-RBP in sequential scans to obtain ribbon location and resting Ca2+ levels. To maximize the capture of Ca2+ signals during brief stimuli, we image the Cal520LA-RBP channel, followed by TAMRA-RBP and Cal520LA-RBP sequential scans to confirm the ribbon locations. We found that the maximum amplitude of depolarization-evoked Ca2+ signals increased with an increase in ribeye fluorescence (Fig. 11A & B and Video 3; r = 0.51, N = 121 ribbons, p< 0.001), consistent with findings reported in cochlear inner hair cells 57,62. Since ribeye fluorescence correlates with the number of ribeye molecules per ribbon, and, therefore, ribbon size 32,57,62,83, our findings suggest that larger ribbons display stronger Ca2+ signals as active zone size scales with ribbon size 62. Given the diverse shapes of RBC synaptic ribbons (Fig. 10A & B, magenta), we measured the longest length and width of ribbons from EM images and compared this with the active zone size. We found that RBC ribbon length and width have a moderate positive correlation (Fig. 11C, r = 0.47, N = 102 ribbons, p< 0.001), and both dimensions have a moderate positive correlation with active zone size, albeit the ribbon width (Fig. 11D, r = 0.52, N = 102 ribbons, p< 0.001) has a stronger correlation with active zone size than the ribbon length (Fig. 11D, r = 0.32, N = 102 ribbons, p< 0.001). These results suggest that heterogeneity in synaptic Ca2+ signals correlates with ribbon dimensions and that active zone and the number of Cav channels 57,62 scale with ribbon size.

Heterogeneity of Ca2+ microdomains in RBC terminals.

Larger ribbons have stronger maximal Ca2+ influx and larger CAZ. A. Images show voltage-clamped RBC filled with a solution containing TAMRA-RBP to label ribbons before depolarization (left, magenta) and Cal520LA-RBP to measure the amplitude of Ca2+ influx (middle, cyan) during depolarization and superimposed of the two (right) to compare the size or ribbon vs. maximal Ca2+ influx. A time-lapse movie of the synaptic terminal of a zebrafish bipolar neuron during Ca2+ influx is provided in Video 3. B. Scatter plot of maximal Ca2+ influx (F/Frest) vs. TAMRA-RBP. Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. C. Scatter plot of maximal synaptic ribbon length vs. maximal synaptic ribbon width as estimated from SBF-SEM images. Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. D. Scatter plot of area of AZ vs. ribbon width (filled black circles) and of area of AZ vs. ribbon length (filled grey circles). Dashed lines are linear regressions, and r is Pearson’s correlation coefficient. AZ, Active Zone.

We note that a large correlation between the inferred ribbon size and active zone size would be expected if the size heterogeneity is partially due to the differences in the degree of labeling by the two ribbon-binding peptides. Therefore, the observed significant but moderate degree of correlation (r=0.32 or r = 0.52) between the active zone area and the ribbon length or width provides an additional indication of large heterogeneity in the calcium influx between ribbons of similar size, increasing further the heterogeneity in neurotransmitter release by different ribbons.

Discussion

Ca²⁺ signaling plays a central role in regulating neurotransmitter release throughout the retina, but the spatial and temporal organization of these signals differs markedly between the outer and inner retinal circuits. In the outer retina, photoreceptors rely on graded, sustained calcium influx through L-type Ca²⁺ channels at ribbon synapses to drive tonic glutamate release. These signals are relatively well characterized: the structure– function relationships between ribbon geometry, Ca²⁺ channel distribution, and synaptic vesicle release have been extensively studied, and recent high-resolution studies 84 have revealed nanometer-scale alignment between Cav channels and release sites. Postsynaptically, the identity of the glutamate receptor (e.g., mGluR6, KAR, AMPAR) and synaptic cleft geometry further tune the flow of visual information. In contrast, calcium dynamics in the inner retina, especially at bipolar cell terminals, are less well understood. The diversity of bipolar cell types, their complex patterns of ribbon organization, and the presence of both graded and spiking output modes add to this complexity. Moreover, presynaptic Ca²⁺ signals in the inner retina are likely to establish distinct micro- or nanodomain mechanisms depending on ribbon structure, local channel density, and the spatial coupling between Ca²⁺ influx and release sensors. Despite numerous studies that have focused in establishing retinal bipolar cells Ca2+ signaling controlling NTR 9,11,12,48,49,85–87, there remains significant uncertainty about how presynaptic Ca²⁺ signals are organized across individual ribbon synapses, especially in small or non-classical synapses where standard imaging lacks the resolution to resolve sub-ribbon-scale features. Our study addresses this gap by applying a new immobile Ca²⁺-ribbon indicator, enabling high-resolution tracking of Ca²⁺ domains at the level of individual ribbons. This approach offers a window into nanodomain calcium signaling, allowing us to directly link structural heterogeneity to functional diversity in vesicle release. By capturing the fine-scale dynamics that govern synaptic output in the inner retina, this approach opens the door to a more complete understanding of how visual signals are shaped at their earliest synaptic relays.

A deep look at the active zone Ca2+ microdomain in RBC terminals

In this study, we developed a quantitative nanophysiology approach using ribeye-bound Ca2+ indicators and fluorescently labeled RBP, combined with ratiometric dual-color imaging, to measure changes in Ca2+ concentration at and along the ribbon axis. By targeting Ca2+ indicators to the ribbon combined with fluorescently labeled RBP, we determined the Ca2+ concentration along the ribbon at different distances from active zone and found that near the ribbon base the Ca2+ concentration could increase to an average of 26 μM upon depolarization. Given the localization of our indicator on the ribbon and minimal sensitivity to exogenous buffers, we believe that our ribbon-proximal signal represents Ca2+ concentrations on the ribbon in locations very near release sites, likely some of those that contribute to vesicle resupply. Given that typical resolution for Ca2+ imaging—whether using confocal or even STED microscopy—is limited to ∼100–300 nm, the novelty of this study lies in the ability of the new immobile Ca²⁺-ribbon indicator to potentially capture signals on even finer spatial scales, approaching the realm of nanophysiology. Using 2D STED microscopy combined with fluorescence lifetime imaging, Moser’s lab achieved ∼50 nm lateral resolution, enabling precise visualization of Ca²⁺ channel clusters at single active zones in mouse cochlear inner hair cells 42. Notably, the [Ca2+] i levels obtained from our higher resolution approach using confocal microscopy are similar to those that were reported for measurements in hair cells using Stimulated Emission Depletion (STED) lifetime measurements, emphasizing that our immobile ribbon-bound Ca2+ indicator similarly enables nanoscopic (∼tens of nanometers) resolution of localized Ca²⁺ signals—complementing STED-based approaches like Neef et al.’s. 42. Previous work on RBCs found slowing of recovery from paired-pulse depression by EGTA, suggesting that Ca2+ levels at distal sites on the ribbon might be important for the recruitment of new vesicles following depletion of the RRP and UFRP 11,30. Here, we measured directly the effect of EGTA on the spread of Ca2+ signals and found that the Ca2+ signal at the distal half of the ribbon was two-fold lower than that at the location proximal to the membrane under conditions of low Ca2+ buffering. It should be noted, however, that diffraction of light during the optical imaging procedure is likely to underestimate spatiotemporal [Ca2+] along the ribbon. Therefore, we used our measurement of the Ca2+ current per ribbon to develop a detailed computational model that allowed us to resolve the expected profile of Ca2+ microdomains governing UFRP fusion. The model results shown in Fig. 7 reveal, on a finer spatial scale, the distribution of Ca2+ around a ribbon in the presence of various quantities of immobile endogenous buffer and added exogenous buffers. Our finding of high proximal [Ca2+] levels on the order of 26 µM recorded using a ribeye-bound low-affinity indicator dye is in qualitative agreement with the simulated high concentration of Ca2+ in the microdomain at the base of the ribbon. Simulations also show that, as expected, the peak Ca2+ microdomain amplitude is relatively insensitive to the concentration of endogenous immobile buffer and added

EGTA, but its decay with distance is accelerated by increasing the concentrations of exogenous and endogenous buffers. In contrast, the addition of 2 mM BAPTA causes a profound reduction of the Ca2+ microdomain amplitude and localizes the signal to a small volume around the ribbon. We also compared the impact of mobile vs immobile endogenous buffers on the Ca2+ distribution, showing the much greater effect of mobile buffers on the microdomain amplitude and distance dependence. Although these results are generally expected 76–78, the simulation results estimate the [Ca2+] levels in the vicinity of the ribbon with greater spatial and temporal resolution. As expected, the ratios of Ca2+ concentration values at the base of the ribbon vs. further away from the ribbon are much greater in the model simulation than our experimentally determined proximal-to-distal Ca2+ ratio values due to the large size of the microscope PSF, which precludes precise localization of the corresponding Ca2+ signals. Further, simulations show the effect of exogenous buffers with greater spatial resolution. However, the simulations fully agree with the overall increase of the proximal-to-distal Ca2+ concentration ratios by the application of exogenous Ca2+ buffers, especially BAPTA. Despite the technical challenges posed by the limited spatiotemporal resolution of optical imaging, the spatial resolution attained with the nanophysiology approach developed in our study is sufficient to differentiate Ca²⁺ signals associated with distal sites on the ribbon from those localized near the plasma membrane. Conceivably, the nanophysiology approaches established here could generally be applied to other classes of ribbon-containing cells, such as rods, cones and hair cells.

Recent ultrastructural and functional studies have provided new insights into the nanoscale organization of calcium channels at ribbon synapses. In rod photoreceptors, Cav1.4 channels have been shown to localize not only at the base but also along both sides of the synaptic ribbon, in close proximity—within nanometers—to other active zone (AZ) proteins 84. This spatial arrangement supports the idea that calcium influx is tightly restricted to subregions of the ribbon, enabling compartmentalized control of synaptic release. These findings are consistent with our own imaging results, which demonstrate localized, submicron Ca²⁺ signals that vary systematically with ribbon geometry. Together, these studies highlight a common theme: that nanodomain Ca²⁺ signaling at ribbon synapses is not only possible but structurally supported by the precise spatial alignment of Ca²⁺ channels and release machinery.

Heterogeneity of Ca2+ microdomains across RBCs terminals

Our quantitative nanophysiology approach provides evidence for the heterogeneity in synaptic Ca2+ signals across zebrafish RBCs, suggesting variability in the Ca2+ microdomains as has been previously reported for hair cells 57,62 and cone photoreceptors. Retinal bipolar neurons contain many ribbon active zones. This ranges from 45-60 in adult goldfish, as quantified by EM 81, or 25-45 in RBCs of adult living zebrafish, as estimated from laser scanning confocal microscopy 88. Our analyses of ribbon number across zebrafish RBCs through SBF-SEM arrived at an estimate of 31-40. The differences in the number of ribbons as determined through confocal microscopy compared to SBF-SEM could be explained by the resolution limit encountered by the confocal image analyses. Our analyses of zebrafish RBC terminal SBF-SEM images and 3D projections provide the first evidence of diversity in synaptic ribbon shape and sizes across zebrafish RBC terminals. Further, our quantitative measurements of SBF-SEM images reveal that synaptic ribbon width and length correlate with the active zone size, consistent with previous findings in hair cell ribbon synapses 82,83 and cone photoreceptor ribbon synapses58. Our observation that maximal active zone Ca2+ influx correlates with ribeye fluorescence in live imaging is consistent with findings previously reported in hair cells 62. Thus, we expect larger active zones, with more Ca2+ influx, to have more Cav channel expression, as demonstrated for hair cell ribbon synapses 62. Heterogeneity among ribbon active zones has been previously reported in hair cells, where synaptic Ca2+ microdomains varied substantially in amplitude and voltage-dependence within individual inner hair cells 57,62. Studies in aquatic tiger salamander cone photoreceptors revealed that the amplitude and midpoint activation voltage of Ca²⁺ signals varied across individual ribbons within the same cone. Additionally, local Ca²⁺ signal dynamics at cone photoreceptor ribbons were found to be independently regulated 58. However, in comparing our findings with studies of ribbon size heterogeneity in hair cells, we note that we do not observe any systematic variation of ribbon size with larger-scale morphological features like the cell position, as compared to the reported interesting tonotopic variation of ribbon size with position along the cochlea 57,62,

The faster, smaller, and more spatially confined Ca2+ signals that are insensitive to the application of high concentrations of exogenous Ca2+ buffers, referred to here as ribbon proximal Ca2+ signals, could be due to Ca2+ influx through Cav channel clusters beneath the synaptic ribbon 25,32,35–48. However, the variability in Ca2+ signals at the distal ribbon, away from the plasma membrane, may result from the spread between two closely spaced microdomains; thus, if proximal Ca2+ signals are variable, the distal Ca2+ signals are also variable between ribbons. Alternatively, the variability in distal Ca2+ signals could be due to additional mechanisms, for example, Ca2+ influx from internal stores. Future studies should aim to elucidate the specific mechanisms underlying the observed heterogeneity in distal Ca2+ signals. For instance, exploring differences in Ca2+ release from intracellular stores or variations in Ca2+ sequestration, as reported in goldfish, could reveal key contributing factors 89. Although the potential impact of these intracellular Ca2+ stores on Ca2+ microdomain heterogeneity in hair cells has not been reported,57 previous studies in goldfish bipolar cell terminals have documented spatially restricted Ca2+ oscillations in voltage-clamped retinal bipolar cells, occurring independently of membrane potential 26. To our knowledge, the sources of Ca2+ for these oscillations remain unknown. However, previous studies in goldfish bipolar cells suggest that the endoplasmic reticulum89 and mitochondria24 may act as internal Ca2+ stores. Notably, IP₃ receptors have been localized to retinal bipolar cell terminals in some species.90,91

In the current study, we investigated possible mechanisms for heterogeneity in proximal Ca2+ signals by measuring the ribbon size, particularly the area of the ribbon adjacent to the plasma membrane, where Cav channel clusters are located. Our SBF-SEM shows substantial variability in synaptic ribbon size, shape, and the area of the ribbon facing the plasma membrane where Cav channel clusters are tethered. Since both Ca2+ signals and the area of the synaptic ribbon facing the plasma membrane are heterogeneous, we propose that the larger ribbons may anchor more channels, leading to larger Ca2+ microdomains. Since local Ca2+ signals control kinetically distinct neurotransmitter release components, heterogeneity in local Ca2+ signals may alter the rate of vesicle release and allow them to function independently. Indeed, heterogeneity in neurotransmitter release kinetics has been proposed for the observed diversity in excitatory postsynaptic current amplitudes and kinetics reported from paired recordings of goldfish bipolar cells86 and ON-type mixed RBCs 92. The heterogeneity in the RBC Ca2+ microdomains, synaptic ribbon shape, size, and active zone area reported in this study may contribute to regulating the dynamic range of RBC ribbon synapses28,93,94 – a hypothesis that needs to be tested in future studies.

Methods

Rearing of zebrafish

Male and female zebrafish (Danio rerio; 16 ∼20 months) were raised under a 14 h light/10 h dark cycle and housed according to NIH guidelines and the University of Tennessee Health Science Center (UTHSC) Guidelines for Animals in Research. All procedures were approved by the UTHSC Institutional Animal Care and Use Committee (IACUC; protocol # 23-0459).

Isolation of zebrafish retinal RBCs

Dissociation of RBCs was performed using established procedures 95. Briefly, retinas were dissected from zebrafish eyes and incubated in hyaluronidase (1100 units/ml) for 20 minutes. The tissue was washed with a saline solution containing 120 mM NaCl, 2.5 mM KCl, 0.5 mM CaCl2, 1 mM MgCl2, 10 mM glucose, and 10 mM HEPES, pH = 7.4] before being cut into quadrants. Each quadrant was incubated at room temperature for 25 to 40 min in the same saline solution, to which was added DL-cysteine and papain (20-30 units/ml; Sigma Millipore, St. Louis, MO) and triturated using a fire-polished glass Pasteur pipette. Individual cells were transferred to glass-bottomed dishes, allowed to attach for 30 minutes, and washed with saline solution before being used for experiments.

Ribeye binding peptides

As a means of localizing the ribbons, custom peptides containing the ribbon binding sequence fused to TAMRA (tetramethylrhodamine; TAMRA: GIDEEKPVDLTAGRRAG) dye were synthesized, purified, and purchased from LifeTein (>95% purity, LifeTein LLC, NJ).

Ca2+ indicators

Free Ca2+ indicators

The potassium salts of high and low-affinity Ca2+ indicators Cal-520® (high affinity, KD 320 nM) and Cal-520N™ AM (low affinity, KD 90 µM), referred to as Cal520HA and Cal520LA, respectively were purchased from AAT Bioquest.

Direct conjugation of Ca2+ indicator to cysteine-containing ribbon-binding peptides

To target Ca2+ indicators to the ribbon, custom-made cysteine-containing ribeye binding peptides NH2-CIEDEEKPVDLTAGRRAC-COOH were synthesized and purchased from LifeTein to directly fuse with the fluorogenic 520® maleimide (purchased from AAT Bioquest) for high affinity (HA) and low affinity (LA Ca2+ indicator dyes). Each peptide at one mM concentration was mixed and incubated with two mM Cal-520® maleimide (20 mM stock solution in DMSO purchased from AAT Bioquest) for one hour at room temperature, then overnight at 4°C. Calibration of Ca2+ indicator dyes was performed as described 25 and as detailed below. The conjugated Ca²⁺ indicators were stored at -20°C in smaller aliquots, each sufficient for a single day’s experiments.

Measurement of dissociation constants (KD) for Ca2+ indicator peptides

The effective KD (Keff) was obtained by measuring the fluorescence of Cal520HA-RBP and TAMRA-RBP in buffered Ca2+ solutions, determining the ratio between them 25, and using the Grynkiewicz equation to determine the Ca2+ concentration [Ca2+] from this ratio 59. However, this was not possible for the low-affinity indicator Cal520LA due to the large Ca2+ levels required to calibrate it. Thus, Keff for Cal520LA-RBP could be larger than the K1/2 provided by the manufacturer for Cal520LA (KD, 90 μM), as reported previously with Keff, measurements of OB-5N in inner hair cells 42. Thus, our estimates of local Ca2+ concentrations obtained using Cal520LA represent the lower bounds of the underlying true values. As noted in the Results section, the same may be true for the high-affinity Cal520HA despite its accurate Keff estimate, due to potential dye saturation effects.

RBC voltage clamp recording

Whole-cell patch-clamp recordings were made from isolated RBCs, as described previously 25,33,96. Briefly, a patch pipette containing pipette solution (120 mM Cs-gluconate, 10 mM tetraethyl-ammonium-Cl, 3 mM MgCl2, 0.2 mM N-methyl-d-glucamine-EGTA, 2 mM Na2ATP, 0.5 mM Na2GTP, 20 mM HEPES, pH = 7.4) was placed on the synaptic terminal, as described previously. The patch pipette solution also contained a fluorescently-labeled RBP peptide (TAMRA-RBP) to mark the positions of the ribbons and either 1) free Ca2+ indicators Cal520HA-free (Figs. 1 and 2) and Cal520LA-free (Fig. 2, 3 and 5) to demonstrate our nanophysiological approach or 2) ribeye-bound Ca2+ indicators Cal520HA-RBP (Figs. 3 and Supplementary Figs. 3 and 5) and Cal520LA-RBP (Figs. 3, 4, 5, and 8) to measure local ribbon-associated Ca2+ signals. Current responses from the cell membrane were recorded under a voltage clamp with a holding potential (VH) of -65 mV that was stepped to 0 mV (t0) for 10 milliseconds. These responses were recorded with a patch clamp amplifier running PatchMaster software (version v2×90.4; HEKA Instruments, Inc., Holliston, MA). Membrane capacitance, series conductance, and membrane conductance were measured via the sine DC lock-in extension in PatchMaster and a 1600 Hz sinusoidal stimulus with a peak-to-peak amplitude of 10 mV centered on the holding potential 97.

Acquisition of confocal images

Confocal images were acquired using an Olympus model IX 83 motorized inverted FV3000RS laser-scanning confocal microscopy system (Olympus, Shinjuku, Tokyo, Japan) running FluoView FV31S-SW software (Version 2.3.1.163; Olympus, Center Valley, PA) equipped with a 60 X silicon objective (NA 1.3), all diode laser combiner with five laser lines (405, 488, 515, 561 & 640 nm), a true spectral detection system, a hybrid galvanometer, and a resonant scanning unit. Fluorescently labeled ribeye binding peptide (RBP) 32 and Ca2+ indicator were delivered to RBC via a whole-cell patch pipette placed directly at the cell terminal. We waited for 30 s after break-in to allow Cal520HA to reach equilibrium with the patch-pipette before obtaining the first fluorescence image. Rapid x-t line scans at the ribbon location were performed to localize synaptic ribbons (Fig.1Aii) and to monitor local changes in Ca2+ concentration at a single ribbon, as we demonstrated previously, to estimate the Ca2+ levels at the plasma membrane and to track a single synaptic vesicle at ribbon locations 25,26. The z-projection from a series of confocal optical sections through the synaptic terminal (Fig. 1Ai) illustrates ribbon labeling (magenta spots). RBP fluorescence was used to localize a synaptic ribbon and to define a region for placing a scan line perpendicular to the plasma membrane, extending from the extracellular space to the cytoplasmic region beyond the ribbon to monitor changes in the Ca2+ concentration along the ribbon axis (Fig. 1Aii). The focal plane of the TAMRA-RBP signal was carefully adjusted for sharp focus to avoid potential errors arising from the high curvature near the top of the terminal and the plane of membrane adherence to the glass coverslip at the bottom of the terminal. Sequential dual laser scanning was performed at rates of 1.51 milliseconds per line. Two-color laser scanning methods allowed observation of Ca2+ signals (Fig.1C, cyan) throughout the full extent of the ribbon in voltage-clamped synaptic terminals, while the ribbon and cell border were imaged with a second fluorescent label. The Exchange of TTL (transistor-transistor logic) pulses between the patch-clamp and imaging computers synchronized the acquisition of electrophysiological and imaging data. The precise timing of imaging relative to voltage-clamp stimuli was established using PatchMaster software to digitize horizontal-scan synced pulses from the imaging computer in parallel with the electrophysiological data (Fig.1B). Acquisition parameters, such as pinhole diameter, laser power, PMT gain, scan speed, optical zoom, offset, and step size were kept constant between experiments. Sequential line scans were acquired at 1-2 millisecond/line and 10 μs/pixel with a scan size of 256 × 256 pixels. Bleed-through between the channels was confirmed with both lasers using the imaging parameters we typically use for experiments. To test bleed-through from the RBP channel (TAMRA-RBP) to the Ca2+ indicator (Cal-520HA and Cal-520LA), whole-cell recordings from RBC terminals were performed with patch pipette solution that contained TAMRA-RBP or the aforementioned Ca2+ indicators, and line-scan images were collected and analyzed using the same procedures used for experimental samples.

Point spread function

The lateral and axial point spread function (PSF) was obtained as described previously 25,34,95. Briefly, an XYZ scan was performed through a single 27 nm bead and the maximum projection in the xy-plane was fit to the Gaussian function using Igor Pro software. We obtained the full width at half maximum (FWHM) values for x and y-width of 268 and 273 nm, respectively, in the lateral (x-y plane) and for y-z width of 561 nm in the axial (y-z axis) resolution.

Photobleaching

We minimized photobleaching and phototoxicity during live-cell scanning by using fast scan speed (10 microsec/pixel), low laser intensity (0.01–0.06% of maximum), and low pixel density (frame size, 256 × 256 pixels). We estimated photobleaching using x-t line scans of Cal520HA or Cal520LA and TAMRA-RBP in the absence of stimulation, with the same imaging parameters used for experimental samples.

Data analysis

Quantitative FluoView x-t and x-y scans were analyzed initially with ImageJ software (imagej.nih.gov) and subsequently with Igor Pro software (Wavemetrics, Portland OR) for curve fitting and production of the figures. Data from PatchMaster software were initially exported to Microsoft Excel (Version 16.81) for normalizing and averaging and exported from MS Excel to Igor Pro (Version 9.05) for curve fitting and production of the l figures.

Analysis of x-t scan data

X-axis profile

To determine the Ca2+ signals along the ribbon axis, we spatially averaged the x-axis profile intensity of RBP (i.e., a horizontal row of pixels, see Fig.1C, top) to determine the position of the center of the ribbon and estimate the location of the plasma membrane. The parameter x0 is the peak of the Gaussian fit, giving the x-position of the center of the ribbon. The x-axis profile is also used to obtain the spatial profile of Ca2+ signals before, during, and after with respect to the ribbon profile. To identify the Ca2+ signals specific to the ribbon location, we fit x-axis intensity profiles with the equation f(x) = s(x) + g(x), where s(x) is a sigmoid function that describes the transition from intracellular to extracellular background fluorescence at the edge of the cell, given by s(x) = b − c / (1 − exp((x1/2 − x)/d)), and g(x) is a Gaussian function that represents the fluorescence of RBP, given by g(x) = a exp(−(x − x0)2/w2), as described 33. The parameters x1/2 and x0 were taken as the x-axis positions of the plasma membrane and the fluorescence emitter, respectively. While parameter b is intracellular background fluorescence, c is extracellular background fluorescence, d is the slope of the sigmoid, a is the peak amplitude of emitter fluorescence, and w is √2 times the standard deviation of the Gaussian function, in practice, the latter parameters were highly constrained by the data or by the measured PSF, leaving only x1/2/ and x0 as free parameters in the fitting. Fig.1D, demonstrates that the peak of the Ca2+ signals (x$, cyan; Fig.1D) during the stimuli is proximal to the ribbon center (x0, magenta; Fig.1D) towards the plasma membrane (x!/#, magenta; Fig.1D), as expected for Ca2+ influx originating from Ca2+ channels localized in the plasma membrane 50–52.

T-axis profile

The temporal profiles of the Cal520, Cal520-RBP, and TAMRA-RBP signals were determined by analyzing the time-axis profile of the x-t line scan to obtain the kinetics of the Ca2+ transient with respect to the ribbon. We determined the baseline kinetics by averaging the fluorescence obtained immediately before depolarization. The timing of depolarization and the amplitude of the Ca2+ current was obtained from the PatchMaster software. The rising phase of the Ca2+ transient was fit with the sigmoid function, the peak of which is referred to as the peak amplitude Ca2+ transient.

We used this baseline profile of ribbon-proximal Ca2+ signals to distinguish signals proximal or distal to the ribbon despite the distance between these two signals being within the PSF. For example, the ribbon-proximal signals were obtained by averaging 5 pixels of the temporal profile of the Ca2+ signals (obtained with Cal520 or Cal520-RBP) between the x1/2 and x0 values obtained in the x-axis profile of TAMRA-RBP. The distal profile was obtained similarly but was 5-pixels after x0 towards the cytoplasm.

Quantifying the kinetics of the Ca2+ transients

The decay phase of the fluorescence transients in Figs. 2 and 3 were fit using the standard bi-exponential sum with time constants τfast and τslow: Where Fpeak and tpeak are the peak fluorescence and the time at which this peak has been reached, and afast and aslow are the relative proportions of the two decay components, specified as percentages and satisfying afast + aslow 100%. The decay parameters were found using a global differential evolution optimization algorithm 98, repeating the optimization 400 times starting with different initial parameters values, to ensure true global minimum of the fit error is reached. We note however that bi-exponential data fitting is known to be an ill-conditioned procedure, highly sensitive to noise and other recording characteristics such as the duration of the measurement. Therefore, much higher signal-to-noise ratio would be required for reliable uncertainty quantification of the corresponding kinetic fit parameters, so we only fit the trial-averages data for fluorescence traces in Figs. 2 and 3 as a rough description of temporal decay of the signal; no precise quantitave conclusions should be drawn from the decay time values. Still, time-course fit of the averaged traces may provide a qualitative-level comparison of stimulus-evoked fluorescence transients recorded with different dyes at different locations with respect to the ribbon.

Analysis of x-y scan data

We analyzed the rate of loading the TAMRA-RBP and Ca2+ indicator into the RBC terminal with ImageJ software by placing a square region of interest (ROI; 5 x 5 μm) and using a Plot z-axis profile function to obtain spatially averaged TAMRA-RBP and Ca2+ indicator fluorescence as a function of time. The rising phase of TAMRA-RBP and Ca2+ indicator fluorescence was obtained by fitting the rising to the peak to the single exponential function using the curve fitting function in Igor Pro 9 software. The rate of the exponential function is defined as the rate of fluorescence loading to the terminal.

For analysis comparing the fluorescence intensity of ribbons and Ca2+ influx elicited by 100-ms stimuli (Fig. 11A & B), 10 images of both TAMRA-RBP and Cal520LA-RBP were collected in sequential, followed by imaged Cal520LA-RBP only during 200-ms depolarization, and the sequence ended with 10 images of both TAMRA-RBP and Cal520LA-RBP in sequential to confirm the ribbon locations. Two color imaging were obtained sequentially with a frame interval of 407 ms, and Cal520LA-RBP only during 100ms-stimuli was 205 ms. Images before and after were then averaged to obtain the ribbon location. Synaptic ribbon fluorescence was quantified as the ratio of TAMRA fluorescence to the fluorescence of the nearby RBC cytoplasm, measured approximately eight to nine pixels away (Fribbon/Fnear by), measuring the pixel with the highest intensity. The change in Cal520LA-RBP fluorescence, used as a proxy for active zone Ca2+ influx, was estimated as ΔF/Frest, where Frest is the fluorescence intensity before depolarization at −65 mV, and ΔF is the difference when depolarized to 0 mV. Ca2+ indicator intensity was calculated as the average fluorescence of nine pixels centered on the pixel with the greatest fluorescence increase.

Experimental design and statistical methods

We did not use any statistical methods to determine the sample size prior to experiments. Mean value variances were reported as ± the standard error of the mean (SEM). The statistical significance of differences in average amplitudes of Ca2+ current, capacitance, synaptic ribbon size and number, and Ca2+ transients were assessed using unpaired, two-tailed t-tests with unequal variance using R Studio (Version 2023.09.0+463) and Igor Pro software.

Amplitude analysis for Figure 8

Amplitude data were processed and analyzed using R Studio software (Version 2023.09.0+463). The data were divided into two groups, that of individual cells and of multiple cells, and analyzed as follows. Multiple individual cells from each condition (proximal 10 mM EGTA Cal520LA-RBP and distal 10 mM EGTA Cal520LA-RBP) were analyzed to compare the variability between the Ca2+ amplitude measurements of the ribbons from each cell. Having verified the normality of the data to ensure the validity of the statistical test, we used one-way ANOVA and Tukey’s honest significant difference test to assess the specific statistical differences between ribbons. To compare the variability between the amplitude measurements of different cells, we compared readings for various cells. If there were multiple measurements for a single ribbon in a specific cell, those readings were averaged. In this case, the normality of the data was again verified and Welch’s ANOVA was employed to assess the between-group differences as it does not assume homogeneity of variance, with the Games-Howell post-hoc test being used to understand the specific differences between cells.

Computational modeling of Ca2+ dynamics

Spatio-temporal Ca2+ dynamics is simulated in a volume shown in Fig. 9, which is a 3D box of dimensions (1.28 x 1.28 x 1.1) µm3. Assuming an approximately spherical bouton of diameter 5µm and an average of 36 ribbons per terminal, one obtains synaptic volume per ribbon of 1.81 µm3, which matches the volume of our box domain. Ribbon serves as an obstacle to Ca2+ and buffer diffusion, and has an ellipsoidal shape with semi-axis of 190nm along the Y- and Z-axes (for a total height of 380nm), and a semi-axis of 70 nm along the X-axis. It is attached to the plasma membrane by a thin ridge (arciform density) of dimensions 60×30×30 nm. Ca2+ ions enter through 4 channels (or clusters of channels) at the base of the ribbon indicated by black disks in Fig. 6, forming a square with a side length of 80 nm; this highly simplified channel arrangement is sufficient to capture the level of detail in spatial [Ca2+] distribution that we seek to resolve. Each of the four clusters admits a quarter of the total Ca2+ current. Ca2+ current values and current pulse durations are listed in Fig. 7. Although the actual distribution of Cav channels is more heterogeneous, channels close to the ribbon play a greater role in neurotransmitter release 2.

The simulation includes a single dominant Ca2+ buffer with molecules possessing a single Ca2+ ion binding site. We assume that only immobile buffer is present in a patch-clamped cell, since any mobile buffer would diffuse into recording pipette. Simulations with mobile buffer in an un-patched cell are also provided (Supplementary Fig. 4B&C). Buffer parameter values were set to values reported by 7, namely dissociation constant of 2 µM, and total concentration of 1.44 mM (total buffering capacity at rest of 720). We also examine the impact of the assumption on buffer concentration by repeating the simulation with a lower buffering capacity of 100 (total buffer concentration of 200 µM). The Ca2+ ion diffusivity is set to 0.22 µm2/ms 99.

To minimize the number of undetermined parameters, we assumed the same parameters for the Ca2+ clearance on all surfaces, with one high-capacity but lower affinity process simulating Na+/Ca2+ exchangers, and one lower capacity but higher affinity process simulating Ca2+ extrusion by ATPase PMCA and SERCA pumps. We used the Hill coefficient of m=2 for the Ca2+ sensitivity of the SERCA pumps on all surfaces except the bottom surface 100, whereas Hill coefficient of m=1 corresponds to PMCA pumps on the bottom surface 101. Thus, the flux of [Ca2+] across the surface (ions extruded per unit area per unit time) is described by where C is the Ca2+ concentration, DC is the Ca2+ diffusivity, and n ⋅ ∇C is the gradient of [Ca2+] normal to the boundary. The constant leak term ΦLeak ensures that the flux is zero at the resting value [Ca2+]rest = C0 = 0.1 µM. The Ca2+ clearance rate and affinity parameters for the pumps and exchangers are listed in Supplementary Table 3. We note however that the density of Na+-Ca2+ exchangers and PMCA/SERCA pumps are rough estimates and have a significant effect on results only for longer stimulation durations of over 100-200 ms. Reaction-diffusion equations for [Ca2+] and buffer are solved numerically using Calcium Calculator software version 6.10.7 102,103.

Serial block-face scanning electron microscopy (SBF-SEM)

The eyecups of adult (16 ∼20-month-old) zebrafish were dissected in bicarbonate-buffered Ames’ medium and, after removing the lens, the eyecups were fixed in 4% glutaraldehyde in 0.1 M sodium-cacodylate buffer at RT for one hour, followed by overnight fixation at 4°C. Thereafter, samples were treated to generate blocks for serial block face scanning electron microscopy (SBF-SEM) according to the published protocol 104, in this case using Zeiss 3-View SBF-SEM. Multiple 35 µm montages were acquired at a 5 nm X-Y resolution and a 50nm Z resolution across the entire retina from the outer (photoreceptor) layer to the inner (ganglion) cell layer. Images were inspected visually and structures of interest were traced using the TrakEM plug-in in ImageJ software (NIH/Fiji), as described below.

Analysis of serial block-face scanning electron microscopy images

SBF-SEM image stacks of the zebrafish retina were imported, aligned, visualized, and analyzed using the TrakEM plugin (version 1.5h) in ImageJ software (NIH). RBC1 were identified based on their characteristic morphology and 4-7 μm terminal size, as previously described 80. RBCs, synaptic ribbons, pre-synaptic membranes, and the area of the ribbon touching the plasma membrane were traced, painted using the TrakEM brush features, and rendered in 3D for structural visualization. The area of the synaptic ribbon touching the RBC membrane was quantified using the measuring tools in ImageJ and images were prepared using Adobe Photoshop software (version 25.9). The distance between each ribbon from three different RBCs and its nearest five ribbons (Fig. 9B and Video 1) was measured as follows: Ribbons on the same plane had their linear distance measured, whereas the distance between ribbons that were close to each other but in different layers was calculated using the Pythagorean Theorem.